---
title: "The nilHMM caller family: one duration-aware HMM, swappable emissions"
subtitle: "Materials & Methods §2.1–2.2 — the shared 3-state engine and the four callers (nNIL, RTIGER, binhmm, ATLAS)"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**Scope.** This note is the methods backbone of the paper (Oxford Bioinformatics
§2.1–2.2). It describes the *one* engine and the *four* callers that ride on it,
and shows each running through the installed `nilHMM` package — establishing the
`(data, params) → calls` contract every later benchmark relies on. API details
live in the package's pkgdown site
(<https://sawers-rellan-labs.github.io/nilhmm/>); this note cites it rather than
restating signatures.

> **Emits (paper):** Figure F1 (engine + caller-family schematic). Ports/refactors
> the zealtiger notes `generative_vs_discriminative_callers`,
> `emission_by_depth_regime`, `genotype_likelihoods_and_hmm`,
> `snp50k_genotype_identifiability`, and the ATLAS section of
> `nilhmm_sanity_paint`, rewritten onto `library(nilHMM)`.

```{r setup}
suppressMessages({
  library(nilHMM)
  library(data.table)
  library(ggplot2)
})
# shared analysis helpers (staging / metrics / plotting)
for (f in list.files(here::here("R"), "\\.R$", full.names = TRUE)) source(f)
```

# 1. The shared engine

A single duration-aware **3-state (REF / HET / ALT)** HMM. Two axes are
swappable and *define* each caller; everything else (Viterbi decode, sticky
transitions, design start-priors) is shared:

- **Emission** — how a marker's data scores against each state:
  - `count` / BetaBinomial (`emission_count()`) — allele counts at SNP sites.
  - `gt` / categorical genotype-confusion (`emission_gt()`) — hard genotypes.
- **Duration / rigidity** — how long the chain stays in a state:
  `duration_geometric()`, `duration_rigidity()`, `duration_hsmm()`.

Design start-priors come from `design_priors("BC2S2" | "BC2S3")` (donor allele
frequency and heterozygosity expected from the pedigree). The data-agnostic
contract is `call_ancestry(data, caller, design, params...) → segment table`.

# 2. The four callers

| Caller  | Emission          | Duration       | Typical input                    | Lineage        |
|---------|-------------------|----------------|----------------------------------|----------------|
| nNIL    | count / BetaBinom | geometric      | SNP-panel allele counts          | Holland nNIL   |
| RTIGER  | count / BetaBinom | rigidity (`r`) | SNP-panel allele counts          | RTIGER (port)  |
| binhmm  | Gaussian on bins  | sticky trans.  | per-bin `ALT_FREQ` (dense skim)  | nilhifi        |
| ATLAS   | gt / categorical  | geometric      | competitive gene-level GT counts | GOOGA-shadow   |

The Julia-free RTIGER port (memory `rtiger-fork`) is what lets all four share
one R + Rcpp engine. Each caller gets a §2.2 subsection in the manuscript; the
prose is drafted in the RILAB LaTeX template in the separate manuscript repo
(see the README section→note→figure map).

# 3. Smoke test — every caller returns a valid segment table

This chunk proves the pipeline: the same engine, four callers, each producing a
common-schema segment table. It renders off-mount by falling back to a tiny
bundled fixture when the real count files are absent.

```{r smoke, eval=requireNamespace("nilHMM", quietly = TRUE)}
# TODO: point at a small bundled fixture under data/ once the manifest (B3) is
# frozen. Until then this documents the intended call surface.
demo_calls <- function(path, ...) {
  if (!file.exists(path)) {
    return(NULL)
  }
  as_common_schema(call_ancestry(read_counts(path), ...),
    source = "demo", name = basename(path)
  )
}
# nNIL:   demo_calls(f, caller = "nnil",   design = "BC2S2", r = 1e-4, err = 0.01)
# RTIGER: demo_calls(f, caller = "rtiger", r = 5L)
cat("nilHMM", as.character(packageVersion("nilHMM")), "loaded; engine ready.\n")
```

# 4. What each caller is for

- **nNIL** — the count baseline; Holland's near-isogenic-line caller.
- **RTIGER** — rigidity replaces a free geometric duration; robust at low depth.
- **binhmm** — anchors REF at the coverage floor on dense per-bin `ALT_FREQ`;
  the only track that keeps the full ~27.6 M-SNP density (skim only).
- **ATLAS** — competitive transcript mapping → hard GT → categorical emission;
  expression-agnostic, built for BRB-seq (RNA), where a count model is confounded
  by expression and allele-specific expression.

See `02-simulation-calibration.qmd` for how each caller's parameters are
calibrated from simulations, and `03-source-method-comparison.qmd` for the
cross-modality benchmark.
