---
title: "Simulation-calibrated ancestry calling: fit on simcross, validate on held-out sims"
subtitle: "Materials & Methods §2.4–2.5 + Results §3.1–3.2 — the paper's core contribution"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**Thesis (the paper's novelty).** Ancestry-caller parameters — rigidity `r`,
error `err`, emission means — can be **calibrated from cheap Broman `simcross`
simulations** rather than from scarce high-coverage truth samples. Holland
calibrated on high-coverage genotyping (our analogue = MolBreeding target-seq),
but those samples are too few to calibrate from; we show **simple simulations
suffice** and that the calibrated parameters transfer to real data (B2.3).

> **Emits (paper):** Figure F2 (simcross calibration), Figure F3 (simulation
> benchmark / caller ranking). Core Results §3.1–3.2.

::: {.callout-important title="STATUS: partially blocked (B5 open input)"}
The generator scaffold lives in `R/simulate.R` (simulations are written to the
gitignored `sim/` dir) and uses `nilHMM::load_map()` + `kbroman/simcross`. The
concrete **simcross design to match BZea** — generations (BC2S2 bulked skim vs
BC2S3), n per condition, cM map + interference — is still open (plan B5) and
must be fixed before the calibration numbers are final. Seeds are pinned in
`R/simulate.R` for reproducibility.
:::

```{r setup}
suppressMessages({
  library(nilHMM)
  library(data.table)
  library(ggplot2)
})
for (f in list.files(here::here("R"), "\\.R$", full.names = TRUE)) source(f)
```

# 1. The simulation model (§2.4)

Forward-simulate NILs with `simcross` on the consensus genetic map (cM +
interference), then **degrade to each source's depth/error regime** so a
simulated NIL looks like a Skim or BRB sample:

- BC2S3 (and BC2S2 bulked skim, memory `bzea-skim-bulked-bc2s2`) pedigrees.
- Per-locus genotype expectations at the panel marker set.
- Coverage-driven missingness `missing(λ) = π + (1 − π)e^{−kλ}` and per-read error.

Generator: `R/simulate.R::simulate_source()`. Truth is the simulated
mosaic; the degraded counts are the caller input.

# 2. Calibration protocol (§2.5a — the contribution)

Fit each caller's parameters **on sim truth**, per data source × caller:

- **Rigidity `r`** by KS-vs-sim minimum-distance, *not* MLE (memory
  `rigidity-not-mle`): choose `r` so the simulated fragment-size distribution
  matches the caller's — `nilHMM::calibrate_r()` / `expected_fragment_dist()`.
- **`err` and emission means** on degraded-sim truth (memory
  `skim-count-caller-calibration`).

Then **transfer the calibrated params to the real 400-sample cohort** (B2.3),
explicitly contrasting with Holland's high-coverage-truth calibration
(target-seq analogue, too few samples). The claim: *sims replace truth samples
for calibration.*

```{r calibrate, eval=FALSE}
# Skeleton — fill once simulate_source() emits (truth, degraded counts) per source.
# map  <- nilHMM::load_map()
# sims <- simulate_source(map, design = "BC2S3", source = "skim", n = 200, seed = 1L)
# r_hat <- nilHMM::calibrate_r(sims$counts, sims$truth, design = "BC2S3")
```

# 3. Validation on held-out sims (§2.5b, Results §3.2)

On sims **not** used for calibration, report per caller × source:

- per-bin state accuracy + REF/HET/ALT confusion (`state_accuracy()`),
- segment-boundary error and breakpoint-count vs truth (`breakpoint_count()`),
- donor-footprint ROC (`donor_footprint()` swept over thresholds),
- HET-vs-ALT calibration.

Reuse the n = 1500 expectation and fragment-size null (memory
`bc2s3-fragment-size-expectation`) and the RTIGER benchmark set (zealtiger
`docs/06`, `07`, `09-rigidity-…`). Deliver the caller ranking on truth here.

```{r validate, eval=FALSE}
# grid <- sims$grid            # shared marker/bin midpoints
# acc  <- state_accuracy(called, sims$truth, grid)
# acc$accuracy; acc$confusion
```

# 4. Next steps

1. Fix the simcross design with the user (B5) → finalize `R/simulate.R`.
2. Run calibration across {Skim, BRB} × {nNIL, RTIGER, binhmm, ATLAS}.
3. Produce F2 (calibration curves) and F3 (benchmark ranking).
