---
title: "Simulation-calibrated ancestry calling: fit on simcross, validate on truth"
subtitle: "Materials & Methods §2.4–2.5 + Results §3.1–3.2 — the paper's core contribution"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**Thesis (the paper's novelty).** Ancestry-caller parameters (`rrate`, `rigidity`,
`err`, emission means) can be **calibrated from cheap Broman `simcross`
simulations** rather than from scarce high-coverage truth samples, and the
calibrated parameters transfer across Skim and BrB modalities.

**Methodological point (this section's argument).** Holland's nNIL grid search
maximized **marker true-positives** (donor-marker recall). That objective rewards
calling *more* donor, which **over-fragments** — it opens spurious short donor
blocks to catch every donor marker. We instead **calibrate on F1** (marker and
donor-fragment) and show the true-positive optimum inflates the breakpoint/
crossover count at little F1 gain. All metrics live in `R/metrics.R`.

::: {.callout-important title="STATUS: metrics ready; simcross truth generator pending"}
The scoring harness (`R/metrics.R`: `marker_f1`, `donor_fragment_f1`,
`fragment_size_ks`, `genotype_fractions`, `single_locus_expectation`,
`calibrate_sweep`) is implemented and tested. The **BC2S2, n = 1500 simcross
truth** is generated by `R/simulate.R` (still a scaffold) → the sim-dependent
chunks below are gated on the truth files existing and are the intended shape.
:::

```{r setup}
suppressMessages({
  library(nilHMM)
  library(data.table)
  library(ggplot2)
})
for (f in list.files(here::here("R"), "\\.R$", full.names = TRUE)) source(f)

# simcross truth + degraded counts land here (gitignored); R/simulate.R writes them.
SIM <- here::here("results/sim")
truth_file <- file.path(SIM, "bc2s2_truth_segments.csv") # common schema
have_sim <- file.exists(truth_file)
```

# 1. The simulation model — BC2S2, n = 1500 (§2.4)

One truth, for simplicity: **BC2S2** (2 backcrosses to B73, 2 selfs), the bulked-
skim generation (memory `bzea-skim-bulked-bc2s2`; `bc2s2_vs_bc2s3` argues bulked
skim references BC2S2, one meiosis short of BC2S3). Forward-simulate with
`simcross` on the maize consensus cM map (~1540 cM, per TeoNAM / Chen 2019) with
**Stahl interference m = 10, p = 0**, **n = 1500** NILs, fixed seed. Truth = the
donor mosaic (common-schema segments). Then **degrade** to each source's regime
(coverage-driven missingness `missing(λ) = π + (1 − π)e^{−kλ}`, per-read error)
so a simulated NIL looks like a Skim / BrB sample.

Generator: `R/simulate.R::simulate_source(design = "BC2S2", ...)`. Expected per-
locus genotype target (the calibration/validation anchor):

```{r expectation}
p0 <- single_locus_expectation(n_bc = 2L, n_self = 2L)
round(p0, 4) # REF 0.844 / HET 0.0625 / ALT 0.0938 ; donor dosage 0.125
```

# 2. Calibration — fit the duration knob on truth (§2.5a, Results §3.1) — Fig F2

Sweep the caller's duration knob (`rrate` for nNIL, `rigidity` for RTIGER) and
score each value against the BC2S2 truth with `calibrate_sweep()`. The curve
gives three coincident optima (F1-max, KS-min, crossover-match) — and shows the
true-positive (recall) optimum sits elsewhere, over-fragmenting.

```{r calibrate, eval=have_sim}
sim <- fread(truth_file) # name, chr, start_bp, end_bp, state (BC2S2 truth)
grid <- fread(file.path(SIM, "grid.csv")) # chr, pos (marker grid)
skim <- read_counts_dir(file.path(SIM, "skim_counts")) # degraded-sim Skim counts (helper TODO)

nnil_sweep <- calibrate_sweep(skim, sim, grid,
  caller = "nnil", param = "rrate",
  values = 10^seq(-6, -2, by = 0.25), design = "BC2S2", err = 0.01
)
rtiger_sweep <- calibrate_sweep(skim, sim, grid,
  caller = "rtiger", param = "rigidity", values = 1:12, design = "BC2S2"
)

# F1 optimum vs the Holland true-positive (marker donor-recall) optimum
f1_opt <- nnil_sweep[which.max(donor_frag_F1)]
tp_opt <- nnil_sweep[which.max(donor_marker_recall)]
rbind(cbind(objective = "F1-optimal", f1_opt), cbind(objective = "TP-optimal", tp_opt))[
  , .(objective, value, donor_frag_F1, donor_marker_recall, donor_frag_FDR, n_breakpoints, ks_fragsize)
]
```

```{r fig-F2, eval=have_sim, fig.width=7, fig.height=4}
# F2a: donor-fragment F1 vs marker donor-recall (Holland's TP objective) across
# the swept parameter; the F1 optimum (dashed) and the TP optimum differ.
# F2b/F2c: n_breakpoints and ks_fragsize vs `value` are the over-fragmentation
# and fragment-size-fit companion panels (same nnil_sweep columns).
ggplot(nnil_sweep, aes(value)) +
  geom_line(aes(y = donor_frag_F1, colour = "donor-fragment F1")) +
  geom_line(aes(y = donor_marker_recall, colour = "marker donor recall (TP)")) +
  geom_vline(xintercept = f1_opt$value, linetype = "dashed") +
  scale_x_log10() +
  labs(
    x = "rrate", y = "score", colour = NULL,
    title = "F2a — F1 vs true-positive objective (nNIL); dashed = F1 optimum"
  )
```

# 3. Validation on held-out sims — caller ranking (§2.5b, Results §3.2) — Fig F3

Score each caller × source at its calibrated parameter against held-out BC2S2
truth: per-state marker F1, donor-fragment F1/FDR/boundary, and the crossover /
breakpoint count vs truth (the over-fragmentation number).

```{r benchmark, eval=have_sim}
score_caller <- function(called, tag) {
  mf <- marker_f1(called, sim, grid)
  ff <- donor_fragment_f1(called, sim)
  data.table(
    caller = tag,
    marker_macroF1 = mf$macro_f1,
    donor_marker_F1 = mf$per_class[class == "donor(>0)", f1],
    donor_frag_F1 = ff$f1, donor_frag_FDR = ff$fdr,
    n_breakpoints = breakpoint_count(called),
    n_breakpoints_truth = breakpoint_count(sim)
  )
}
# rbindlist(list(score_caller(skim_nnil, "Skim-nNIL"), score_caller(skim_rtiger, "Skim-RTIGER"), ...))
```

# 4. Fragment-size fit (supplementary) — the `bc2s2_vs_bc2s3` view

Called donor-block sizes vs the simulated-truth distribution (ECDF + KS). Over-
fragmentation shows as mass shifted to short blocks; a good fit tracks the truth.

```{r fragsize, eval=have_sim, fig.width=7, fig.height=4}
sizes <- rbind(
  data.table(src = "truth", mb = donor_block_sizes(sim))
  # data.table(src = "Skim-nNIL", mb = donor_block_sizes(skim_nnil)), ...
)
ggplot(sizes, aes(mb, colour = src)) +
  stat_ecdf() +
  scale_x_log10() +
  labs(x = "donor block size (Mb)", y = "ECDF", title = "Fragment-size fit vs BC2S2 truth")
```

# 5. Genotype-frequency vs the one-locus expectation (supplementary)

The `skim_brbseq_skimbin` view: per-NIL genome fractions vs the Mendelian
single-locus expectation `p0` and vs the simcross distribution. **Reference
(Chen 2019 / TeoNAM, BC1S4):** ~15% homozygous teosinte, ~8% heterozygous —
donor dosage runs *below* neutral (selection), and a *modest* real het excess
exists; the RNA failure mode (BrB over-calling HET) is a far grosser inflation.

```{r genofreq, eval=have_sim, fig.width=7, fig.height=4}
gf <- genotype_fractions(sim) # + rbind each caller's genotype_fractions, tagged
ggplot(melt(gf[, .(name, HET, ALT)], id.vars = "name"), aes(variable, value)) +
  geom_violin() +
  geom_hline(
    data = data.table(variable = c("HET", "ALT"), y = c(p0["HET"], p0["ALT"])),
    aes(yintercept = y), linetype = "dashed"
  ) +
  labs(x = NULL, y = "genome fraction", title = "Genotype freq vs one-locus expectation (dashed)")
```

# Next steps

1. Wire `R/simulate.R::simulate_source("BC2S2", n = 1500, m = 10)` → write
   `results/sim/{bc2s2_truth_segments.csv, grid.csv, <source>_counts/}`.
2. Run the F2 sweep for nNIL (`rrate`) and RTIGER (`rigidity`) on the degraded
   Skim and BrB sims; confirm F1-opt ≠ TP-opt and the breakpoint blow-up.
3. Produce F3 (caller ranking) + the supplementary fragment-size and
   genotype-frequency panels; carry the calibrated parameters to
   `03-source-method-comparison.qmd`.
```
