---
title: "Simulation-calibrated ancestry calling: fit on simcross, validate on truth"
subtitle: "Materials & Methods §2.4–2.5 + Results §3.1–3.2 — the paper's core contribution"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**Thesis (the paper's novelty).** Ancestry-caller parameters (`rrate`, `rigidity`,
`err`, emission means) can be **calibrated from cheap Broman `simcross`
simulations** rather than from scarce high-coverage truth samples, and transfer
across Skim and BrB modalities.

**Methodological point (this section's argument).** Holland's nNIL grid search
maximized **marker true-positives** (donor-marker recall). On well-covered truth
that objective **saturates** — donor recall is ~0.99 across four orders of
magnitude of `rrate` — so it cannot discriminate a good parameter from a bad one,
and any value inside the plateau looks "optimal." The discriminating axis is on
the *precision* side: raising `rrate` licenses ever-shorter donor blocks, so the
false-discovery rate and breakpoint count climb while recall does not move. We
therefore calibrate on **F1** (marker and donor-fragment), which has a
well-defined optimum whose breakpoint count matches truth. Metrics: `R/metrics.R`;
truth: `R/simulate.R`.

::: {.callout-note title="Compute is separated from this note"}
The sweeps, RTIGER EM, and benchmark run in **`scripts/02_calibrate.R`**, which
writes small summary CSVs to `results/sim/` (gitignored). This note only loads and
plots them, so it renders in seconds. Regenerate everything with
`Rscript scripts/02_calibrate.R --generate`. Truth: BC2S2, n = 1500, **50 000
markers** — the caller runs directly on per-marker skim counts (no binning), at a
density where a donor block spans enough markers for the duration prior to
integrate over.
:::

```{r setup}
suppressMessages({
  library(data.table)
  library(ggplot2)
})
SIM <- here::here("results/sim")
f <- function(x) file.path(SIM, x)
have_calib <- file.exists(f("nnil_rrate_sweep.csv"))
if (have_calib) {
  params <- setNames(fread(f("calib_params.csv"))$value, fread(f("calib_params.csv"))$key)
  nnil_sweep <- fread(f("nnil_rrate_sweep.csv"))
  rtiger_sweep <- fread(f("rtiger_rigidity_sweep.csv"))
  benchmark <- fread(f("benchmark.csv"))
  frag <- fread(f("frag_sizes.csv"))
  geno <- fread(f("geno_fractions.csv"))
}
```

# 1. The simulation model — BC2S2, n = 1500 (§2.4)

**BC2S2** (2 backcrosses to B73, 2 selfs — the bulked-skim generation, memory
`bzea-skim-bulked-bc2s2`). `simcross` on the maize consensus cM map
(`maize_map_v5_clean.rds`, ~1783 cM) with **Stahl interference m = 10, p = 0**,
**50 000 markers**, fixed seed. Truth = the donor mosaic (dosage 0/1/2 at each
marker → common-schema segments). Degrade to the skim regime (coverage-driven
missingness `π + (1−π)e^{−kλ}`, per-read error): skim ≈ 0.43× / ~72% missing.

```{r expectation, eval=have_calib}
# Truth genome composition vs the single-locus expectation. (TeoNAM / Chen 2019 is
# a *different* design — BC1S4, ~15% teosinte-hom / ~8% het — cited elsewhere for
# orientation only, not a target here.)
data.table(
  metric = c("REF", "HET", "ALT", "dosage"),
  truth_mean = round(params[c("REF_truth", "HET_truth", "ALT_truth", "dosage_truth")], 4),
  bc2s2_expectation = round(params[c("REF_expect", "HET_expect", "ALT_expect", "dosage_expect")], 4)
)
```

# 2. Calibration — fit the duration knob on truth (§2.5a, Results §3.1) — Fig F2

`scripts/02_calibrate.R` sweeps `rrate` (nNIL) and `rigidity` (RTIGER) with
`calibrate_sweep()` and scores each against truth. The F1 optimum, the KS-min, and
the truth-matching breakpoint count roughly coincide; the marker true-positive
(recall) objective is flat across the whole sweep and cannot pick a winner.

```{r nnil-sweep, eval=have_calib}
nnil_sweep[, .(
  rrate = signif(value, 2), donor_marker_recall = round(donor_marker_recall, 3),
  donor_frag_F1 = round(donor_frag_F1, 3), donor_frag_FDR = round(donor_frag_FDR, 3),
  n_breakpoints, truth_bp, ks_fragsize = round(ks_fragsize, 3)
)]
```

```{r F1-vs-TP, eval=have_calib}
# Marker donor-recall (Holland's TP objective) is saturated, so a recall-only search
# is *indifferent* between the F1 optimum and the most permissive rrate. Donor-fragment
# F1, FDR, and breakpoint count are not — they pick the parameter that matches truth.
rbind(
  cbind(objective = "F1-optimal", nnil_sweep[which.max(donor_frag_F1)]),
  cbind(objective = "most permissive (recall ties)", nnil_sweep[which.max(value)])
)[, .(objective,
  rrate = signif(value, 2), donor_marker_recall = round(donor_marker_recall, 3),
  donor_frag_F1 = round(donor_frag_F1, 3), donor_frag_FDR = round(donor_frag_FDR, 3),
  n_breakpoints, truth_bp
)]
```

```{r fig-F2, eval=have_calib, fig.width=7, fig.height=4.2}
d <- nnil_sweep[, .(
  rrate = value, `donor-fragment F1` = donor_frag_F1,
  `marker donor recall (TP)` = donor_marker_recall, `donor-fragment FDR` = donor_frag_FDR
)]
dm <- melt(d, id.vars = "rrate", variable.name = "metric", value.name = "score")
ggplot(dm, aes(rrate, score, colour = metric)) +
  geom_line() +
  geom_vline(xintercept = params["rrate_star"], linetype = "dashed") +
  scale_x_log10() +
  labs(
    x = "rrate", y = "score", colour = NULL,
    title = "F2 — F1 has a peak; the TP (recall) objective is flat (dashed = F1 optimum)"
  )
```

# 3. Benchmark at the calibrated parameter (§2.5b, Results §3.2) — Fig F3

Each caller at its F1-optimal knob (`rrate = ` `r if (have_calib) signif(params["rrate_star"],2)`,
`rigidity = ` `r if (have_calib) params["rigidity_star"]`), scored on a held-out
set vs truth: per-state marker F1, donor-fragment F1/FDR, fragment-size KS, and
breakpoint count vs truth.

```{r benchmark, eval=have_calib}
benchmark[]
```

# 4. Fragment-size fit (supplementary) — the `bc2s2_vs_bc2s3` view

```{r fragsize, eval=have_calib, fig.width=7, fig.height=4}
ggplot(frag, aes(mb, colour = src)) +
  stat_ecdf() +
  scale_x_log10() +
  labs(
    x = "donor block size (Mb)", y = "ECDF",
    title = "Fragment-size fit vs BC2S2 truth (KS in the benchmark table)"
  )
```

# 5. Genotype-frequency vs the one-locus expectation (supplementary)

The `skim_brbseq_skimbin` view. Reference (Chen 2019 / TeoNAM, BC1S4): ~15%
teosinte-hom, ~8% het — donor dosage runs *below* neutral (selection) and a
*modest* real het excess exists; the RNA failure mode is a far grosser inflation.

```{r genofreq, eval=have_calib, fig.width=7, fig.height=4}
ggplot(melt(geno[, .(src, HET, ALT)], id.vars = "src"), aes(src, value)) +
  geom_violin(scale = "width") +
  facet_wrap(~variable, scales = "free_y") +
  geom_hline(
    data = data.table(
      variable = c("HET", "ALT"),
      y = c(params["HET_expect"], params["ALT_expect"])
    ),
    aes(yintercept = y), linetype = "dashed"
  ) +
  labs(x = NULL, y = "genome fraction", title = "Genotype freq vs one-locus expectation (dashed)") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

# Next steps

- Add the BrB regime and the per-source calibrated parameters; carry them to
  `03-source-method-comparison.qmd`.
- Add the donor-footprint ROC and HET-vs-ALT calibration panels.
```
