---
title: "Cross-modality concordance on the paired BrB + skim cohort"
subtitle: "Results §3.3 — transfer sim-calibrated params to ~400 NILs sequenced two ways"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**The benchmark cohort is the ~400 NILs that have BOTH BRB-seq and skim data.**
The paired design is the point: the *same* NIL genotype seen through two
modalities, so cross-source disagreement is **method, not sample**. We apply the
sim-calibrated parameters from `02-simulation-calibration.qmd` and quantify
concordance per sample.

> **Emits (paper):** Figure F4 (400-cohort source × method concordance),
> Results §3.3. Chromosome painting appears here only as supplementary QC (see
> `nilhmm_sanity_check_paint.qmd`), never as a main figure.

::: {.callout-important title="STATUS: blocked (B5 open input)"}
Needs the **~400 paired-cohort sample manifest** (IDs, donor species, per-source
depth). Trace from `sample_metatada.csv` / memory `bzea-sample-donor-metadata`
plus the BRB and skim sample maps; the vary-skim sweep members are a subset.
Record it in `DATA.md` before running. This note is the analysis
harness that consumes that manifest.
:::

```{r setup}
suppressMessages({
  library(nilHMM)
  library(data.table)
  library(ggplot2)
})
for (f in list.files(here::here("R"), "\\.R$", full.names = TRUE)) source(f)
mnt <- bzea_mounts()
```

# 1. The three source paths (calibrated params from B2.2)

- **Skim** (low-cov WGS + TP teosinte–B73 wideseq filtering): nNIL, RTIGER, binhmm.
- **BrB** (3′ RNA-seq + TP filtering): nNIL, RTIGER, ATLAS.
- **Target-seq** (MolBreeding 45K, high coverage): `gt` path — *completeness
  only*, and doubles as the optional Holland-style calibration cross-check; not
  headline.

# 2. Concordance metrics (per sample)

Computed on a shared marker/bin grid so tracks with different breakpoints are
comparable (helpers in `R/metrics.R`):

- **pairwise segment agreement** — `pairwise_agreement()`
- **donor-footprint Jaccard** — `donor_jaccard()`
- **breakpoint concordance** — compare `breakpoint_count()` across tracks
- **control cleanliness on B73** — `control_donor_rate()` (should be ~0)

```{r comparison, eval=FALSE}
# manifest <- fread(here::here("data/paired_cohort_manifest.csv"))  # from B5
# for each sample: stage skim + brb calls -> common schema -> metrics on grid.
# jac <- donor_jaccard(skim_calls, brb_calls, grid)
# agr <- pairwise_agreement(skim_calls, brb_calls, grid)
```

# 3. Read (to write once data is in)

Cross-source / cross-caller concordance distributions, and where the modalities
systematically disagree (expected: BRB-RTIGER RNA false-donor floor; ATLAS most
conservative). This is F4.
