---
title: "Supplementary QC: nilHMM sanity paint (source × caller tracks)"
subtitle: "Supplementary figure — visual check that independent callers agree and the B73 control paints clean"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**Role in the paper: supplementary QC only.** Chromosome painting is a *visual*
coherence check — do the independent tracks agree on where the donor blocks are,
and does the B73 control stay REF? — **not** a main figure (plan B2.3). The
quantitative concordance is in `03-source-method-comparison.qmd`.

This is the zealtiger `nilhmm_sanity_paint` note ported onto the installed
package: `devtools::load_all("agent/nilhmm")` → `library(nilHMM)`, and the paint
factored into `R/plotting.R`. Six tracks = 2 data sources (Skim, BrB) × nilHMM
callers, naming `<Source>-<Caller>` (memory `method-naming-source-caller`).

::: {.callout-note title="STATUS: data migration pending (B3)"}
The count/bin inputs still live in the zealtiger working repo. `ZT` below
resolves to that repo as a stand-in data source; once `DATA.md` freezes
provenance (B3), point these at migrated fixtures under `data/`.
:::

```{r setup}
suppressMessages({
  library(nilHMM)
  library(data.table)
  library(ggplot2)
})
for (f in list.files(here::here("R"), "\\.R$", full.names = TRUE)) source(f)
mnt <- bzea_mounts()
CAS <- mnt$cassini
MNT <- mnt$ancestry
# zealtiger working repo = current data source for the sweep inputs (B3 migrates these)
ZT <- pick(
  "/Users/fvrodriguez/Desktop/zealtiger",
  "/Users/fvrodriguez/repos/zealtiger"
)
```

```{r samples, eval=file.exists(file.path(ZT, "results/sim_calibration/coverage_sweep_members.csv"))}
mem <- fread(file.path(ZT, "results/sim_calibration/coverage_sweep_members.csv"))[sweep == "vary_skim"][order(vary_cov)]
mem[, tax := sub("\\..*", "", nil)]
tax2dir <- c(
  Zx = "mexicana/pangene", Zl = "luxurians/pangene", Zd = "diploperennis/pangene",
  Zv = "parviglumis/pangene", Zh = "huehuetenangensis/pangene"
)
mem[, adir := tax2dir[tax]]
wraplab <- function(cov, ped, skim, brb) sprintf("%s\n%s\nskim:%s\nbrb:%s", cov, ped, skim, brb)
rows <- rbind(
  data.table(
    rowlab = wraplab("B73 ctrl", "B73 (control)", "PN10_SID893", "PN3_SID213"),
    skim = "PN10_SID893", brb = "PN3_SID213", adir = "mexicana/pangene"
  ),
  mem[, .(
    rowlab = wraplab(sprintf("%.2f×", vary_cov), nil, skim_name, brb_name),
    skim = skim_name, brb = brb_name, adir
  )]
)
```

# ATLAS: competitive-mapping ancestry (GOOGA-shadow)

Per-gene hard call from competitive-mapping counts (recurrent ≥ .95 → REF, donor
≥ .95 → ALT, both ≥ .25 → HET, min 5 reads), placed at B73 coordinates, then
smoothed through the `gt`-emission geometric-duration HMM. Categorical (not
count) because 3′ RNA depth is expression-driven and ASE pushes HET off 0.5.

```{r atlas, eval=dir.exists(CAS)}
g2p <- fread(file.path(CAS, "data/pangene/gene_to_pangene.tsv"))[species == "B73", .(gene, pan_gene)]
co <- fread(file.path(CAS, "data/meta/b73_gene_coords.tsv"),
  header = FALSE, col.names = c("gene", "chr", "start", "end")
)
gmap <- merge(g2p, co, by = "gene")
gmap[, chr := as.integer(sub("chr", "", chr))]
gmap <- unique(gmap[!is.na(chr)], by = "pan_gene")[, .(pan_gene, chr, pos = end)]
gcall <- function(nr, nd) {
  nt <- nr + nd
  fr <- fifelse(nt > 0, nr / nt, 0)
  fd <- fifelse(nt > 0, nd / nt, 0)
  g <- rep(3L, length(nt))
  ok <- nt >= 5L
  g[ok & fr >= .95] <- 0L
  g[ok & fd >= .95] <- 2L
  g[ok & fr >= .25 & fd >= .25] <- 1L
  g
}
gtmodel <- nilHMM:::fit(
  list(), emission_gt(), duration_geometric(1e-3), design_priors("BC2S3")
)
atlas_seg <- function(brb, adir) {
  f <- file.path(CAS, "results", adir, paste0(brb, ".pangene_counts.tsv"))
  if (!file.exists(f)) {
    return(NULL)
  }
  ct <- merge(fread(f), gmap, by = "pan_gene")
  ct[, g := gcall(n_recur, n_donor)]
  ct <- ct[order(chr, pos)]
  rbindlist(lapply(sort(unique(ct$chr)), function(cc) {
    d <- ct[chr == cc]
    p <- nilHMM:::decode(gtmodel, list(g = d$g))
    r <- rle(p)
    e <- cumsum(r$lengths)
    s <- c(1L, head(e, -1) + 1L)
    data.table(chr = cc, start_bp = d$pos[s], end_bp = d$pos[e], state = r$values)
  }))
}
```

# binhmm on the dense skim bins

3-state Gaussian-emission HMM on per-bin `ALT_FREQ`, REF anchored at the
coverage floor (hard-EM with REF fixed, sticky transitions, collapse guards).

```{r binhmm, eval=dir.exists(MNT)}
binhmm_gauss <- function(skim, design = "BC2S2", stay = 0.995) {
  f <- file.path(MNT, paste0(skim, "_bin_genotypes.tsv"))
  if (!file.exists(f)) {
    return(NULL)
  }
  d <- fread(f)
  dt <- data.table(
    chr = as.integer(sub("chr", "", d$CONTIG)), bin = d$BIN_POS,
    af = d$ALT_FREQ, s = d$BIN_START, e = d$BIN_END
  )[order(chr, bin)]
  af <- dt$af
  pri <- design_priors(design)
  START <- c(1 - pri$f_1 - pri$f_2, pri$f_1, pri$f_2)
  trans <- matrix((1 - stay) / 2, 3, 3)
  diag(trans) <- stay
  seg_all <- function(state) {
    rbindlist(lapply(sort(unique(dt$chr)), function(cc) {
      i <- which(dt$chr == cc)
      i <- i[order(dt$bin[i])]
      p <- state[i]
      rr <- rle(p)
      rr$values[rr$values > 0L & rr$lengths < 2L] <- 0L
      p <- inverse.rle(rr)
      r <- rle(p)
      e <- cumsum(r$lengths)
      s <- c(1L, head(e, -1) + 1L)
      data.table(chr = cc, start_bp = dt$s[i][s], end_bp = dt$e[i][e], state = r$values)
    }))
  }
  base_bins <- af[af <= stats::quantile(af, 0.6)]
  ref_mean <- mean(base_bins)
  sd_base <- max(1e-3, sd(base_bins))
  hi <- af[af > ref_mean + 4 * sd_base]
  if (length(hi) < 4) {
    return(seg_all(rep(0L, length(af))))
  }
  alt_mean <- as.numeric(stats::quantile(hi, 0.75))
  het_mean <- ref_mean + (alt_mean - ref_mean) / 2
  sd_ref <- max(sd_base, (alt_mean - ref_mean) / 6)
  means <- c(ref_mean, het_mean, alt_mean)
  sds <- c(sd_ref, max(sd_ref, sd(hi)), max(sd_ref, sd(hi)))
  emit <- function(m, s) {
    vapply(1:3, function(k) stats::dnorm(af, m[k], max(s[k], 1e-3), log = TRUE), numeric(length(af)))
  }
  path <- integer(length(af))
  for (it in 1:8) {
    le <- emit(means, sds)
    np <- integer(length(af))
    for (cc in unique(dt$chr)) {
      i <- which(dt$chr == cc)
      i <- i[order(dt$bin[i])]
      np[i] <- viterbi_log_cpp(log(START), log(trans), le[i, , drop = FALSE])
    }
    if (identical(np, path)) break
    path <- np
    for (k in 2:3) {
      if (any(path == k - 1L)) {
        means[k] <- mean(af[path == k - 1L])
        sds[k] <- max(sd_ref, sd(af[path == k - 1L]))
      }
    }
  }
  if (means[2] < ref_mean + 4 * sd_ref) path[path == 1L] <- 0L
  if (means[3] < ref_mean + 4 * sd_ref) path[path == 2L] <- 0L
  seg_all(path)
}
```

# Count / RTIGER via the engine

```{r engine}
eng <- function(path, ...) {
  if (is.na(path) || !file.exists(path)) {
    return(NULL)
  }
  seg <- as.data.table(call_ancestry(read_counts(path), ...))
  seg[, .(chr, start_bp, end_bp, state)]
}
SK50 <- file.path(ZT, "data/rtiger_50K")
BRBW <- file.path(ZT, "results/sim_calibration/brbseq_ks_wideseq/counts")
```

# Assemble + paint

```{r paint, fig.width=16, fig.height=11, dpi=130, eval=exists("rows")}
methods <- c("Skim-binhmm", "Skim-nNIL", "Skim-rtiger", "BrB-nNIL", "BrB-rtiger", "BrB-atlas")
grab <- function(fn) {
  tryCatch(fn, error = function(e) {
    message("  ! ", conditionMessage(e))
    NULL
  })
}
all <- list()
for (i in seq_len(nrow(rows))) {
  rl <- rows$rowlab[i]
  sk <- rows$skim[i]
  br <- rows$brb[i]
  ad <- rows$adir[i]
  sk_counts <- find1(SK50, sk)
  brb_counts <- file.path(BRBW, paste0(br, ".tsv"))
  tr <- list(
    `Skim-binhmm` = grab(binhmm_gauss(sk)),
    `Skim-nNIL` = grab(eng(sk_counts, caller = "nnil", design = "BC2S2", r = 1e-4, err = 0.01)),
    `Skim-rtiger` = grab(eng(sk_counts, caller = "rtiger", r = 5L)),
    `BrB-nNIL` = grab(eng(brb_counts, caller = "nnil", design = "BC2S3", r = 1e-8, fit_means = TRUE)),
    `BrB-rtiger` = grab(eng(brb_counts, caller = "rtiger", r = 7L)),
    `BrB-atlas` = grab(atlas_seg(br, ad))
  )
  for (m in names(tr)) {
    if (!is.null(tr[[m]]) && nrow(tr[[m]])) {
      all[[length(all) + 1L]] <- tr[[m]][, `:=`(rowlab = rl, track = m)]
    }
  }
}
df <- rbindlist(all)
df[, rowlab := factor(rowlab, levels = c(rows$rowlab[1], rev(rows$rowlab[-1])))]
p <- paint_tracks(
  df,
  title = "nilHMM sanity paint — vary-skim sweep (B73 control + NILs)",
  subtitle = "Supplementary QC: 2 sources x nilHMM callers | all nilHMM, no Julia"
)
dir.create(here::here("results"), showWarnings = FALSE)
fwrite(df, here::here("results/sanity_paint_segments.tsv"), sep = "\t")
p
```

# Sanity read

- The B73 control paints ~all REF (clean control → clean paint).
- NIL donor blocks land on the same chromosomes across independent callers and
  both data sources — the coherence result.
- Caller-specific behavior as expected: binhmm broadest/most permissive;
  BrB-rtiger noisiest (RNA false-donor floor); ATLAS/BrB-nNIL most conservative.
