---
title: "Genome coverage theory: the Poisson model of missing data"
subtitle: "Missing-data modelling in BZea, part 1 of 3 — the coverage baseline"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**Purpose.** Establish the theoretical baseline for reasoning about missing
genotype data in a low-coverage sequencing population. Before we fit any model
to the BZea data (`missing-data-floor-model.qmd`) or compare fits across
datasets (`missing-data-model-comparison.qmd`), we need the expectation that a
purely random, Poisson sampling of reads would produce. Everything downstream is
measured as a *departure* from this baseline.

::: {.callout-note title="Where this sits in the series"}
1. **This note** — Poisson coverage theory and the missing-data expectation it implies.
2. `missing-data-floor-model.qmd` — the exponential-with-floor model, fit to BZea BAM alignments.
3. `missing-data-model-comparison.qmd` — the floor model applied to three BZea datasets.
:::

# Coverage calculation

## Theory

Ewens & Grant (2005) describe the coverage of a genome as a Poisson process. The
probability of observing $x$ reads at a given site is

$$P(x) = \frac{\lambda^x e^{-\lambda}}{x!}$$

where:

- $x$ is the number of short reads covering a given site on the genome,
- $\lambda = \dfrac{LN}{G}$ is the sequencing redundancy (mean coverage),

and:

- $G$ is the genome length (in bp),
- $L$ is the short-read average length,
- $N$ is the number of short reads sequenced.

## For the BZea population

The NAM v5 B73 genome is about 2.2 Gb:

```{r genome-size}
G <- 2.2e9
```

The mean read length is 150 bp:

```{r read-length}
L <- 150
```

The nominal coverage for BZea is 0.8:

```{r lambda}
lambda <- 0.8
```

Then the implied number of sequenced reads per sample is:

```{r n-reads}
N <- G * lambda / L
N / 1e6
```

i.e. each sample has ~12 million reads. Checking the other direction:

```{r coverage-check}
12e6 * 150 / 2.2e9
```

If 1400 samples were sequenced:

```{r total-reads}
n <- 1400
total_reads <- n * 12e6
```

and a lane yields ~4e9 reads:

```{r lanes}
total_reads / 4e9
```

so roughly 4 lanes were sequenced.

**Fraction of the genome covered** (at least one read at a site) under the
Poisson model:

```{r frac-covered}
p <- 1 - exp(-lambda)
p
```

**Expected number of samples with at least one read** at each position:

```{r expected-covered-samples}
round(n * p, 0)
```

with standard deviation:

```{r sd-covered}
sqrt(n * p * (1 - p))
```

**Expected number of missing genotypes per position** (median):

```{r expected-missing}
n - qbinom(0.50, n, p)
```

# Missing data

Expected fraction of missing data per position:

```{r frac-missing}
1 - qbinom(0.50, n, p) / n
```

# Probability of extreme missingness at a site

Now consider a specific question that matters for filtering: **what is the
probability that a site has at least 98% missing data?**

We assume a cohort of **1400** individuals.

```{r params}
n <- 1400 # number of individuals
p_missing <- 0.98 # missing-data proportion threshold
k_missing <- ceiling(n * p_missing) # number of missing genotypes at threshold

n
p_missing
k_missing
```

We model the number of missing genotypes at a site as a
$\text{Binomial}(n, p_0)$, where $p_0$ is the (unknown) true per-genotype
missingness rate. For illustration, assume **random missingness** with
$p_0 = 0.5$:

```{r prob-high-missing}
p0 <- 0.5

# Probability of having >= k_missing missing genotypes at a site
prob_high_missing <- pbinom(k_missing - 1,
  size = n, prob = p0,
  lower.tail = FALSE
)
prob_high_missing
```

# Dispersion

For the Binomial, the dispersion (variance-to-mean ratio) is

$$\frac{\text{Variance}}{\text{Mean}} = \frac{n p (1-p)}{n p} = 1 - p$$

```{r dispersion}
dispersion <- (n * p0 * (1 - p0)) / (n * p0)
dispersion
```

# Notes

- If $p_0$ is estimated from the real data (mean missingness rate across sites),
  this probability will be more realistic than the illustrative $p_0 = 0.5$.
- For **overdispersed** data (extra variability beyond the Binomial), a
  Beta-Binomial model may be more appropriate.
- Critically, real sequencing data has an *irreducible* missingness floor that
  the pure Poisson model cannot capture — that is the motivation for the
  exponential-with-floor model developed in `missing-data-floor-model.qmd`.

# References

Ewens, W. J., & Grant, G. (2005). *Statistical methods in bioinformatics: An
introduction* (2nd ed.). Springer. <https://doi.org/10.1007/b137845>
