---
title: "Comparing missing-data estimates across three BZea datasets"
subtitle: "Missing-data modelling in BZea, part 3 of 3 — fitting the floor model to WGS, Wideseq, and SNP50K"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**Purpose.** Take the exponential-with-floor model developed in
`missing-data-floor-model.qmd` and evaluate how stable its parameters ($\pi$,
$k$) are when the *same* BZea samples are viewed through three different data
products: whole-genome `picard` WGS metrics, the 27 M-SNP Wideseq binned counts,
and a filtered 50 K SNP panel.

::: {.callout-note title="Where this sits in the series"}
1. `missing-data-coverage-theory.qmd` — Poisson coverage theory (the baseline).
2. `missing-data-floor-model.qmd` — the exponential-with-floor model, fit to BZea BAM alignments.
3. **This note** — the floor model applied to, and compared across, three BZea datasets.
:::

```{r setup}
suppressMessages({
  library(here)
  library(readr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggpubr)
  library(cowplot)
  library(minpack.lm)
  library(viridis)
  library(knitr)
})
```

## Helper functions

These wrap the per-dataset workflow from part 2 so the same fit and plots can be
applied to every dataset. (In the original notebook they were sourced from a
project helper file; they are reproduced here so this note is self-contained.)

```{r helpers}
# Fit the exponential-with-floor model to a per-sample data frame that already
# has columns `lambda` (mean coverage) and `missing_obs` (observed missingness).
# Returns the fit object, the fitted parameters, and the data with predictions.
fit_exp_floor_model <- function(per_sample) {
  per_sample <- per_sample %>%
    mutate(missing_poisson = exp(-lambda))

  dat <- per_sample %>%
    filter(
      is.finite(lambda), is.finite(missing_obs),
      missing_obs > 0, missing_obs < 1
    )

  # Starting values: linear fits for the floor intercept and the decay rate
  pi_est <- coef(lm(missing_obs ~ missing_poisson, data = per_sample))[1]
  k_est <- -coef(lm(log(missing_obs) ~ lambda, data = dat))[2]

  fit <- nlsLM(
    missing_obs ~ pi + (1 - pi) * exp(-k * lambda),
    data  = dat,
    start = list(pi = unname(pi_est), k = unname(k_est)),
    lower = c(0, 0),
    upper = c(1, 10)
  )

  params <- as.list(coef(fit))
  per_sample$missing_exp_floor <- predict(fit, newdata = per_sample)

  list(fit = fit, params = params, data = per_sample)
}

# Coverage histogram for a per-sample data frame (used as an inlay).
generate_coverage_histogram <- function(per_sample) {
  per_sample %>%
    ggplot(aes(x = lambda)) +
    xlab("Coverage") +
    ylab("Sample Count") +
    geom_histogram(aes(fill = after_stat(x))) +
    geom_vline(
      xintercept = mean(per_sample$lambda, na.rm = TRUE),
      color = "red"
    ) +
    scale_fill_viridis_c() +
    ggpubr::theme_classic2(base_size = 8) +
    theme(legend.position = "none")
}

# Validation plot: observed vs floor-model expected missingness, 1:1 line,
# floor as a dashed vertical line, with the coverage histogram as an inlay.
generate_floor_model_plot <- function(model_results, coverage_hist_plot,
                                      plot_title = "") {
  ps <- model_results$data
  pi <- model_results$params$pi
  xy_limits <- c(0.2, 1.0)

  scatter <- ps %>%
    ggplot(aes(x = missing_exp_floor, y = missing_obs, col = lambda)) +
    geom_point() +
    geom_vline(xintercept = pi, linetype = "dashed") +
    geom_abline(slope = 1, intercept = 0, color = "red") +
    scale_color_viridis_c(limits = c(0, 1.5)) +
    guides(col = guide_colorbar(title = "Coverage")) +
    scale_x_continuous(limits = xy_limits, breaks = (2:10) / 10) +
    scale_y_continuous(limits = xy_limits, breaks = (2:10) / 10) +
    coord_fixed(ratio = 1) +
    xlab("Expected Missing (Floor Model)") +
    ylab("Observed Missing") +
    ggtitle(plot_title) +
    ggpubr::theme_classic2(base_size = 12) +
    theme(legend.position = "top")

  cowplot::ggdraw() +
    cowplot::draw_plot(scatter) +
    cowplot::draw_plot(cowplot::as_grob(coverage_hist_plot),
      x = 0.55, y = 0.16, width = 0.2, height = 0.25
    )
}
```

# 1 Data loading and processing

We load and process three datasets. Each needs a slightly different route to the
same two metrics: mean coverage (`lambda`) and observed missingness
(`missing_obs`).

## 1.1 Dataset 1: WGS metrics summary

Derived from `WGSmetrics_summary.tsv` (`picard CollectWgsMetrics` output),
reduced to a per-sample summary by `scripts/make_missing_data_summaries.R`:
$\lambda = \texttt{MEAN\_COVERAGE}$, $\texttt{missing\_obs} = 1 - \texttt{PCT\_1X}$.

```{r wgs-load}
per_sample_wgs <- read_tsv(
  here("data/missing_data/wgs_per_sample.tsv"),
  show_col_types = FALSE
) %>%
  filter(is.finite(lambda) & is.finite(missing_obs))

head(per_sample_wgs)
```

## 1.2 Dataset 2: BZea Wideseq (27 M binned genotypes)

Derived from `all_samples_bin_genotypes.tsv` — the same source used in part 2,
via the shared per-sample summary from `scripts/make_missing_data_summaries.R`.

```{r wideseq-load}
per_sample_bzea <- read_tsv(
  here("data/missing_data/wideseq_per_sample.tsv"),
  show_col_types = FALSE
) %>%
  select(SAMPLE, lambda, missing_obs) %>%
  filter(is.finite(lambda) & is.finite(missing_obs))

head(per_sample_bzea)
```

## 1.3 Dataset 3: 50 K filtered SNPs

Derived from `allelic_counts50K.tsv` (per-site REF/ALT read counts), reduced to a
per-sample summary by `scripts/make_missing_data_summaries.R`.

```{r snp50k-load}
per_sample_snp50K <- read_tsv(
  here("data/missing_data/snp50k_per_sample.tsv"),
  show_col_types = FALSE
) %>%
  select(SAMPLE, lambda, missing_obs) %>%
  filter(is.finite(lambda) & is.finite(missing_obs))

head(per_sample_snp50K)
```

# 2 Model fitting

Apply `fit_exp_floor_model()` to all three processed datasets.

```{r fit-all}
wgs_model_results <- fit_exp_floor_model(per_sample_wgs)
bzea_model_results <- fit_exp_floor_model(per_sample_bzea)
snp50K_model_results <- fit_exp_floor_model(per_sample_snp50K)

cat("WGS Metrics Model Parameters:\n")
cat("Floor (pi):", round(wgs_model_results$params$pi, 3), "\n")
cat("Decay (k):", round(wgs_model_results$params$k, 3), "\n\n")

cat("BZea Binned Model Parameters:\n")
cat("Floor (pi):", round(bzea_model_results$params$pi, 3), "\n")
cat("Decay (k):", round(bzea_model_results$params$k, 3), "\n\n")

cat("SNP50K Model Parameters:\n")
cat("Floor (pi):", round(snp50K_model_results$params$pi, 3), "\n")
cat("Decay (k):", round(snp50K_model_results$params$k, 3), "\n")
```

# 3 Visual comparison of model fits

Generate the validation plots for all three models and display them
side-by-side.

```{r compare-plots}
hist_wgs <- generate_coverage_histogram(wgs_model_results$data)
hist_bzea <- generate_coverage_histogram(bzea_model_results$data)
hist_snp50K <- generate_coverage_histogram(snp50K_model_results$data)

plot_wgs <- generate_floor_model_plot(
  model_results = wgs_model_results,
  coverage_hist_plot = hist_wgs,
  plot_title = "Whole Genome (0.8X goal)"
)

plot_bzea <- generate_floor_model_plot(
  model_results = bzea_model_results,
  coverage_hist_plot = hist_bzea,
  plot_title = "Wideseq (27M Teosinte SNPs)"
)

plot_snp50K <- generate_floor_model_plot(
  model_results = snp50K_model_results,
  coverage_hist_plot = hist_snp50K,
  plot_title = "Filtered SNPs (50K)"
)

combined_plot <- plot_grid(plot_wgs, plot_bzea, plot_snp50K, ncol = 3)
print(combined_plot)
```

# 4 Summary statistics comparison

```{r summary-table}
summary_stats <- data.frame(
  Dataset = c("WGS", "Wideseq", "SNP50K"),
  Samples = c(
    nrow(per_sample_wgs), nrow(per_sample_bzea),
    nrow(per_sample_snp50K)
  ),
  Mean_Coverage = c(
    round(mean(per_sample_wgs$lambda), 2),
    round(mean(per_sample_bzea$lambda), 2),
    round(mean(per_sample_snp50K$lambda), 2)
  ),
  Mean_Missingness = c(
    round(mean(per_sample_wgs$missing_obs), 3),
    round(mean(per_sample_bzea$missing_obs), 3),
    round(mean(per_sample_snp50K$missing_obs), 3)
  ),
  Floor_Pi = c(
    round(wgs_model_results$params$pi, 3),
    round(bzea_model_results$params$pi, 3),
    round(snp50K_model_results$params$pi, 3)
  ),
  Decay_k = c(
    round(wgs_model_results$params$k, 3),
    round(bzea_model_results$params$k, 3),
    round(snp50K_model_results$params$k, 3)
  )
)

knitr::kable(
  summary_stats,
  caption = "Comparison of Model Parameters Across Datasets",
  col.names = c(
    "Dataset", "N Samples", "Mean Coverage", "Mean Missingness",
    "Floor (pi)", "Decay (k)"
  )
)
```

The fitted parameters, for reference:

| Dataset | N Samples | Mean Coverage | Mean Missingness | Floor ($\pi$) | Decay ($k$) |
| ------- | --------- | ------------- | ---------------- | ------------- | ----------- |
| WGS     | 1437      | 0.39          | 0.744            | 0.446         | 1.661       |
| Wideseq | 1434      | 0.59          | 0.670            | 0.346         | 1.256       |
| SNP50K  | 1439      | 0.43          | 0.703            | 0.161         | 1.042       |

# 5 Interpretation

The floor $\pi$ moves in the expected direction with SNP ascertainment: the
unfiltered whole-genome view carries the highest irreducible missingness
($\pi = 0.446$), the 27 M teosinte-variant Wideseq panel sits in the middle
($\pi = 0.346$), and the curated 50 K SNP panel — chosen to be well-behaved and
mappable — has the lowest floor ($\pi = 0.161$). Filtering to informative,
mappable sites removes much of the irreducible missingness, exactly as the
mechanistic interpretation of $\pi$ (repetitive/structural/unmappable sequence)
predicts. The decay rate $k$ likewise decreases toward 1 (the Poisson value) as
the panel is cleaned, consistent with the filtered data behaving more like the
idealized random-sampling baseline from `missing-data-coverage-theory.qmd`.
