---
title: "Comparing missing-data estimates across three BZea datasets"
subtitle: "Missing-data modelling in BZea, part 3 of 3 — fitting the floor model to WGS, Wideseq, and SNP50K"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**Purpose.** Take the exponential-with-floor model developed in
`missing-data-floor-model.qmd` and evaluate how stable its parameters ($\pi$,
$k$) are when the *same* BZea samples are viewed through three different data
products: whole-genome `picard` WGS metrics, the 27 M-SNP Wideseq binned counts,
and a filtered 50 K SNP panel.

::: {.callout-note title="Where this sits in the series"}
1. `missing-data-coverage-theory.qmd` — Poisson coverage theory (the baseline).
2. `missing-data-floor-model.qmd` — the exponential-with-floor model, fit to BZea BAM alignments.
3. **This note** — the floor model applied to, and compared across, three BZea datasets.
:::

```{r setup}
suppressMessages({
  library(here)
  library(readr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggpubr)
  library(cowplot)
  library(minpack.lm)
  library(viridis)
  library(knitr)
})
```

## Helper functions

These wrap the per-dataset workflow from part 2 so the same fit and plots apply
to every dataset. They live in `R/missing_data_models.R` (ported verbatim from
the original `modeling_functions.R`) and are sourced here, so the fit and the
figure share one canonical definition and cannot drift from a retyped copy.

```{r helpers}
source(here("R/missing_data_models.R"))
```

# 1 Data loading and processing

We load and process three datasets. Each needs a slightly different route to the
same two metrics: mean coverage (`lambda`) and observed missingness
(`missing_obs`).

## 1.1 Dataset 1: WGS metrics summary

Derived from `WGSmetrics_summary.tsv` (`picard CollectWgsMetrics` output),
reduced to a per-sample summary by `scripts/make_missing_data_summaries.R`:
$\lambda = \texttt{MEAN\_COVERAGE}$, $\texttt{missing\_obs} = 1 - \texttt{PCT\_1X}$.

```{r wgs-load}
per_sample_wgs <- read_tsv(
  here("data/missing_data/wgs_per_sample.tsv"),
  show_col_types = FALSE
) %>%
  filter(is.finite(lambda) & is.finite(missing_obs))

head(per_sample_wgs)
```

## 1.2 Dataset 2: BZea Wideseq (27 M binned genotypes)

Derived from `all_samples_bin_genotypes.tsv` — the same source used in part 2,
via the shared per-sample summary from `scripts/make_missing_data_summaries.R`.

```{r wideseq-load}
per_sample_bzea <- read_tsv(
  here("data/missing_data/wideseq_per_sample.tsv"),
  show_col_types = FALSE
) %>%
  select(SAMPLE, lambda, missing_obs) %>%
  filter(is.finite(lambda) & is.finite(missing_obs))

head(per_sample_bzea)
```

## 1.3 Dataset 3: 50 K filtered SNPs

Derived from `allelic_counts50K.tsv` (per-site REF/ALT read counts), reduced to a
per-sample summary by `scripts/make_missing_data_summaries.R`.

```{r snp50k-load}
per_sample_snp50K <- read_tsv(
  here("data/missing_data/snp50k_per_sample.tsv"),
  show_col_types = FALSE
) %>%
  select(SAMPLE, lambda, missing_obs) %>%
  filter(is.finite(lambda) & is.finite(missing_obs))

head(per_sample_snp50K)
```

# 2 Model fitting

Apply `fit_exp_floor_model()` to all three processed datasets.

```{r fit-all}
wgs_model_results <- fit_exp_floor_model(per_sample_wgs)
bzea_model_results <- fit_exp_floor_model(per_sample_bzea)
snp50K_model_results <- fit_exp_floor_model(per_sample_snp50K)

cat("WGS Metrics Model Parameters:\n")
cat("Floor (pi):", round(wgs_model_results$params$pi, 3), "\n")
cat("Decay (k):", round(wgs_model_results$params$k, 3), "\n\n")

cat("BZea Binned Model Parameters:\n")
cat("Floor (pi):", round(bzea_model_results$params$pi, 3), "\n")
cat("Decay (k):", round(bzea_model_results$params$k, 3), "\n\n")

cat("SNP50K Model Parameters:\n")
cat("Floor (pi):", round(snp50K_model_results$params$pi, 3), "\n")
cat("Decay (k):", round(snp50K_model_results$params$k, 3), "\n")
```

# 3 Visual comparison of model fits

Generate the validation plots for all three models and display them
side-by-side.

```{r compare-plots, fig.width=18, fig.height=6}
hist_wgs <- generate_coverage_histogram(wgs_model_results$data)
hist_bzea <- generate_coverage_histogram(bzea_model_results$data)
hist_snp50K <- generate_coverage_histogram(snp50K_model_results$data)

plot_wgs <- generate_floor_model_plot(
  model_results = wgs_model_results,
  coverage_hist_plot = hist_wgs,
  plot_title = "Whole Genome (0.8X goal)"
)

plot_bzea <- generate_floor_model_plot(
  model_results = bzea_model_results,
  coverage_hist_plot = hist_bzea,
  plot_title = "Wideseq (27M Teosinte SNPs)"
)

plot_snp50K <- generate_floor_model_plot(
  model_results = snp50K_model_results,
  coverage_hist_plot = hist_snp50K,
  plot_title = "Filtered SNPs (50K)"
)

combined_plot <- plot_grid(plot_wgs, plot_bzea, plot_snp50K, ncol = 3)
print(combined_plot)
```

# 4 Summary statistics comparison

```{r summary-table}
summary_stats <- data.frame(
  Dataset = c("WGS", "Wideseq", "SNP50K"),
  Samples = c(
    nrow(per_sample_wgs), nrow(per_sample_bzea),
    nrow(per_sample_snp50K)
  ),
  Mean_Coverage = c(
    round(mean(per_sample_wgs$lambda), 2),
    round(mean(per_sample_bzea$lambda), 2),
    round(mean(per_sample_snp50K$lambda), 2)
  ),
  Mean_Missingness = c(
    round(mean(per_sample_wgs$missing_obs), 3),
    round(mean(per_sample_bzea$missing_obs), 3),
    round(mean(per_sample_snp50K$missing_obs), 3)
  ),
  Floor_Pi = c(
    round(wgs_model_results$params$pi, 3),
    round(bzea_model_results$params$pi, 3),
    round(snp50K_model_results$params$pi, 3)
  ),
  Decay_k = c(
    round(wgs_model_results$params$k, 3),
    round(bzea_model_results$params$k, 3),
    round(snp50K_model_results$params$k, 3)
  )
)

knitr::kable(
  summary_stats,
  caption = "Comparison of Model Parameters Across Datasets",
  col.names = c(
    "Dataset", "N Samples", "Mean Coverage", "Mean Missingness",
    "Floor (pi)", "Decay (k)"
  )
)
```

# 5 Interpretation

The floor $\pi$ moves in the expected direction with SNP ascertainment: the
unfiltered whole-genome view carries the highest irreducible missingness
($\pi = 0.446$), the 27 M teosinte-variant Wideseq panel sits in the middle
($\pi = 0.346$), and the curated 50 K SNP panel — chosen to be well-behaved and
mappable — has the lowest floor ($\pi = 0.161$). Filtering to informative,
mappable sites removes much of the irreducible missingness, exactly as the
mechanistic interpretation of $\pi$ (repetitive/structural/unmappable sequence)
predicts. The decay rate $k$ likewise decreases toward 1 (the Poisson value) as
the panel is cleaned, consistent with the filtered data behaving more like the
idealized random-sampling baseline from `missing-data-coverage-theory.qmd`.
