---
title: "BAM-alignment genome coverage and missing data in the ZEAL population"
subtitle: "Missing-data modelling in ZEAL, part 2 of 3 — the exponential-with-floor model"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**Purpose.** Extend the Poisson baseline from `missing-data-coverage-theory.qmd`
into a model that actually fits the ZEAL data. We analyze **BAM alignments
directly** — before any genotype calling or VCF filtering — and show that
per-sample missingness follows an *exponential-with-floor* law rather than the
pure Poisson expectation.

::: {.callout-note title="Where this sits in the series"}
1. `missing-data-coverage-theory.qmd` — Poisson coverage theory (the baseline).
2. **This note** — the exponential-with-floor model, fit to ZEAL BAM alignments.
3. `missing-data-model-comparison.qmd` — the floor model applied to three ZEAL datasets.
:::

```{r setup}
suppressMessages({
  library(here)
  library(readr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggpubr)
  library(cowplot)
  library(minpack.lm) # nlsLM: robust Levenberg-Marquardt nonlinear least squares
  library(viridis)
})
```

# 1 Introduction

This analysis examines genome coverage and the distribution of missing read data
in the **ZEAL** maize–teosinte population without calling genotypes or relying on
VCF files. Instead, we analyze the **BAM alignments** directly, prior to any
genotype calling. We used GATK's `CollectAllelicCounts` on the BAM files,
restricting read counting to sites corresponding to ~27 million known teosinte
variants (the `wideseq_ref` panel). For each site and sample the tool counted the
number of reads supporting each allele, and these counts were aggregated into
1 Mb bins for the initial file in the workflow.

By aggregating the per-bin read counts across the genome, we estimate coverage
($\lambda$) and compare the **observed fraction of missing genotype calls**
against the **expected missing fraction** under the Poisson coverage model.

## 1.1 Theoretical background

Modeling missingness in sequencing data involves both stochastic sampling effects
and systematic biases. The foundational concept is the Poisson coverage model,
where the probability of observing zero reads at a site decreases exponentially
with increasing coverage — this reflects the stochastic nature of read sampling.

Additionally, certain genomic regions inherently exhibit higher missingness due
to low-complexity/repetitive sequences or structural variation. This leads to an
irreducible baseline missingness that persists even at high coverage with short
reads. These two facts are the basis for the **exponential-with-floor** model:
a parsimonious and mechanistically interpretable summary of missingness across
samples.

## 1.2 Models of missingness

### 1.2.1 Poisson model

If reads are randomly and independently distributed across the genome, the
probability of missing a site is

$$P_{\text{missing}} = e^{-\lambda}$$

where $\lambda$ is the mean coverage (average reads per variant site) and
$P_{\text{missing}}$ is the probability of no reads covering a site. This serves
as a baseline to evaluate whether missingness arises purely from random
sequencing variation or whether additional biases (e.g. mapping issues)
contribute.

### 1.2.2 Exponential-with-floor model

We model per-sample missingness as

$$P_{\text{missing}} = \pi + (1 - \pi)\, e^{-k \lambda}$$

where:

- $\lambda$ is the mean coverage.
- $\pi$ (**floor parameter**) is the fraction of sites that remain uncalled even
  at high coverage with short reads. This baseline missingness likely corresponds
  to repetitive DNA, structural variants, or highly polymorphic regions where
  short reads fail to map or are discarded by mapping filters. For example,
  `picard CollectWgsMetrics` reports a `PCT_EXC_TOTAL` statistic that can be
  interpreted as a floor parameter — for the *alux* project, at nominal coverage
  $3.2\times$ I obtained a mean $\text{PCT\_EXC\_TOTAL} = 0.45 \sim \pi$.
- $k$ (**decay rate**) is how fast missingness decays with increasing coverage
  relative to the Poisson model. I interpret the product $k\lambda$ as an
  *effective coverage* (perhaps the coverage of the informative variants).

# 2 Data loading and processing

The input is a small per-sample summary of the aggregated 1 Mb bin allele-count
table (`all_samples_bin_genotypes.tsv`), reduced to per-sample variant/depth
counts by `scripts/make_missing_data_summaries.R` (run once on-mount; see
`DATA.md`). The notebook reads only this summary, so it renders with no mount
attached. The genotype **calls** in the source file are irrelevant here — we use
only depth and variant counts.

```{r data-load}
summary_file <- here("data/missing_data/wideseq_per_sample.tsv")
stopifnot(file.exists(summary_file))
per_sample <- read_tsv(summary_file, show_col_types = FALSE) %>%
  mutate(missing_poisson = exp(-lambda))
```

```{r per-sample}
# Overall (pooled) population statistics
overall_stats <- per_sample %>%
  summarize(
    VARIANT_COUNT             = sum(VARIANT_COUNT, na.rm = TRUE),
    INFORMATIVE_VARIANT_COUNT = sum(INFORMATIVE_VARIANT_COUNT, na.rm = TRUE),
    DEPTH_SUM                 = sum(DEPTH_SUM, na.rm = TRUE)
  ) %>%
  mutate(
    lambda          = DEPTH_SUM / VARIANT_COUNT,
    missing_obs     = 1 - INFORMATIVE_VARIANT_COUNT / VARIANT_COUNT,
    missing_poisson = exp(-lambda)
  )

overall_stats
```

The population-level mean coverage is $\lambda \approx 0.59$, with an observed
missingness of $\approx 0.67$ — already above the Poisson expectation of
$\approx 0.554$.

# 3 Coverage distribution analysis

```{r coverage-hist}
lambda_bar <- overall_stats$lambda
p_hist <- per_sample %>%
  ggplot(aes(x = lambda)) +
  xlab("Coverage") +
  ylab("Sample Count") +
  geom_histogram(aes(fill = after_stat(x))) +
  geom_vline(xintercept = lambda_bar, color = "red") +
  scale_fill_viridis_c() +
  ggpubr::theme_classic2(base_size = 8) +
  theme(legend.position = "none")

print(p_hist)
```

The coverage distribution has a mean of ~0.59 with substantial spread across
samples.

# 4 Model fitting

## 4.1 Simple exponential model

A first pass: regress log-missingness on coverage to get a rough decay rate.

```{r fit-linear}
fit_linear <- lm(log(missing_obs) ~ lambda, data = per_sample)
per_sample$missing_exp_k <- exp(predict(fit_linear))

k_est <- -coef(fit_linear)[2]
cat("Estimated k:", round(k_est, 3), "\n")

cat(
  "Effective coverage (k*lambda) range:",
  paste0(round(k_est * range(per_sample$lambda), 3), collapse = " - "), "\n"
)
```

## 4.2 Exponential-with-floor model

Now the full model, fit by robust nonlinear least squares (`nlsLM`), using the
linear-model estimates as starting values.

```{r fit-floor}
# Restrict to samples with well-defined missingness in (0, 1)
dat <- per_sample %>%
  filter(
    is.finite(lambda), is.finite(missing_obs),
    missing_obs > 0, missing_obs < 1
  )

# Initial estimate of the floor from a linear fit of observed vs Poisson
fit_missing <- lm(missing_obs ~ missing_poisson, data = per_sample)
pi_est <- coef(fit_missing)[1]

fit_exp_floor <- nlsLM(
  missing_obs ~ pi + (1 - pi) * exp(-k * lambda),
  data  = dat,
  start = list(pi = pi_est, k = k_est),
  lower = c(0, 0),
  upper = c(1, 10)
)

model_params <- coef(fit_exp_floor)
pi <- model_params[1]
k <- model_params[2]

cat("Floor parameter (pi):", round(pi, 3), "\n")
cat("Decay rate (k):", round(k, 3), "\n")

per_sample$missing_exp_floor <- predict(fit_exp_floor, newdata = per_sample)
overall_stats$missing_exp_floor <- predict(
  fit_exp_floor,
  newdata = data.frame(lambda = overall_stats$lambda)
)
```

# 5 Model comparison

```{r decay-plot}
decay_plot <- per_sample %>%
  select(-missing_obs) %>%
  ungroup() %>%
  tidyr::pivot_longer(
    cols = starts_with("missing_"),
    names_to = "model", values_to = "pred",
    names_prefix = "missing_"
  ) %>%
  ggplot(aes(x = lambda, y = pred, col = model, group = model)) +
  geom_point(
    data = per_sample, aes(x = lambda, y = missing_obs),
    inherit.aes = FALSE, alpha = 0.3
  ) +
  geom_line(linewidth = 1) +
  xlab("Coverage") +
  ylab("Missing Fraction") +
  geom_point(x = lambda_bar, y = overall_stats$missing_obs, color = "red") +
  ggpubr::theme_classic2(base_size = 16) +
  theme(legend.position = c(0.2, 0.25), legend.box = "horizontal")

print(decay_plot)
```

# 6 Model validation

Common scales for the validation scatter plots.

```{r validation-scales}
xy_limits <- c(0.2, 1.0)
legend_breaks <- c(0, 0.5, 1.0, 1.5)
legend_labels <- as.character(legend_breaks)
```

## 6.1 Poisson model assessment

```{r poisson-scatter}
p_scatter_poisson <- per_sample %>%
  ggplot(aes(x = missing_poisson, y = missing_obs, col = lambda)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  scale_color_viridis_c(
    breaks = legend_breaks, labels = legend_labels,
    limits = c(0, 1.5)
  ) +
  guides(col = guide_colorbar(title = "Coverage")) +
  geom_point(
    x = overall_stats$missing_poisson, y = overall_stats$missing_obs,
    color = "red"
  ) +
  scale_x_continuous(limits = xy_limits, breaks = (2:10) / 10) +
  scale_y_continuous(limits = xy_limits, breaks = (2:10) / 10) +
  coord_fixed(ratio = 1) +
  xlab("Expected Missing (Poisson)") +
  ylab("Observed Missing") +
  ggpubr::theme_classic2(base_size = 16) +
  theme(legend.position = "top")

inlay_grob <- cowplot::as_grob(p_hist)

final_plot_poisson <- cowplot::ggdraw() +
  cowplot::draw_plot(p_scatter_poisson) +
  cowplot::draw_plot(inlay_grob, x = 0.55, y = 0.16, width = 0.2, height = 0.25)

print(final_plot_poisson)
```

The Poisson model systematically **underpredicts** missingness: observed points
sit above the 1:1 line.

## 6.2 Exponential-with-floor model assessment

```{r floor-scatter}
p_scatter_floor <- per_sample %>%
  ggplot(aes(x = missing_exp_floor, y = missing_obs, col = lambda)) +
  geom_point() +
  geom_vline(xintercept = pi, linetype = "dashed") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  scale_color_viridis_c(
    breaks = legend_breaks, labels = legend_labels,
    limits = c(0, 1.5)
  ) +
  guides(col = guide_colorbar(title = "Coverage")) +
  geom_point(
    x = overall_stats$missing_exp_floor,
    y = overall_stats$missing_obs, color = "red"
  ) +
  scale_x_continuous(limits = xy_limits, breaks = (2:10) / 10) +
  scale_y_continuous(limits = xy_limits, breaks = (2:10) / 10) +
  coord_fixed(ratio = 1) +
  xlab("Expected Missing (Floor Model)") +
  ylab("Observed Missing") +
  ggpubr::theme_classic2(base_size = 16) +
  theme(legend.position = "top")

final_plot_floor <- cowplot::ggdraw() +
  cowplot::draw_plot(p_scatter_floor) +
  cowplot::draw_plot(inlay_grob, x = 0.55, y = 0.16, width = 0.2, height = 0.25)

print(final_plot_floor)
```

The floor model places points on the 1:1 line; the dashed line marks the floor
$\pi$, the minimum missingness attainable no matter how high the coverage.

# 7 Results summary

```{r results-summary}
cat("Population-level Statistics:\n")
cat("Mean coverage (lambda):", round(overall_stats$lambda, 3), "\n")
cat("Observed missingness:", round(overall_stats$missing_obs, 3), "\n")
cat("Poisson expectation:", round(overall_stats$missing_poisson, 3), "\n")
cat("Floor model expectation:", round(overall_stats$missing_exp_floor, 3), "\n\n")

cat("Model Parameters:\n")
cat("Floor parameter (pi):", round(pi, 3), "\n")
cat("Decay rate (k):", round(k, 3), "\n")
```

The exponential-with-floor model substantially improves the fit over the simple
Poisson model. The floor parameter (~0.35) indicates that about a third of sites
remain uncallable even at high coverage — likely repetitive sequence, structural
variants, or mapping artifacts. The decay rate (~1.26) implies
slower-than-Poisson decay of missingness with increasing coverage.

Part 3 (`missing-data-model-comparison.qmd`) evaluates how stable $\pi$ and $k$
are when the same model is fit to three different ZEAL data products.

# 8 References

Li, H., Ruan, J., & Durbin, R. (2008). Mapping short DNA sequencing reads and
calling variants using mapping quality scores. *Bioinformatics*, 24(3), 223–230.

Nielsen, R., Paul, J. S., Albrechtsen, A., & Song, Y. S. (2011). Genotype and SNP
calling from next-generation sequencing data. *Nature Reviews Genetics*, 12(6),
443–451.
