---
title: "nilHMM sanity check paint: source × caller tracks, one engine, no Julia"
subtitle: "Chromosome painting of the vary-skim sweep NILs (+B73 control) — 6 tracks = 2 data sources (Skim, BRB) crossed with nilHMM callers, all through the nilHMM package"
author: "Fausto Rodríguez-Zapata"
date: "2026-07-02"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**Context.** This is the cross-track sanity check for the unified `nilHMM`
engine, where every caller runs through one duration-aware 3-state
(REF/HET/ALT) HMM with the emission as the only swappable piece. We paint the
same **vary-skim coverage-sweep NILs** (plus a clean B73 control) and eyeball
whether the independent tracks agree on where the donor blocks are, and whether
the control stays REF. Memory: `wideseq-chrpainting-code`,
`generative_vs_discriminative_callers`, `cassini-googa-caller`,
`method-naming-source-caller`, `skim-count-caller-calibration`,
`rtiger-betabinomial-cost`.

**The six tracks are *not* six callers.** Each track is a combination of a
**data source** and a **nilHMM caller method** (naming convention
`<DataSource>-<Caller>`, source first — memory `method-naming-source-caller`).
There are only **two data sources** and **four distinct callers**; Skim and BRB
each run three tracks and *share* the nNIL and RTIGER callers:

- **Data sources (2):** `Skim` = low-coverage WGS; `BrB` = 3′ RNA-seq (BRB-seq).
- **Callers (4 distinct):** `binhmm` (Gaussian on per-bin `ALT_FREQ`), `nNIL`
  (count / BetaBinomial HMM), `RTIGER` (Julia-free port), `ATLAS`
  (GOOGA-shadow gene-level GT).

The tracks are ordered **first by data source, then by caller** (all three Skim
tracks, then all three BRB tracks):

| # | Track | Data source | nilHMM caller | Input marker set |
|---|---|---|---|---|
| 1 | **Skim-binhmm** | Skim (low-cov WGS) | `binhmm` — 3-state Gaussian on per-bin `ALT_FREQ`, REF anchored at the floor | **dense ~27.6 M-SNP set** — per-1 Mb `ALT_FREQ` `bin_genotypes` from GATK `CollectAllelicCounts` at the full teosinte-informative wideseq positions (mount, `bzeaseq/ancestry`) |
| 2 | **Skim-nNIL** | Skim (low-cov WGS) | `nNIL` — count / BetaBinomial HMM (`r=1e-4`, `err=0.01`) | **50K panel** counts (thinned SNP50K) |
| 3 | **Skim-rtiger** | Skim (low-cov WGS) | `RTIGER` port (`r=5`) | **50K panel** counts (thinned SNP50K) |
| 4 | **BrB-nNIL** | BrB (3′ RNA-seq) | `nNIL` — count HMM, `fit_means` (`r=1e-8`) | **wideseq-thinned** BRB counts |
| 5 | **BrB-rtiger** | BrB (3′ RNA-seq) | `RTIGER` port (`r=7`) | **wideseq-thinned** BRB counts |
| 6 | **BrB-atlas** | BrB (3′ RNA-seq) | `ATLAS` — GOOGA-shadow gene-level GT (`≥.95`/`≥.25`, min 5 reads) → geometric-duration HMM | cassini **pangene** counts, per donor species |

So the grid is 2 sources × 3 caller-slots = 6 tracks, drawing on 4 unique
callers (`binhmm` is Skim-only, `ATLAS` is BRB-only, `nNIL`/`RTIGER` run on both).

::: {.callout-note title="Marker-set provenance — the tracks run at different SNP densities"}
The three Skim tracks are **not** all on the same markers. **Skim-binhmm runs on
the full dense set of ~27.6 M teosinte-informative SNPs** (2.0–4.0 M per
chromosome; total 27,603,777), whereas **Skim-nNIL / Skim-rtiger run on the much
sparser ~52 K "SNP50K" set**. These are two different marker sets with two
different provenances:

- **The ~27.6 M panel** (`wideseq_ref`) is a **teosinte-informative SNP set** —
  positions where teosinte (the donor) carries variation relative to B73 (the
  recurrent parent), so a read there is diagnostic of donor-vs-recurrent ancestry.
  That informativeness is the whole point of the panel; it is a **reference/filter
  set, not sample genotypes**. It is built from the **Schnable et al. 2023** *Zea*
  pan-genome VCF by keeping the teosinte/*Tripsacum* + B73 samples and biallelic
  SNPs — the MAF ≥ 0.05 cutoff is merely the mechanism that selects the
  donor-informative (species-common) sites
  (`bzeaseq/get_variants_from_schnable2023.sh`). GATK `CollectAllelicCounts`
  tallies each skim/BRB sample's reads at these positions — so the BRB count tracks
  (BrB-nNIL, BrB-rtiger) draw on the same panel; only Skim-binhmm keeps the full
  27.6 M density.
- **The ~52 K SNP50K set is an *intersection* — not a genotyping array, and not
  merely a coarse thinning of the 27.6 M panel.** Its unfiltered base is the
  **BZea NIL BAMs → GATK joint genotyping (N. Tandukar) + Beagle imputation**
  (196,120 SNPs); intersecting those genotyped sites with the 27.6 M
  teosinte-informative panel yields **51,991 SNPs** (so the 50K is itself
  teosinte-informative, being a subset of the panel). The Skim-nNIL / Skim-rtiger
  counts are reads tallied at those ~52 K sites. Here the teosinte panel is the
  **filter** and the
  BZea genotyping is the base — full chain + paths in
  `sawers-rellan-labs/BzeaSeq` → `docs/donor_id_jaccard.md`
  ("SNP50K & marker-set provenance").

The MAF ≥ 0.05 filter keeps donor-informative (species-common) sites, which
raises the donor dose and widens REF/HET/ALT separation —
see [`genotype_likelihoods_and_hmm` §5](genotype_likelihoods_and_hmm.html#what-survives-the-information-ceiling)
and [`snp50k_genotype_identifiability` §2.1/§2.6/§3](snp50k_genotype_identifiability.html#consequences).
:::

> **Parameters are illustrative, not fully re-calibrated.** This note is a
> coherence check across sources and callers, not a final call set.

---

# 1. Setup — engine, paths, samples

```{r setup}
suppressMessages({
  library(nilHMM)
  library(data.table)
  library(ggplot2)
})
# Inputs are staged locally under data/ by stage_sanity_check_paint.R (see DATA.md),
# organized by (source x input type); paths below point into that layout.
SS <- here::here("data/skimsweep") # the vary-skim cohort
SKIM <- file.path(SS, "skim") # skim counts (counts_50k/) + bins (bins/)
BRB <- file.path(SS, "brb") # brb counts (counts_wideseq/) + pangene (pangene/)
REF <- here::here("data/ref") # genome-level ATLAS gene maps

# samples: control + 10 vary-skim NILs (ordered by skim lambda)
mem <- fread(file.path(SS, "coverage_sweep_members.csv"))[sweep == "vary_skim"][order(vary_cov)]
mem[, tax := sub("\\..*", "", nil)]
tax2dir <- c(
  Zx = "mexicana", Zl = "luxurians", Zd = "diploperennis",
  Zv = "parviglumis", Zh = "huehuetenangensis"
)
mem[, adir := tax2dir[tax]]
# wrapped left-hand label: cov + FULL pedigree string, then skim/brb plate ids
wraplab <- function(cov, ped, skim, brb) sprintf("%s\n%s\nskim:%s\nbrb:%s", cov, ped, skim, brb)
rows <- rbind(
  data.table(
    rowlab = wraplab("B73 ctrl", "B73 (control)", "PN10_SID893", "PN3_SID213"),
    skim = "PN10_SID893", brb = "PN3_SID213", adir = "mexicana"
  ),
  mem[, .(
    rowlab = wraplab(sprintf("%.2f×", vary_cov), nil, skim_name, brb_name),
    skim = skim_name, brb = brb_name, adir
  )]
)
```

# 2. ATLAS: competitive-mapping ancestry (GOOGA-shadow)

**ATLAS** = **A**ncestry by **T**ranscript-**L**evel **A**lignment **S**cores.

**The basic principle is competitive mapping: count the reads that match each
founder genome.** Where the SNP-panel callers read an *allele fraction* at
predefined variant sites, ATLAS instead aligns each transcript read against
**both founder references at once** — B73 (the recurrent parent) and the
teosinte **donor** transcriptome — and lets the **alignment scores** decide
which parent that read came from. Per read, the best alignment score (with a
margin `k` on the score gap `ΔAS`) votes the read as **REF**, **DONOR**, or
**AMBIGUOUS**. A read that maps equally well to both parents landed on sequence
that is *identical* between them (conserved — at ~2 % B73-vs-donor divergence
this is ~98 % of positions), so it carries no ancestry information: **AMBIGUOUS
= non-informative → dropped** (it is *not* a HET call). Summing the informative
votes over a gene gives the two competitive-mapping counts, `n_recur`
(B73-matching reads) and `n_donor` (donor-matching reads); ancestry is then just
"which founder's genome captured the reads at this gene."

**Caveat — the "donor" reference is a species-level stand-in, not the actual
donor accession.** We do not have the genome/transcriptome of the specific
teosinte accession used in each cross, so the donor side of the competitive
alignment is a **representative reference for the donor's species** (per
`tax2dir`: mexicana, luxurians, diploperennis, parviglumis, huehuetenangensis).
A read is scored donor-vs-B73 against that proxy, not against the true parent —
the same reference-data gap as the SNP-panel accession ceiling
([`snp50k_genotype_identifiability` §2.6](snp50k_genotype_identifiability.html#the-site-selection-ceiling)):
sequence the accession carries but the species proxy lacks (or vice versa) is
mis-scored, so `n_donor` is only as good as the stand-in's match to the real donor.

**GOOGA is the template.** GOOGA (Fishman lab, *PLoS Genet* 2024,
[pgen.1011072](https://doi.org/10.1371/journal.pgen.1011072)) is a *Mimulus* F2
genetic-map-building pipeline whose ancestry step is **threshold-then-HMM**.
ATLAS is a "GOOGA-shadow": it ports that step onto the shared nilHMM engine —

1. **Per-gene hard call from the count fractions** (GOOGA `s2.py`, verbatim):
   `n_recur + n_donor < 5` → missing; recurrent fraction `≥ 0.95` → **REF** (AA);
   donor fraction `≥ 0.95` → **ALT / donor-hom** (BB); *both* parents `≥ 0.25`
   → **HET** (AB); otherwise missing. (Thresholds `0.95` / `0.25`, min 5 reads.)
2. **Place each gene at its B73 coordinate** and order it along the chromosome.
3. **Smooth with the engine's categorical genotype-error (`gt`) emission +
   geometric-duration HMM** — the same duration/rigidity layer as every other
   caller, but a *categorical confusion* emission instead of count/BetaBinomial.

Why categorical and not a count model: these are **3′ RNA (BRB-seq) reads**, so
a gene's read depth is driven by **expression**, not coverage, and
**allele-specific expression** pushes a true HET's donor fraction off 0.5 — both
confound a count/BetaBinomial emission. Discretizing to a hard genotype *first*
(exactly what GOOGA, itself an eQTL/ASE pipeline, does) makes the emission
**expression-agnostic**. Memory: `cassini-googa-caller`,
`feedback-reproduce-from-source`.

The code below is the consumer-side staging: read the cassini pangene counts per
donor species, apply the GOOGA hard call, map genes to B73 coordinates, and
decode each chromosome through the `gt`-emission geometric-duration HMM.

```{r atlas}
g2p <- fread(file.path(REF, "gene_to_pangene.tsv"))[species == "B73", .(gene, pan_gene)]
co <- fread(file.path(REF, "b73_gene_coords.tsv"),
  header = FALSE,
  col.names = c("gene", "chr", "start", "end")
)
gmap <- merge(g2p, co, by = "gene")
gmap[, chr := as.integer(sub("chr", "", chr))]
gmap <- unique(gmap[!is.na(chr)], by = "pan_gene")[, .(pan_gene, chr, pos = end)]
gcall <- function(nr, nd) {
  nt <- nr + nd
  fr <- fifelse(nt > 0, nr / nt, 0)
  fd <- fifelse(nt > 0, nd / nt, 0)
  g <- rep(3L, length(nt))
  ok <- nt >= 5L
  g[ok & fr >= .95] <- 0L
  g[ok & fd >= .95] <- 2L
  g[ok & fr >= .25 & fd >= .25] <- 1L
  g
}
gtmodel <- nilHMM:::fit(
  list(), emission_gt(), duration_geometric(1e-3), design_priors("BC2S3")
)
atlas_seg <- function(brb, adir) {
  f <- file.path(BRB, "pangene", adir, paste0(brb, ".pangene_counts.tsv"))
  if (!file.exists(f)) {
    return(NULL)
  }
  ct <- merge(fread(f), gmap, by = "pan_gene")
  ct[, g := gcall(n_recur, n_donor)]
  ct <- ct[order(chr, pos)]
  rbindlist(lapply(sort(unique(ct$chr)), function(cc) {
    d <- ct[chr == cc]
    p <- nilHMM:::decode(gtmodel, list(g = d$g))
    r <- rle(p)
    e <- cumsum(r$lengths)
    s <- c(1L, head(e, -1) + 1L)
    data.table(chr = cc, start_bp = d$pos[s], end_bp = d$pos[e], state = r$values)
  }))
}
```

# 3. binhmm (corrected) on the dense skim bins

3-state **Gaussian**-emission HMM on per-bin `ALT_FREQ`, porting the nilhifi
`chromosome_painting.R` fix: REF is **anchored at the baseline floor** (mean
fixed, sd from the low bins) so REF noise stays REF; HET/ALT are seeded from the
data and refined by hard-EM with REF fixed; sticky transitions + design start
priors + collapse guards. This replaces the free K=3 GMM (which always invents a
3rd cluster → HET over-call).

```{r binhmm}
binhmm_gauss <- function(skim, design = "BC2S2", stay = 0.995) {
  f <- file.path(SKIM, "bins", paste0(skim, "_bin_genotypes.tsv"))
  if (!file.exists(f)) {
    return(NULL)
  }
  d <- fread(f)
  dt <- data.table(
    chr = as.integer(sub("chr", "", d$CONTIG)), bin = d$BIN_POS,
    af = d$ALT_FREQ, s = d$BIN_START, e = d$BIN_END
  )[order(chr, bin)]
  af <- dt$af
  pri <- design_priors(design)
  START <- c(1 - pri$f_1 - pri$f_2, pri$f_1, pri$f_2)
  trans <- matrix((1 - stay) / 2, 3, 3)
  diag(trans) <- stay
  seg_all <- function(state) {
    rbindlist(lapply(sort(unique(dt$chr)), function(cc) {
      i <- which(dt$chr == cc)
      i <- i[order(dt$bin[i])]
      p <- state[i]
      rr <- rle(p)
      rr$values[rr$values > 0L & rr$lengths < 2L] <- 0L
      p <- inverse.rle(rr) # drop 1-bin donor specks
      r <- rle(p)
      e <- cumsum(r$lengths)
      s <- c(1L, head(e, -1) + 1L)
      data.table(chr = cc, start_bp = dt$s[i][s], end_bp = dt$e[i][e], state = r$values)
    }))
  }

  # bulk of the bins = REF baseline (NILs are mostly REF)
  base_bins <- af[af <= stats::quantile(af, 0.6)]
  ref_mean <- mean(base_bins)
  sd_base <- max(1e-3, sd(base_bins)) # anchor REF at the FLOOR (not 0)
  hi <- af[af > ref_mean + 4 * sd_base] # bins clearly above the floor
  if (length(hi) < 4) {
    # collapse guard: no donor signal -> call everything REF
    return(seg_all(rep(0L, length(af))))
  }

  alt_mean <- as.numeric(stats::quantile(hi, 0.75)) # seed ALT from data
  het_mean <- ref_mean + (alt_mean - ref_mean) / 2 # HET midway floor..ALT
  # Floor REF width to the signal scale: a hyper-tight baseline (high coverage)
  # must not push mildly-elevated single bins into donor (anti-fragmentation).
  sd_ref <- max(sd_base, (alt_mean - ref_mean) / 6)
  means <- c(ref_mean, het_mean, alt_mean)
  sds <- c(sd_ref, max(sd_ref, sd(hi)), max(sd_ref, sd(hi)))
  # per-bin log-emission: dnorm of alt_frac under each of the 3 state means
  emit <- function(m, s) {
    vapply(
      1:3,
      function(k) stats::dnorm(af, m[k], max(s[k], 1e-3), log = TRUE),
      numeric(length(af))
    )
  }
  path <- integer(length(af))
  for (it in 1:8) { # hard-EM: Viterbi -> refit HET/ALT means (REF fixed at 0)
    le <- emit(means, sds)
    np <- integer(length(af))
    for (cc in unique(dt$chr)) {
      i <- which(dt$chr == cc)
      i <- i[order(dt$bin[i])]
      np[i] <- nilHMM:::viterbi_log_cpp(log(START), log(trans), le[i, , drop = FALSE])
    }
    if (identical(np, path)) break
    path <- np
    for (k in 2:3) {
      if (any(path == k - 1L)) {
        means[k] <- mean(af[path == k - 1L])
        sds[k] <- max(sd_ref, sd(af[path == k - 1L]))
      }
    }
  }
  # collapse guards: fold a state whose mean isn't separated from the REF floor
  # HET indistinct from floor
  if (means[2] < ref_mean + 4 * sd_ref) path[path == 1L] <- 0L
  # ALT indistinct from floor
  if (means[3] < ref_mean + 4 * sd_ref) path[path == 2L] <- 0L
  seg_all(path)
}
```

# 4. Count / RTIGER via the engine

```{r engine}
find1 <- function(dir, s) {
  f <- list.files(dir, paste0("^", s, "\\.tsv$"), full.names = TRUE, recursive = TRUE)
  if (length(f)) f[1] else NA
}
eng <- function(path, ...) {
  if (is.na(path) || !file.exists(path)) {
    return(NULL)
  }
  seg <- as.data.table(call_ancestry(read_counts(path), ...))
  seg[, .(chr, start_bp, end_bp, state)]
}
# parent of counts/ (NILs) + counts_checks/B73/ (control); find1 recurses
SK50 <- file.path(SKIM, "counts_50k")
BRBW <- file.path(BRB, "counts_wideseq")
```

# 5. Assemble all tracks

```{r assemble}
methods <- c(
  "Skim-binhmm", "Skim-nNIL", "Skim-rtiger",
  "BrB-nNIL", "BrB-rtiger", "BrB-atlas"
)

# run a caller, returning NULL (with a message) instead of aborting on error
grab <- function(fn) {
  tryCatch(fn, error = function(e) {
    message("  ! ", conditionMessage(e))
    NULL
  })
}

all <- list()
for (i in seq_len(nrow(rows))) {
  rl <- rows$rowlab[i]
  sk <- rows$skim[i]
  br <- rows$brb[i]
  ad <- rows$adir[i]

  sk_counts <- find1(SK50, sk) # skim 50K panel counts
  brb_counts <- file.path(BRBW, paste0(br, ".tsv")) # wideseq-thinned BRB counts

  tr <- list(
    `Skim-binhmm` = grab(binhmm_gauss(sk)),
    `Skim-nNIL` = grab(eng(sk_counts,
      caller = "nnil", design = "BC2S2", r = 1e-4, err = 0.01
    )),
    `Skim-rtiger` = grab(eng(sk_counts, caller = "rtiger", r = 5L)),
    `BrB-nNIL` = grab(eng(brb_counts,
      caller = "nnil", design = "BC2S3", r = 1e-8, fit_means = TRUE
    )),
    `BrB-rtiger` = grab(eng(brb_counts, caller = "rtiger", r = 7L)),
    `BrB-atlas` = grab(atlas_seg(br, ad))
  )

  # keep the non-empty tracks, tagging each with its row label and method
  for (m in names(tr)) {
    if (!is.null(tr[[m]]) && nrow(tr[[m]])) {
      all[[length(all) + 1L]] <- tr[[m]][, `:=`(rowlab = rl, method = m)]
    }
  }
}
df <- rbindlist(all)
df[, .N, by = method]
```

# 6. Paint

```{r paint, fig.width=16, fig.height=11, dpi=130}
yb <- data.table(method = methods, k = 5:0) # top -> bottom
df <- merge(df, yb, by = "method")
df[, `:=`(ylo = k / 6 + 0.01, yhi = (k + 1) / 6 - 0.01)]
# B73 ctrl on top, then skim-cov descending (facet switch = 'y')
df[, rowlab := factor(rowlab, levels = c(rows$rowlab[1], rev(rows$rowlab[-1])))]
df[, chr := factor(chr, levels = 1:10, labels = paste0("chr", 1:10))]
df[, state := factor(state, 0:2, c("REF", "HET", "ALT"))]
pal <- c(REF = "gold", HET = "springgreen4", ALT = "purple4")
p <- ggplot(df, aes(xmin = start_bp / 1e6, xmax = end_bp / 1e6, ymin = ylo, ymax = yhi, fill = state)) +
  geom_rect() +
  facet_grid(rowlab ~ chr, scales = "free_x", space = "free_x", switch = "y") +
  scale_fill_manual(values = pal, drop = FALSE) +
  scale_y_continuous(
    breaks = (5:0) / 6 + 1 / 12, labels = methods, position = "right",
    expand = expansion(mult = 0.01)
  ) +
  labs(
    title = "nilHMM sanity check paint — vary-skim sweep (B73 control + 10 NILs)",
    subtitle = "6 tracks = data source × nilHMM caller (2 sources: Skim, BrB | 4 callers: binhmm, nNIL, RTIGER, ATLAS) | ordered by source then caller | all nilHMM, no Julia",
    x = "Position (Mb)", fill = "Genotype"
  ) +
  ggpubr::theme_classic2(base_size = 8) +
  theme(
    plot.title = element_text(size = 12),
    plot.subtitle = element_text(size = 12),
    legend.position = "bottom",
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    panel.spacing.x = unit(0.04, "lines"),
    panel.spacing.y = unit(0.15, "lines"),
    strip.background = element_blank(),
    strip.text.y.left = element_text(
      angle = 0, hjust = 1, size = 12, lineheight = 0.9
    ),
    axis.text.y.right = element_text(size = 11),
    axis.ticks.y = element_blank(),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
dir.create(here::here("results"), showWarnings = FALSE)
ggsave(here::here("results/nilhmm_sanity_check_paint.png"), p, width = 16, height = 11, dpi = 130, limitsize = FALSE)
fwrite(df, here::here("results/nilhmm_sanity_check_paint_segments.tsv"), sep = "\t")
p
```

# 7. Sanity read

- **The anchor holds:** the B73 control (top row) is essentially all-REF across
  all six tracks — a clean control paints clean.
- **NILs show localized donor blocks that broadly agree across callers.** The
  major introgressions land in the same chromosomes across independent callers
  and across the two data sources — the core sanity result.
- **Caller-specific behavior, as expected:** Skim-binhmm paints the broadest
  blocks and is more permissive toward ALT; BrB-rtiger is the noisiest (the
  known BRB-RTI RNA false-donor floor); BrB-atlas and BrB-nNIL are the most
  conservative.
