---
title: "nNIL reproduction benchmark: faithful port + calibration foil on Tao Zhong's population"
subtitle: "Supplement (optimization): nilHMM nNIL reproduces Holland's caller; chip-calibration vs simulation-calibration land in the same place"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

::: {.callout-important title="SCAFFOLD, numbers not yet measured"}
Scaffolded ahead of the live benchmark (Phase 1 of
`agent/nilhmm-draft-completion-plan.md`). Result tables/figures are placeholders
guarded by `file.exists()`. Prose and method are final; numbers are `[TO MEASURE]`.

**Data prerequisite (blocking):** the Zhong et al. 2025 nNIL population genotypes
(888 nNILs GBS + 24 SNP-chip re-genotyped truth) must be sourced before this note
can run. Confirm availability/repository with J. Holland; record provenance in
`DATA.md`. See [[teonam-genotype-source-deadend]] for the kind of dead-ends to
avoid, do not approximate with a lookalike panel.
:::

**Thesis (two parts).**

1. **Faithful port.** nilHMM's `nNIL` caller (categorical genotype-confusion
   emission + geometric duration; the `rrate` axis of the engine) is a
   reimplementation of Holland's original nested-NIL HMM \[Zhong 2025\]. Run on the
   same nNIL genotypes it should reproduce Holland's introgression calls.
2. **Calibration foil.** Zhong et al. set the HMM's parameters by optimizing
   against 24 of 888 lines re-genotyped on a SNP chip, the scarce-truth calibration
   the main text contrasts with. We reproduce that chip-calibrated operating point,
   then show the **simulation-calibrated** `rrate` (from `simcross`, no chip) lands
   at effectively the same segmentation. This is the concrete evidence for the
   paper's central claim that simulation substitutes for scarce truth.

::: {.callout-note title="Compute is separated from this note"}
Run by `scripts/bench_nnil_reproduction.R` (to be written): loads the nNIL panel,
runs nilHMM `nNIL`, compares against (a) the published Holland calls and (b) the
chip truth on the 24-line subset, then overlays the simulation-calibrated `rrate`
operating point. Writes CSVs to `results/bench/` (gitignored). This note loads and
plots them. Regenerate: `Rscript scripts/bench_nnil_reproduction.R --generate`.

Same offline/logging discipline as `rtiger-optimization-benchmark.qmd`: sources
`scripts/logging.R` with per-line ETA, `--smoke` before `--generate`, run persisted
via `tee agent/bench_nnil_<ts>_<pid>.log`. Rendering never triggers the run.
:::

```{r setup}
suppressMessages({
  library(data.table)
  library(ggplot2)
})
BENCH <- here::here("results/bench")
f <- function(x) file.path(BENCH, x)
have_repro <- file.exists(f("nnil_reproduction.csv"))
have_calib <- file.exists(f("nnil_chip_vs_sim_calibration.csv"))
```

## 1. Faithful reproduction of Holland's nNIL calls

Concordance of nilHMM `nNIL` against the published Zhong et al. introgression
calls on the 888-line panel: marker-level agreement and donor-segment overlap.

```{r repro-table}
if (have_repro) {
  rp <- fread(f("nnil_reproduction.csv"))
  knitr::kable(rp, caption = "nilHMM nNIL vs published Holland calls, marker concordance and donor-segment Jaccard.")
} else {
  cat(
    "[TO MEASURE] Reproduction concordance. Target: high marker concordance and",
    "donor-segment agreement, establishing the port is faithful (cf. reproduce-not-approximate)."
  )
}
```

## 2. Chip-calibration vs simulation-calibration

The heart of the foil. On the 24-line chip-truth subset, reproduce the operating
point Zhong et al. selected by chip optimization; then place the `rrate` chosen by
`simcross` calibration (no chip) on the same axis and show the resulting
segmentations coincide.

```{r calib-plot}
if (have_calib) {
  cc <- fread(f("nnil_chip_vs_sim_calibration.csv"))
  knitr::kable(cc, caption = "Chip-calibrated vs simulation-calibrated nNIL operating point and resulting segmentation agreement.")
} else {
  cat(
    "[TO MEASURE] Chip-calibrated rrate vs simulation-calibrated rrate, and the",
    "segmentation agreement at each. Claim to support: they land in the same place,",
    "so simulation substitutes for the scarce chip truth."
  )
}
```

## 3. Throughput at population scale

nNIL is the geometric-duration categorical caller; the C++/Rcpp core lets it run
across the full 888-line panel cheaply (cf. `rtiger-optimization-benchmark.qmd`).

```{r scale-note}
cat(
  "[TO MEASURE] Wall time / memory for the full 888-line nNIL fit through the",
  "nilHMM C++ core."
)
```
