---
title: "Reproducing Holland's nNIL HMM grid search with `caller_grid`"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

::: {.callout-note}
## Scope: an independent reimplementation of Holland's grid, not a re-analysis of his CSV
The companion foil (`nnil-calibration-foil`) decomposes Holland's *published* grid
(File_S04) to show the chip-supervised objective is driven by `nir`. This notebook
instead **re-runs the entire 945-configuration grid** on the same real 24-line nNIL
GBS, scored against the chip calls, through nilHMM's unified `caller_grid()` kernel,
and asks two questions: does our caller reproduce Holland's result, and what is the
marker-mismatch floor actually made of. Compute lives in the tracked scripts
`scripts/nnil_foil/12_holland_grid.R` and `13_het_error_decomposition.R`; this
notebook reads their outputs. Markers are on maize **v5** (`S<chr>_<pos_v5>`).
:::

Zhong et al. (2025) calibrated the `nnil` caller with a **full-factorial grid
search** over five HMM parameters (File_S04), picking the configuration that
minimises the disagreement between the GBS calls and the chip calls on the 24
both-platform lines. We reproduced that search verbatim through `caller_grid()`,
the unified emission-by-transition sweep kernel: for each of the 315 emission
combinations we build the genotype emission once and batch the three-value `r`
transition grid through the C++ Viterbi, scoring every configuration's calls
against the chip. The question this answers is whether our reimplemented caller
lands on Holland's optimum and his variance structure, and, since it does, what
sets the residual ~2% mismatch that neither caller can beat.

```{r setup}
suppressMessages({
  library(data.table)
  library(ggplot2)
  library(patchwork)
  library(here)
  library(knitr)
})
FOIL <- here("data/nnil_foil")
theme_foil <- theme_classic(base_size = 10) +
  theme(
    text = element_text(family = "sans"),
    plot.tag = element_text(face = "bold"),
    legend.position = "top", legend.title = element_blank()
  )
col_ours <- "#0072B2" # caller_grid
col_holl <- "#D55E00" # Holland File_S04

rep <- fread(file.path(FOIL, "holland_grid_reproduction.csv")) # nir,germ,gert,p,r,mismatch,holland
sens <- fread(file.path(FOIL, "holland_grid_sensitivity.csv")) # param, ours, Holland
decomp <- fread(file.path(FOIL, "het_error_decomposition.csv")) # per-nir mismatch decomposition
conf <- fread(file.path(FOIL, "het_error_confusion.csv")) # V1=chip truth, V2=nnil call, N=genome fraction
f4 <- fread(here("agent/nNIL/File_S04.nNIL_gbs_vs_chip_data_HMMgridSearch.csv"))
params <- c("nir", "germ", "gert", "p", "r")
```

## 1. The grid: 945 = 7 × 3 × 3 × 5 × 3

Holland's search is the **Cartesian product** of the sample points of five
parameters. The grid size is simply the product of the per-parameter level counts,
not a pairwise count:

$$
\underbrace{7}_{\texttt{nir}}\times
\underbrace{3}_{\texttt{germ}}\times
\underbrace{3}_{\texttt{gert}}\times
\underbrace{5}_{\texttt{p}}\times
\underbrace{3}_{\texttt{r}}
= 945 .
$$

```{r grid-levels}
lv <- function(col) sort(unique(f4[[col]]))
roles <- c(
  nir = "emission non-informative rate (a donor-informative site emits no donor signal)",
  germ = "homozygote genotyping-error rate",
  gert = "heterozygote genotyping-error rate",
  p = "allele-bias term (splits the homozygote error between the two alleles)",
  r = "intermarker recombination fraction (transition / segment-length prior)"
)
levels_tab <- rbindlist(lapply(params, function(pp) {
  v <- lv(pp)
  data.table(
    parameter = pp, role = roles[[pp]], `n points` = length(v),
    `sample points` = paste(signif(v, 3), collapse = ", ")
  )
}))
levels_tab[, check := `n points`]
kable(levels_tab[, .(parameter, role, `n points`, `sample points`)],
  caption = sprintf(
    "The five swept parameters and their sample points. Product = %s = %d configurations.",
    paste(levels_tab$`n points`, collapse = " x "), prod(levels_tab$`n points`)
  )
)
```

`p` and `r` carry the most sample points (5 and 3), but as the next section shows
that spread is nearly wasted: almost all of the objective's movement is along `nir`.

## 2. `caller_grid` reproduces Holland's grid

Running the same 945 configurations through `caller_grid()` and scoring against the
chip reproduces File_S04 essentially exactly.

```{r reproduction-agreement}
pear <- cor(rep$mismatch, rep$holland)
spear <- cor(rep$mismatch, rep$holland, method = "spearman")
our_opt <- rep[which.min(mismatch)]
his_opt <- f4[which.min(mismatchMean)]
opt_tab <- rbind(
  data.table(source = "caller_grid (ours)", our_opt[, .(nir, germ, gert, p, r, mismatch = round(mismatch, 4))]),
  data.table(source = "Holland File_S04", his_opt[, .(nir, germ, gert, p, r, mismatch = round(mismatchMean, 4))])
)
kable(opt_tab, caption = sprintf(
  "Grid optima agree on the configuration. Per-configuration correlation across all 945: Pearson %.3f, Spearman %.3f.",
  pear, spear
))
```

The two grids pick the **same optimal configuration** (`nir` = 0.9, `germ` = 0.01,
`gert` = 1e-4, `p` = 0.9, `r` = map value), and every configuration's mismatch is
in lockstep (Pearson `r` = `{r} round(pear, 3)`). The variance decomposition of the
objective matches too: `nir` carries essentially all of it under both callers.

```{r variance-table}
setnames(sens, old = names(sens), new = sub("ours \\(caller_grid\\)", "ours", names(sens)))
st <- copy(sens)
for (cc in setdiff(names(st), "param")) st[[cc]] <- sprintf("%.1f%%", 100 * st[[cc]])
kable(st, caption = "Variance share of the GBS-vs-chip mismatch attributed to each parameter (main-effects ANOVA), our caller_grid vs Holland's File_S04.")
```

```{r fig-marginals, fig.width = 8, fig.height = 4.2, fig.cap = "Marginal mean GBS-vs-chip mismatch versus each parameter (log x). Top row, our caller_grid; bottom row, Holland's File_S04. Each ROW has its own y-scale, because the absolute floors differ (~1.8% vs ~0.07%); putting them on one axis would flatten Holland's curve and hide the shape. On their own scales the two rows have the SAME shape and the SAME minimum location: a steep fall along nir to a minimum near nir = 0.9 (open circle marks each panel's minimum), and near-flat responses along germ, gert, p, and r. Strip labels (top) give the parameter's variance share in our grid."}
marg <- rbindlist(lapply(params, function(pp) {
  o <- rep[, .(mm = mean(mismatch)), by = c(pp)][, .(level = get(pp), mm, param = pp, src = "caller_grid (ours)")]
  h <- rep[, .(mm = mean(holland)), by = c(pp)][, .(level = get(pp), mm, param = pp, src = "Holland File_S04")]
  rbind(o, h)
}))
vshare <- setNames(sens[["ours"]], sens$param)
marg[, param := factor(param, levels = params, labels = sprintf("%s (%.0f%%)", params, 100 * vshare[params]))]
marg[, src := factor(src, levels = c("caller_grid (ours)", "Holland File_S04"))]
mins <- marg[, .SD[which.min(mm)], by = .(src, param)] # minimum per panel
ggplot(marg, aes(level, mm, colour = src)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1.1) +
  geom_point(data = mins, shape = 21, size = 3, stroke = 0.9, fill = NA) +
  facet_grid(src ~ param, scales = "free") + # rows: source (own y); cols: parameter (own x)
  scale_x_log10() +
  scale_colour_manual(values = c("caller_grid (ours)" = col_ours, "Holland File_S04" = col_holl)) +
  labs(x = "parameter value (log scale)", y = "mean GBS-vs-chip mismatch") +
  theme_foil +
  theme(
    legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(angle = 0)
  )
```

The rows are on different scales by design: the **absolute** mismatch level differs,
our optimum at
`{r} sprintf("%.4f", our_opt$mismatch)` (~1.8%), Holland's File_S04 at
`{r} sprintf("%.4f", his_opt$mismatchMean)` (~0.07%). Pearson = 1 means this is a
pure vertical offset, so it changes no conclusion drawn from the grid, but it does
raise the question the last section answers: what is that ~2% floor.

## 3. The ~2% marker-mismatch floor is donor-recall, not HET-calling error

A natural guess is that the floor is a heterozygote-calling-error minimum
(spurious/dropped hets that no `nir` removes). It is not. Decomposing the
chip-truth by nnil-call confusion at the optimum `nir` shows the mismatch is
dominated by **REF↔ALT (homozygote↔homozygote)** disagreement, with only about a
quarter touching a HET call.

```{r floor-decomp}
d95 <- decomp[nir == 0.95]
floor_tab <- data.table(
  component = c("total mismatch", "REF<->ALT (donor recall)", "HET-involving"),
  `genome fraction` = round(c(d95$mismatch, d95$hom_hom, d95$het_involving), 4),
  `% of mismatch` = sprintf("%.0f%%", 100 * c(1, d95$hom_hom / d95$mismatch, d95$het_involving / d95$mismatch))
)
kable(floor_tab, caption = "Decomposition of the marker-mismatch floor at the optimum (nir = 0.95).")
```

```{r confusion-table}
cw <- dcast(conf, V1 ~ V2, value.var = "N")
setnames(cw, "V1", "chip")
cw <- cw[match(c("REF", "HET", "ALT"), chip), .(chip, REF, HET, ALT)]
kable(cw, digits = 5, caption = "Chip-truth (rows) by nnil-call (columns) confusion at nir = 0.95, as a genome fraction. Rows are the chip truth state, columns the nnil call. The largest off-diagonal cell is chip-ALT called REF (donor under-recall).")
```

The single largest error cell is **chip-ALT called REF**
(`{r} sprintf("%.4f", conf[V1 == "ALT" & V2 == "REF"]$N)`): the caller misses
roughly 0.9% of the genome the chip calls donor, calling it recurrent. That is a
donor-recall / boundary-placement gap (the caller slightly under-fragments), not a
het artifact. Removing every HET-involving disagreement leaves ~1.3% mismatch,
nowhere near Holland's 0.07%. So:

::: {.callout-important}
## The het-error-floor hypothesis is refuted
At the optimum the ~1.8% GBS-vs-chip floor is **~74% genuine REF↔ALT donor-recall
disagreement** (dominated by chip-donor blocks the caller calls recurrent) and only
**~26% HET-involving**, and this split is stable across `nir` (0.594, 0.9, 0.95).
The offset from Holland's absolute 0.07% is therefore a donor-recall / scoring
difference between our `nnil` and his `hmmlearn`, not a het-calling artifact. It
does not affect the simulation-calibration comparison (`fig:simcal`), which rests
on the `nir` argmin agreeing, not on the floor's composition.
:::

## Methods

`caller_grid()` (nilHMM, the unified `viterbi_sweep_cpp` kernel) runs the emission
grid at the R level (one genotype-emission build per combination) and batches the
transition (`r`) grid through the C++ Viterbi, reproducing a cold `call_ancestry()`
at every configuration. The objective is Holland's File_S16 criterion: total
GBS-vs-chip marker mismatch, `1 - marker accuracy` over all states including the
>90% REF background (`marker_dice`, `R/metrics.R`). Grid run:
`scripts/nnil_foil/12_holland_grid.R`; floor decomposition:
`scripts/nnil_foil/13_het_error_decomposition.R`. Both read the same real 24-line
GBS and chip calls as `08_nir_sweep.R`.
