---
title: "Calibrating the nnil caller against a simulation"
subtitle: "The non-informative rate calibrated on a count-from-parents simulation reproduces calibration against the scarce SNP-chip truth"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
library(here)
FIG <- here::here("results/sim/nested_nnil/fig_nnil_sim_calibration.png")
```

## Thesis

Holland's nested-NIL (nnil) introgression caller has one parameter that matters,
the **non-informative rate `nir`**, the probability that a donor's SNP allele is
identical in state to B73 so the marker carries no ancestry signal. Zhong et al.
(2025) set `nir` by grid search against a **scarce, expensive truth**: 24 of the 888
nNILs re-genotyped on a SNP chip. Here we show that calibrating `nir` against a
**simulation**, where the truth is known for free at every marker, lands at the
same operating point and produces the same calls on the real data. Simulation can
therefore stand in for the scarce chip truth.

The simulation is built from the **real donor genotypes** (not a scalar knob), so a
donor's non-informativeness emerges mechanically and per-donor; this is the fidelity
that makes the substitution credible.

## Methods

*(Paper-ready. Cross-references to files are to the Zhong et al. 2025 supplement
and to this repository's tracked scripts.)*

**Count-from-parents simulation.** We simulated the genotyped nNIL population from
the genotypes of its actual donors. The 16 NAM founders that produced the 888
GBS-genotyped nNILs were taken from the SNP-chip data of Zhong et al. (2025, File
S2) and placed on the GBS marker frame, the 11,310 markers shared between the
20,221 QC-passing chip SNPs and the 64,024-marker GBS set, lifted AGPv3→v4→v5. For
each donor we simulated a **BC5S4** family (five backcrosses to the recurrent parent
B73, then four generations of selfing, the generation Zhong et al. report for the
sequenced material) with `simcross` (Stahl interference m = 10), yielding the latent
ancestry mosaic (0/1/2 donor dose) at every marker. The **observed** allelic dose
was the ancestry dose multiplied by the founder's allele at that marker, so a marker
is non-informative wherever the donor carries the B73/REF allele. Donor
non-informativeness (`nir`) thus emerges **per-donor and spatially** from the real
founder genotypes rather than from a scalar parameter; the realized per-donor
non-informative rate (`f0` ≈ 0.55) matches the value measured directly from the
founder genotypes (0.535). Because Zhong et al. report no per-site read depth and
the raw reads are not deposited, we inferred an effective coverage from the observed
GBS missing-data rate (`mr` = 0.093), giving λ ≈ 2.7 reads per site (Poisson, with a
1% structural-dropout floor); allele read counts were drawn under this regime with a
1% per-read error. We simulated 40 lines per donor (640 nNILs total).

**Genotype calling.** Simulated read counts were converted to hard genotypes by
maximum likelihood (a flat prior; a single ALT read yields a homozygous-ALT call),
matching the likelihood-based calling of the TASSEL-GBS pipeline used by Zhong et al.

**Ancestry caller.** Introgressions were called with `nnil` (nilHMM), a
reimplementation of Holland's nested-NIL hidden Markov model: a categorical
genotype-confusion emission parameterised by the non-informative rate `nir` (plus
genotyping-error rates) and a geometric duration. Each line is decoded independently
under a single, donor-agnostic `nir`, as in Zhong et al.; the adjacent-marker
recombination fraction was fixed at the genetic-map value ($r = 2L/100n$).

**Calibration.** `nir` was chosen to minimise the **total marker mismatch** between
the caller's states and a reference, the criterion of Zhong et al. (File S16),
swept over `nir` ∈ [0.01, 0.95]. Two references define two calibrations:
*simulation* (mismatch against the simulated latent ancestry, over all 640 nNILs)
and *chip* (mismatch against the SNP-chip introgression calls, over the 24
both-platform nNILs, reproducing Zhong et al.). For interpretation we also measured
each donor's non-informative rate `f0` directly, as the fraction of markers at which
the donor genotype equals the B73/REF allele: from the simulated data (realized) and
from the real founder genotypes.

**Evaluation.** We compared donor introgression-size distributions (empirical CDFs)
of the calibrated caller against the truth on each dataset, and used a QQ plot to
compare the per-line introgression sizes obtained on the **same real 24-line GBS**
under the simulation-chosen versus chip-chosen `nir`.

**A note on scope.** Count-emission callers such as rtiger are excluded from this
comparison: the released nNIL GBS provides genotype calls but no read counts, and
manufacturing counts from the calls to feed a count caller would be circular. rtiger
can be validated on the simulation (where counts exist) but has no real-data leg here.

**Data and code (this repository).** Founder genotypes on the v5 GBS frame:
`scripts/nnil_foil/10_founder_genotypes.py`. Effective-coverage / missing-data
regime fit: `scripts/nnil_foil/11_fit_gbs_regime.R`. Count-from-parents simulator:
`scripts/simulate_nested_nils.R`. Calibration, evaluation, and this figure (with a
cached compute step): `scripts/fig_nnil_sim_calibration.R`. Decomposition of the
chip-calibration mismatch floor by GBS genotype: `scripts/nnil_chip_floor_decomposition.R`. Real GBS calls:
`data/nnil_equiv/geno_recoded.csv`; chip truth: `data/nnil_foil/chip_truth_projected.csv`.

## Figure

```{r}
#| label: fig-nnil-calibration
#| echo: false
#| out-width: "100%"
#| fig-cap: "**Ancestry HMM parameter calibrations from `simcross` simulation or chip data are equivalent in the nNIL population.** **(A)** Marker-mismatch calibration curves: total marker mismatch of the `nnil` caller versus `nir`, scored against the simulated ancestry (blue; 640 BC5S4 nNILs from 16 real NAM founders) and against the SNP-chip introgression calls on the 24 both-platform nNILs (orange); open circles mark the chosen operating points, coloured as the resulting caller in panels C and D: blue sim-calibrated at `nir`\\* = 0.70 and orange chip-calibrated at `nir`\\* = 0.80, both on the shallow high-`nir` plateau near Zhong et al.'s 0.90. Dashed verticals mark the donors' measured non-informative rate `f0` (blue, realized in the simulation; black, from the chip founder genotypes), ≈ 0.54, well below the mismatch-optimal `nir`, because the >90% REF background makes marker mismatch a weak criterion that tolerates over-smoothing. **(B)** Donor introgression-size ECDF on the simulated GBS: sim-calibrated `nnil` calls versus the simulated ancestry truth (dotted). **(C)** On the 24 real nNILs, introgression-size ECDFs of the chip- and sim-calibrated `nnil` calls versus the chip calls (Zhong et al. 2025, dotted). Significance: p-value from a two-sample Kolmogorov-Smirnov test. **(D)** QQ plot of per-line introgression sizes from the sim-calibrated versus chip-calibrated `nnil`, both applied to the same real GBS; points on the solid grey 1:1 line indicate the two calibrations yield equivalent calls. rtiger is omitted because it requires read counts, which the released GBS lacks."
if (file.exists(FIG)) {
  knitr::include_graphics(FIG)
} else {
  cat(
    "Figure not yet generated. Run: Rscript scripts/fig_nnil_sim_calibration.R",
    "(NNIL_SIMCAL_RECOMPUTE=1 to rebuild the cached compute)."
  )
}
```

## Results

- **The calibrations agree, and near Holland's value.** Minimising marker mismatch
  gives `nir`\* = 0.70 on the simulation and 0.80 on the chip; the chip curve is
  nearly flat from 0.5 to 0.9, so both, and Zhong et al.'s 0.90, sit on one shallow
  minimum. Calibrating on the simulation recovers the same operating point as
  calibrating on the scarce chip truth (panel A).
- **The same calls on real data.** Applying the sim-chosen (`nir` = 0.70) and
  chip-chosen (`nir` = 0.80) operating points to the identical 24-line real GBS gives
  essentially indistinguishable introgression-size distributions: the QQ falls on
  the 1:1 line (panel D), and the two calibrated ECDFs overlie each other in panel C.
  **Simulation substitutes for the scarce chip truth.**
- **`nir` is inflated above biology.** The donors' actual non-informative rate is
  ≈ 0.54 (measured identically in the simulation and in the real founders; panel A
  dotted lines), well below the mismatch-optimal 0.7 to 0.9. Marker mismatch is
  dominated by the >90% REF background, so it is minimised by an over-smoothing `nir`
  far above the true donor non-informativeness, a caveat for interpreting `nir` as a
  biological quantity, and a motivation for donor-recovery-based criteria.
- **Simulation fidelity.** The sim-calibrated caller recovers the simulated ancestry
  fragment distribution (panel B), confirming the generative model and the calibration
  are self-consistent.
- **The real-data floor is missed donor, not heterozygous calls.** The chip
  calibration bottoms out at 1.9% marker mismatch against 0.1% for the simulation
  (panel A). Cross-classifying every disagreeing marker on the 24 both-platform NILs
  by its GBS genotype (`scripts/nnil_chip_floor_decomposition.R`) shows this floor is
  not the heterozygosity one might expect from low-coverage ML (TASSEL-GBS) calling.
  Heterozygous GBS calls are indeed unreliable, 49% of them disagree with the chip
  against 1.7% for REF-homozygous calls, but they are far too rare to matter:
  heterozygotes are only 0.2% of scored genotypes and 5.5% of all mismatches. The
  floor is instead dominated by REF-homozygous and missing calls (78% and 9% of
  mismatches). The single largest bucket, 52% of all disagreements, is the caller
  reading B73 where the chip reads donor: donor introgressions missed because the
  informative donor allele dropped out under the 8.8% GBS missing rate and low
  coverage. The 1.9% gap is therefore real-data donor-allele dropout that the
  count-from-parents simulation does not fully reproduce.

## Caveats / to do

- The `nir` grid is coarse (step 0.1); the plateau (0.5 to 0.9) is real but the exact
  argmin (0.70 vs 0.80) is within noise. A finer sweep would state the plateau
  precisely.
- Marker mismatch is Zhong et al.'s criterion and is used here so the sim↔chip
  comparison is on their terms; a donor-recovery metric (e.g. Dice) would sharpen the
  optimum but is a separate analysis.
