---
title: "Pedigree caller on a ZEAL-structured cohort"
subtitle: "Does family coupling beat per-individual calling at ZEAL's scale and skim depth? — 82 donor founders, ~17 lines/founder, n ≈ 1394"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
params:
  design: "BC2S3"
  founders: 82          # ZEAL donor accessions (data/zeal/bzea_50K_cohort_ref_metadata.csv)
  lines_per_founder: 17 # ZEAL mean lines/founder (17.1); fixed-sibs approximation of a skewed real dist
  n_markers: 600
  chr: 1
  seed: 11
---

::: {.callout-note title="Dependency"}
This notebook uses `nilHMM`'s `call_ancestry(caller = "pedigree")`. Until nilhmm
**PR #10** (`feat/pedigree-counts-emission`) is merged to `main`, reproduce by
installing `nilHMM` from that branch (`R CMD INSTALL` the branch working tree); the
setup guard below fails loudly against an older build. Method: paper supplement
Text S9 and `nilhmm/design/PEDIGREE_HMM.md`.
:::

The count/geometric callers decode each individual's chromosome independently; the
`pedigree` caller instead couples a selfing family's relatives by loopy belief
propagation over the (pedigree × genome) grid (supplement Text S9). This notebook
asks the honest question on a **ZEAL-structured** simulated cohort: at ZEAL's scale
(82 donor founders, ~17 lines per founder, ~1400 lines) and skim depth, does the
coupling beat per-individual calling? The answer here is a **characterization, not
a win** — see the Take.

## Setup

```{r setup}
suppressMessages({
  library(nilHMM) # simulate_family, simulate_counts, call_states, to_segments
  library(data.table)
  library(ggplot2)
})
stopifnot("pedigree" %in% eval(formals(nilHMM::call_states)$caller)) # guard: caller must be installed
set.seed(params$seed)

# Per-marker dosage accuracy of a call_states() table vs the simcross truth mosaic.
dosage_acc <- function(calls, truth) {
  kt <- paste(truth$name, truth$chr, truth$pos)
  ts <- truth$state[match(paste(calls$name, calls$chr, calls$pos), kt)]
  mean(calls$state == ts, na.rm = TRUE)
}
```

## 1. A ZEAL-matched cohort

The ZEAL SNP50K cohort is **1439 lines** descended from **82 teosinte donor
founders** (accessions), a **mean of 17.1 lines per founder** — a right-skewed
distribution (median 11, range 3–100) spanning five teosinte taxa
(*mexicana* 35 founders, *parviglumis* 36, *diploperennis* 4, *luxurians* 5,
*huehuetenangensis* 2; from `data/zeal/bzea_50K_cohort_ref_metadata.csv`, pedigree
strings like `Zx.0580_P3_P1_P4.1.1.1.B`). `simulate_family()` draws fixed-size
families, so we approximate the population by its **mean**: `r params$founders`
founders × `r params$lines_per_founder` lines. (The fixed-sibs simplification is a
known departure from the skewed real distribution; it matches total n, founder
count, and mean family size — the structure that drives sibling coupling.)

```{r cohort}
sim <- simulate_family(params$design,
  families = params$founders, sibs = params$lines_per_founder,
  chr = params$chr, n_markers = params$n_markers, seed = params$seed
)
truth <- sim$truth

data.table(
  quantity = c("total lines", "donor founders", "mean lines/founder"),
  `ZEAL (real)` = c("1439", "82", "17.1"),
  `simulated` = c(
    length(unique(truth$name)),
    length(unique(sim$pedigree$family)),
    round(length(unique(truth$name)) / length(unique(sim$pedigree$family)), 1)
  )
) |> knitr::kable()
```

## 2. Accuracy versus sequencing depth

Degrade the same truth mosaic across a depth ladder bracketing ZEAL's skim regime
(~0.4×) and call three ways on the identical reads: two single-chain callers
(`nnil`, `rtiger`) and the family-coupled `pedigree` caller.

```{r depth-sweep}
depths <- c(0.1, 0.2, 0.4, 0.8, 1.6, 3.2)
depth_res <- rbindlist(lapply(depths, function(d) {
  o <- simulate_counts(truth, depth = d, error = 0.01, seed = params$seed)
  o$cm <- truth$cm[match(paste(o$name, o$chr, o$pos), paste(truth$name, truth$chr, truth$pos))]
  data.table(
    depth    = d,
    zero_cov = mean((o$n_ref + o$n_alt) == 0),
    nnil     = dosage_acc(call_states(o, caller = "nnil", design = params$design, rrate = 1e-3), truth),
    rtiger   = dosage_acc(call_states(o, caller = "rtiger", design = params$design, rigidity = 3L), truth),
    pedigree = dosage_acc(call_states(o, caller = "pedigree", pedigree = sim$pedigree, design = params$design), truth)
  )
}))
knitr::kable(depth_res, digits = 3)

ggplot(
  melt(depth_res[, .(depth, nnil, rtiger, pedigree)],
    id.vars = "depth",
    variable.name = "caller", value.name = "accuracy"
  ),
  aes(depth, accuracy, colour = caller)
) +
  geom_vline(xintercept = 0.4, linetype = "dotted") +
  annotate("text", x = 0.4, y = -Inf, label = " ZEAL ~0.4×", hjust = 0, vjust = -0.5, size = 3) +
  geom_line() +
  geom_point() +
  scale_x_log10() +
  labs(
    x = "depth (×, log)", y = "per-marker dosage accuracy",
    title = "At ZEAL scale/depth the pedigree caller reaches parity, not a gain",
    subtitle = "rtiger's within-individual rigidity is the strong baseline; coupling does not overtake it"
  ) +
  theme_minimal()
```

## 3. What the coupling *does* buy: full-grid missing-data fill

At skim depth a large fraction of markers carry no reads. The count emission emits
flat there (`n = 0`) and the pedigree fills the state from relatives (Text S9.3):
the caller leaves **no marker uncalled**. We score accuracy specifically on the
zero-coverage markers at ZEAL's ~0.4×.

```{r zero-cov}
o <- simulate_counts(truth, depth = 0.4, error = 0.01, seed = params$seed)
o$cm <- truth$cm[match(paste(o$name, o$chr, o$pos), paste(truth$name, truth$chr, truth$pos))]
o$depth <- o$n_ref + o$n_alt
cp <- call_states(o, caller = "pedigree", pedigree = sim$pedigree, design = params$design)
zc <- o[o$depth == 0, c("name", "chr", "pos")]
kt <- paste(truth$name, truth$chr, truth$pos)
ts_z <- truth$state[match(paste(zc$name, zc$chr, zc$pos), kt)]
pz <- cp$state[match(paste(zc$name, zc$chr, zc$pos), paste(cp$name, cp$chr, cp$pos))]
cat(sprintf("zero-coverage markers at 0.4x: %.0f%% of the grid\n", 100 * nrow(zc) / nrow(o)))
cat(sprintf("pedigree calls left uncalled: %d\n", sum(is.na(pz))))
cat(sprintf("pedigree accuracy on the zero-coverage markers: %.3f\n", mean(pz == ts_z, na.rm = TRUE)))
```

## 4. Why parity, not a gain

The gap to `rtiger` is expected from the model, not a bug. The current kernel
couples relatives with a **per-marker** selfing factor $T^{\mathrm{sib}}$, which
drops *transmission linkage*: a parent transmits a donor **segment** as a coherent
block, but the per-marker factor only pools siblings marker-by-marker (weak
marginal pooling), not over the shared blocks they actually inherited
(`nilhmm/design/PEDIGREE_HMM.md` §2). Meanwhile `rtiger` already captures
block coherence *within* each individual through its rigidity prior — the strong
baseline. Block-coherent *cross-sibling* pooling — the mechanism that would let the
coupling overtake `rtiger` — needs the **phased copy-switch transmission kernel**
(§18), deferred to a follow-up branch.

## Take

On a ZEAL-structured cohort (82 founders, ~17 lines/founder, n ≈ 1394) at skim
depth, the `pedigree` caller:

- **reaches parity with, but does not beat, the single-chain callers** — at ~0.4× it
  matches `nnil` and trails `rtiger` by a small margin, and the three converge by
  ~1×. This is consistent with the design's own prediction (§2): the per-marker
  transmission factor cannot pool siblings over inherited blocks.
- **calls 100% of markers, including the fully-uncovered majority**, filling
  zero-coverage sites from relatives — a real capability the single-chain callers
  achieve only through their own spatial smoothing.

So the caller is validated as **correct and principled infrastructure** (depth-aware
family coupling + the `refine_ancestry()` workflow + the foundation for the phased
kernel), **not** as an accuracy improvement over `rtiger` on ZEAL-like data. The
block-coherent gain is the goal of the phased copy-switch kernel (§18), a separate
branch.
