---
title: "RTIGER reimplementation benchmark: nilHMM C++/Rcpp vs Julia"
subtitle: "Supplement (optimization): correctness-first equivalence, then throughput/memory at 10^4-10^5 markers"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

::: {.callout-tip title="Measured on real Arabidopsis data"}
Numbers, tables and figures below are the live benchmark
(`scripts/bench_rtiger_cpp_vs_julia.R`), not placeholders. The C++/Rcpp core is
run against the upstream-original Julia RTIGER on the shared Col x Ler panel; the
original's 110k convergence fit (~3.4 h) is read from the preserved fork runs
rather than regenerated.
:::

**Thesis.** nilHMM reimplements the RTIGER rigid-HMM core in **C++/Rcpp**
(`nilhmm/src/rtiger.cpp`, `viterbi_batch.cpp`, `emission_count.cpp`, ...), removing
the Julia backend that upstream RTIGER \[Campos-Martin 2023\] requires through
`JuliaCall`. The reimplementation is a **prerequisite**, not a contribution: the
paper's simulation-scale calibration and benchmarking (a caller rerun across many
parameter settings on 10^4-10^5-marker panels) are only tractable on a laptop once
the per-fit cost and the Julia runtime dependency are gone. This note establishes
two things, in order of importance:

1. **Correctness first.** The C++/Rcpp core returns the *same answer* as the Julia
   RTIGER, Viterbi-identical ancestry paths and fitted parameters agreeing to
   float-summation tolerance, from identical initialization on real data. **Result:
   0 mismatches at every panel size, same iteration count, params to <=8e-6.**
2. **Then speed/memory.** Only after equivalence is established do we report the
   throughput and memory gains. **Result: near-linear (markers^0.94) vs the
   original's quadratic (markers^2.10); ~10,000x at 110k markers.**

**Reference point.** The prior *optimized-Julia vs original-Julia* fork
(`faustovrz/RTIGER@optimize-julia-core`, `docs/optimization.md`) established the
method used here, the shared-panel marker sweep, the Viterbi-identical
equivalence oracle, and the peak-RSS sweep. That fork made the original's Julia
core ~610x faster; the C++/Rcpp port is comparable to or faster than that
optimized Julia while removing the runtime entirely, so measured against the
*upstream original* it lands at ~10,000x.

::: {.callout-note title="Compute is separated from this note"}
The benchmark is run by `scripts/bench_rtiger_cpp_vs_julia.R`, which builds the
shared *Arabidopsis* Col x Ler panel, thins it by odd index to five sizes,
fits the nilHMM C++ core at each size (fixed iterations for throughput; to
convergence + a per-iteration delta trace for equivalence), compares against the
preserved original-Julia oracle, and writes summary CSVs to `results/bench/` plus
the two figures to `nilhmm-paper/figures/`. This note only loads and displays
them. Regenerate with `Rscript scripts/bench_rtiger_cpp_vs_julia.R --generate`
(persist a run via `... 2>&1 | tee agent/bench_rtiger_<ts>_<pid>.log`; sources
`scripts/logging.R`, supports `--smoke`). The whole C++ sweep runs in well under a
minute; the original-Julia baseline is NOT re-run (its 110k convergence fit is
~3.4 h upstream) but read from the preserved fork dumps in `data/bench_ref/`.
Rendering this note (`freeze: auto`) only reads the CSVs, never the benchmark.

Inputs staged from the fork (see `DATA.md`): the 109,703-locus shared-3 panel
(`data/rtiger_shared3_input/`) and the original-Julia reference runs
(`data/bench_ref/`), originally from BioProject PRJNA720439 / SRP313911.

Machine/toolchain for the timings is recorded per run in
`results/bench/rtiger_env.txt` (Apple M4, 10 cores, 24 GB, macOS 26.5; nilHMM
C++/Rcpp built native arm64 with Apple clang 17 at `-O2`; R 4.5.2; single-threaded;
original-Julia baseline on Julia 1.12.6, same hardware). The core's
forward/backward, Viterbi, and emission M-step are hand-written scalar C++
(`LinkingTo`: only Rcpp and RcppParallel; no Armadillo, Eigen, or external
BLAS/LAPACK), so the system linear-algebra library cannot affect these timings.
:::

```{r setup}
suppressMessages({
  library(data.table)
})
BENCH <- here::here("results/bench")
FIG <- here::here("nilhmm-paper/figures")
f <- function(x) file.path(BENCH, x)
fg <- function(x) file.path(FIG, x)
have_equiv <- file.exists(f("rtiger_equivalence.csv"))
have_scale <- file.exists(f("rtiger_marker_scaling.csv"))
```

```{r env-table}
if (file.exists(f("rtiger_env.txt"))) {
  env <- fread(f("rtiger_env.txt"),
    header = FALSE, sep = "\t",
    col.names = c("field", "value")
  )
  knitr::kable(env, caption = "Machine and toolchain for the benchmark timings (results/bench/rtiger_env.txt, written per run).")
}
```

## 1. Why reimplement in C++/Rcpp

Two properties of the upstream implementation motivate the port:

- **Runtime dependency.** Upstream RTIGER runs its time-critical core in Julia via
  `JuliaCall`; a matching Julia + package environment must be installed and loaded.
  For an R package intended to run many callers on many parameter settings, that
  cross-runtime dependency is both a friction and a reproducibility hazard.
- **Per-fit cost at density.** The paper works at 10^4-10^5 markers per line and
  reruns callers across calibration sweeps. The fit must be cheap enough to loop.

The C++/Rcpp core keeps the rigid-HMM arithmetic (rHMM EM + rigid Viterbi + segment
boundary postprocessing) but runs in-process with no external runtime.

## 2. Equivalence (correctness first)

The claim the port must earn: **same answer as Julia RTIGER**. Following the fork's
oracle, both cores are run to full convergence from identical deterministic
initialization on the shared panel at all five sizes, and outputs compared
position-by-position.

```{r equiv-table}
eq <- fread(f("rtiger_equivalence.csv"))
knitr::kable(
  eq,
  col.names = c(
    "level", "markers/sample", "iters (C++)", "iters (Julia)",
    "Viterbi total", "mismatches", "param max-diff"
  ),
  caption = "nilHMM C++/Rcpp vs upstream-original Julia RTIGER, run to convergence from the identical deterministic init: same EM iteration count, Viterbi-identical ancestry paths (0 mismatches), parameters agreeing to float-summation tolerance."
)
```

The two cores descend the identical path to convergence at the full panel
(right); the left panel is the speed contrast:

![Shared panel, 109,703 markers/sample, both cores run to convergence (19 iterations each). Left: per-iteration wall time (original Julia ~648 s/iter vs nilHMM C++/Rcpp ~0.1 s/iter). Right: each iteration's EM parameter-convergence delta (max |Delta(alpha,beta)|, RTIGER's convergence criterion, not a log-likelihood change), C++ against original Julia on a log-log 1:1 line, early (yellow) to late (purple).](../nilhmm-paper/figures/rtiger_equivalence_110K_cpp.png)

## 3. Throughput, scaling with markers

Shared-panel head-to-head, per-iteration wall time vs markers/sample, on the five
thinned sizes (6,857 -> 13,713 -> 27,426 -> 54,852 -> 109,703). On a log-log
plot the slope is the empirical complexity exponent: the original is
markers^2.10 (quadratic), the C++/Rcpp core markers^0.94 (linear).

```{r scaling-table}
sc <- fread(f("rtiger_marker_scaling.csv"))
wide <- dcast(sc, mps ~ core, value.var = "per_iter")
setnames(wide, c("mps", "cpp", "orig"),
  c("markers/sample", "C++/Rcpp s/iter", "orig Julia s/iter"),
  skip_absent = TRUE
)
wide[, `speed-up` := round(`orig Julia s/iter` / `C++/Rcpp s/iter`)]
knitr::kable(wide[order(-`markers/sample`)],
  caption = "Per-iteration wall time (fixed-iteration fit, eps=0.01, rigidity 2), and the per-iteration speed-up."
)
```

![RTIGER scaling on the shared panel, original array-retaining Julia vs nilHMM C++/Rcpp (both measured on the same machine); points are measurements, dashed lines the power-law fits. Left: per-iteration wall time (original ~markers^2.10 quadratic, C++/Rcpp ~markers^0.94 linear). Right: peak resident memory; the C++ core streams sufficient statistics and stays ~5x lower.](../nilhmm-paper/figures/rtiger_marker_scaling.png)

## 4. Memory

The C++ E-step folds each chain into fixed-size sufficient statistics (the s x s
transition counts, the start/emission accumulators, and a map over the distinct
(k, n) read-count pairs) and reduces them in a fixed chain order; it never retains
the per-sample zeta/gamma/alpha/beta arrays. EM scratch is therefore
O(threads . T_max), not O(N . T): it does **not** grow with the sample count N. A
peak-RSS sweep (cycling 2 to 60 samples at 50,000 markers each,
`/usr/bin/time -l`) confirms it: peak resident memory grows as only
~162 MiB + 2.1 MiB/sample, where the intercept is the R/package runtime and the
per-sample slope is the retained input count vectors, not EM working memory.

```{r mem-table}
if (file.exists(f("rtiger_memory_sweep.csv"))) {
  mm <- fread(f("rtiger_memory_sweep.csv"))
  lf <- lm(peak_rss_mib ~ n_samples, mm)
  knitr::kable(mm,
    col.names = c("samples", "markers/sample", "peak RSS (MiB)"),
    caption = sprintf(
      "nilHMM C++/Rcpp peak resident memory vs sample count at fixed markers/sample. Fit: %.0f MiB + %.2f MiB/sample (R2=%.3f).",
      coef(lf)[1], coef(lf)[2], summary(lf)$r.squared
    )
  )
}
```

Regenerate with `Rscript scripts/bench_rtiger_cpp_vs_julia.R --memory`.

## 5. Reproducing

`scripts/bench_rtiger_cpp_vs_julia.R --generate` builds the shared panel from the
staged Col x Ler inputs, runs the C++ core at each size (fixed iters for §3, to
convergence + delta trace for §2), compares against the preserved original-Julia
oracle in `data/bench_ref/`, and writes the CSVs and figures shown above. Only the
built nilHMM C++ core is needed; the original Julia baseline is read from the fork
dumps, not re-run.
