---
title: "Simulation-calibrated ancestry calling: two-stage parameter search"
subtitle: "Materials & Methods §2.4–2.5 + Results §3.1–3.2 — calibrate on simcross, validate on truth"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**Thesis.** Ancestry-caller segmentation parameters (`rrate` for nNIL, `rigidity`
for RTIGER) can be **calibrated from cheap Broman `simcross` simulations** where
the donor mosaic is known truth, rather than tuned by eye on real data.

**Method — a two-stage search** (`R/calibrate.R`, driven by
`scripts/02_calibrate.R`):

1. **Log sweep** — score a coarse log-spaced grid against truth to *bracket* the
   optimum (`sweep_calibrate` → one shared `nilHMM::caller_sweep` fit + scoring).
2. **Golden-ratio refine** — golden-section search inside the bracket on the
   continuous `log10(parameter)` to *localize* it (`golden_refine`).

The objective is **donor-fragment Dice** (reciprocal-overlap of called vs true
introgression blocks) — it penalizes both missed donor and the spurious 1-marker
blocks that over-fragmentation produces, unlike marker recall which saturates.

::: {.callout-note title="Compute is separated from this note"}
`scripts/02_calibrate.R` runs the sim, calibration, and benchmark, writing summary
CSVs to `results/sim/` (gitignored). This note only loads and plots them.
Regenerate with `Rscript scripts/02_calibrate.R --generate`.
:::

```{r setup}
suppressMessages({
  library(data.table)
  library(ggplot2)
})
SIM <- here::here("results/sim")
f <- function(x) file.path(SIM, x)
# Pinned BC2S2 params snapshot: the generic calib_params.csv gets overwritten by
# BC1S4 runs (character `design` row breaks this BC2S2 note), so read a dated copy.
CALIB <- "calib_params_bc2s2_20260706024342.csv"
have <- file.exists(f(CALIB))
if (have) {
  P <- fread(f(CALIB))
  par <- setNames(P$value, P$key)
  nnil_sw <- fread(f("nnil_rrate_sweep.csv"))
  rtig_sw <- fread(f("rtiger_rigidity_sweep.csv"))
  nnil_rf <- fread(f("nnil_rrate_refine.csv"))
  rtig_rf <- fread(f("rtiger_rigidity_refine.csv"))
  bench <- fread(f("benchmark.csv"))
  frag <- fread(f("frag_sizes.csv"))
  geno <- fread(f("geno_fractions.csv"))
}
```

# 1. The simulation and its truth (§2.4)

**BC2S2**, n = 1500, `simcross` on the maize consensus map (~1783 cM, Stahl
interference m = 10 / p = 0), 50 000 markers, degraded to the skim regime
(~0.43× / ~72% missing). Truth = the donor mosaic. The simulated genome
composition should match the single-locus (Mendelian) expectation:

```{r truth, eval=(exists("have") && have)}
data.table(
  metric = c("REF", "HET", "ALT", "donor dosage"),
  simulated = round(par[c("REF_truth", "HET_truth", "ALT_truth", "dosage_truth")], 4),
  expected = round(par[c("REF_expect", "HET_expect", "ALT_expect", "dosage_expect")], 4)
)
```

# 2. Two-stage calibration

For each caller the coarse **log sweep** (line) brackets the optimum, and the
**golden-ratio refine** (points) homes in; the dashed line is the refined
parameter carried to the benchmark.

```{r calib-fig, eval=(exists("have") && have), fig.width=9, fig.height=4}
calib_panel <- function(sweep, refine, star, xlab, xtrans) {
  ggplot(sweep, aes(value)) +
    geom_line(aes(y = donor_frag_dice, colour = "donor-fragment Dice")) +
    geom_line(aes(y = donor_frag_FDR, colour = "donor-fragment FDR")) +
    geom_point(data = refine, aes(value, objective, colour = "golden probes"), size = 1.6) +
    geom_vline(xintercept = star, linetype = "dashed") +
    xtrans +
    labs(x = xlab, y = "score", colour = NULL) +
    theme(legend.position = "top")
}
p_nnil <- calib_panel(nnil_sw, nnil_rf, par["rrate_star"], "nNIL rrate", scale_x_log10()) +
  ggtitle(sprintf("nNIL — refined rrate = %.2g", par["rrate_star"]))
p_rtig <- calib_panel(
  rtig_sw, rtig_rf, par["rigidity_star"], "RTIGER rigidity",
  scale_x_continuous(trans = "log2")
) +
  ggtitle(sprintf("RTIGER — refined rigidity = %d", as.integer(par["rigidity_star"])))
patchwork_ok <- requireNamespace("patchwork", quietly = TRUE)
if (patchwork_ok) {
  print(patchwork::wrap_plots(p_nnil, p_rtig))
} else {
  print(p_nnil)
  print(p_rtig)
}
```

```{r calib-tables, eval=(exists("have") && have)}
# The log-sweep curves (Dice peaks then over-fragments as FDR/breakpoints climb).
show <- function(s, pcol) {
  s[, .(
    value = signif(value, 3), donor_frag_dice = round(donor_frag_dice, 3),
    donor_frag_FDR = round(donor_frag_FDR, 3), donor_marker_recall = round(donor_marker_recall, 3),
    n_breakpoints, truth_bp, ks = round(ks_fragsize, 3)
  )]
}
cat("nNIL rrate log sweep:\n")
print(show(nnil_sw))
cat("\nRTIGER rigidity log sweep (powers of 2, feasibility-capped):\n")
print(show(rtig_sw))
```

# 3. Benchmark at the calibrated parameters (§2.5, §3.2)

Each caller run at its refined optimum on a held-out set, scored vs truth:

```{r benchmark, eval=(exists("have") && have)}
bench[]
```

# 4. Fragment-size fit (supplementary)

Donor-block size distributions vs the BC2S2 truth (the calibrated callers should
track the truth's fragment sizes; KS in the benchmark table).

```{r fragsize, eval=(exists("have") && have), fig.width=7, fig.height=4}
ggplot(frag, aes(mb, colour = src)) +
  stat_ecdf() +
  scale_x_log10() +
  labs(x = "donor block size (Mb)", y = "ECDF", title = "Fragment-size fit vs truth")
```

# 5. Genotype-frequency vs the one-locus expectation (supplementary)

```{r genofreq, eval=(exists("have") && have), fig.width=7, fig.height=4}
ggplot(melt(geno[, .(src, HET, ALT)], id.vars = "src"), aes(src, value)) +
  geom_violin(scale = "width") +
  facet_wrap(~variable, scales = "free_y") +
  geom_hline(data = data.table(
    variable = c("HET", "ALT"), y = c(par["HET_expect"], par["ALT_expect"])
  ), aes(yintercept = y), linetype = "dashed") +
  labs(x = NULL, y = "genome fraction", title = "Genotype frequency vs one-locus expectation (dashed)") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

# Next steps

- Add the BrB regime and carry the per-source calibrated parameters to
  `source-method-comparison.qmd`.
- Sharpen the objective's soft plateau near the optimum with a KS or
  breakpoint-match tie-breaker.
