---
title: "Single-locus validation of the BC2S3 simulation"
subtitle: "Does the simcross BC2S3 simulator reproduce the breeding-scheme genotype frequencies? ŌĆö on the TeoNAM native v5 map"
author: "Fausto Rodr├Łguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
params:
  n: 1500
  m: 10
  p: 0
  seed: 1
---

Does the Broman/`simcross` BC2S3 simulation reproduce the **single-locus
breeding-scheme expectation** for the REF/Het/ALT genome fractions? This notebook
runs that check on the **TeoNAM native v5 cM map** (the same 118K-marker grid the
per-coverage caller calibration uses), validating the simulator this repo's
coverage-degradation sweeps rely on. It is a *circularity* check: before trusting a
caller's recovery of the truth, confirm the truth-generator itself is unbiased.

The sim stack is zealhmm's own (`R/simulate.R`: `.bcsft_pedigree`, `.simulate_dosage`);
the single-locus expectation and the design-correct goodness-of-fit test are in
`R/single_locus.R` (ported from the exploratory `zealtiger` repo). Why Hotelling and
not Pearson, why Mb on real marker coordinates, and the ergodicity distinction are
explained inline.

## Setup

```{r setup}
suppressMessages({
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(simcross)
  library(patchwork)
  library(here)
})
source(here("R/simulate.R")) # .bcsft_pedigree, .chr_cm_lengths, .simulate_dosage
source(here("R/single_locus.R")) # single_locus_p0, state_fractions_mb, hotelling_fractions, plot_fraction_forest
set.seed(params$seed)

# TeoNAM native v5 cM map ŌĆö the 118K-marker grid used by the caller calibration.
mc <- fread(here("data/teonam/markers_v5_gwas118k_cm.tsv"))
map <- mc[, .(chr = as.integer(chr), bp = as.integer(pos_v5), cm = as.numeric(cm))][order(chr, bp)]
markers <- map # the native markers directly carry chr/bp/cm ŌĆö no re-gridding

bp <- .bcsft_pedigree(n_bc = 2L, n_self = 3L)
ped <- bp$ped
nid <- bp$nil_id
cmlen <- .chr_cm_lengths(map)
cat(sprintf(
  "simcross %s | native markers: %d | genome: %.0f cM | N: %d\n",
  as.character(packageVersion("simcross")), nrow(markers), sum(cmlen$L), params$n
))
```

## 1. The single-locus expectation (BC2S3 transition matrix)

The genotype distribution at any single locus after F1 ŌåÆ 2 backcrosses ŌåÆ 3
selfings, from the breeding-scheme transition matrices (map- and
interference-independent ŌĆö a marginal quantity).

```{r p0}
p0 <- single_locus_p0(n_bc = 2L, n_self = 3L)
tibble(state = names(p0), expected = as.numeric(p0)) |>
  mutate(note = c("hom recurrent", "het", "hom donor")) |>
  knitr::kable(digits = 6)
cat(sprintf(
  "donor dosage = ALT + 0.5*Het = %.4f (expect 0.1250)\n",
  p0["ALT"] + 0.5 * p0["Het"]
))
```

## 2. f_i validation (Mb-based, Hotelling)

Simulate `n` NILs on the native map, tabulate per-NIL Mb fractions, and test the
population means against `p0` with the design-correct Hotelling/Wald Žć┬▓(2) (sample
= unit; empirical between-sample covariance ŌĆö **not** per-marker binomial, which
linkage overdispersion would make reject a correct sampler).

```{r fi-check}
F <- t(vapply(seq_len(params$n), function(i) {
  state_fractions_mb(markers, .simulate_dosage(ped, cmlen, markers,
    m = params$m, p = params$p, nil_id = nid
  ))
}, numeric(3)))
colnames(F) <- c("REF", "Het", "ALT")
res <- hotelling_fractions(F, p0)
knitr::kable(res$table, digits = 5)
cat(sprintf(
  "Hotelling T^2 = %.3f (df=2), chi2 p = %.4g | N_eff(ALT) = %.1f vs %d markers\n",
  res$T2, res$p_chisq,
  with(res$table[res$table$state == "ALT", ], expected * (1 - expected) / sd_fi^2),
  nrow(markers)
))
cat(if (all(res$table$in_ci)) {
  "PASS: all expectations inside 95% Wald CIs.\n"
} else {
  "CHECK: an expectation outside its CI.\n"
})
```

The fractions span ~2 orders of magnitude, so a single linear forest plot can show
magnitude *or* intervals but not both. Panel **A** (log axis) carries magnitude;
panel **B** (standardized deviation, ┬▒1.96 band) carries the pass criterion ŌĆö
scale-free.

```{r forest, fig.width=10, fig.height=3.2}
plot_fraction_forest(res$table,
  title = sprintf(
    "BC2S3 vs single-locus expectation (n=%d, native v5 map)",
    params$n
  )
)
```

`N_eff` (not the marker count) is the ergodicity-limited effective sample size:
with ~33 crossovers/genome the per-NIL fraction averages over few independent
blocks, so the right SE is `sd(f_i)/ŌłÜN`, not a per-marker binomial. This is the
same ergodicity distinction developed in
[the wideseq coverage-ergodicity note](wideseq-coverage-ergodicity.html).

## 3. Marker-resolution sweep (deferred)

::: {.callout-note title="Deferred to a native-map re-run"}
The exploratory version swept marker density (Cohen's *d* of the Het fraction vs
spacing) to show Het is undercounted only below ~0.5 Mb spacing ŌĆö a
truth-*measurement* floor distinct from low-coverage Het miscalling. That sweep
was computed on the maize consensus map; reproducing it on the TeoNAM native grid
needs a density-varying marker grid (zealhmm's `build_marker_grid` currently
integer-overflows on bp-scale maize spans ŌĆö a small fix to make before running).
Left as future work; the layer-1 truth validation above (┬¦1ŌĆō┬¦2) does not depend on
it.
:::

## Conclusion

On the TeoNAM native v5 map, the `simcross` BC2S3 simulator reproduces the
single-locus genotype fractions ŌĆö all three states inside their 95% CIs, donor
dosage 0.125 ŌĆö so the truth-generator behind the coverage-degradation sweeps is
unbiased at the single-locus margin. The only known deviation is a
marker-resolution undercount of Het at coarse spacing (deferred sweep, ┬¦3). Layer 1
(truth) is validated; caller observation/recovery under coverage is the calibration
and sweep notebooks.
