---
title: "TeoNAM composite genetic map (bcsft(1,4), B73 v5)"
subtitle: "Native per-marker cM (cm) — results view"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**What this is.** A *results* view of the composite TeoNAM genetic map — the native
per-marker cM (`cm`) built by `scripts/teonam_qtl_permap.R` → `scripts/teonam_qtl_map.R`
(per-family maps → union → Chen-2019 flanking imputation → `bcsft(1,4)` `est.map`,
markers ordered on **B73 v5**). That build takes **5+ hours**, so **this notebook only
loads the finished map and renders it — it runs no `est.map`.**

The map is **island-free on chr7**: the v2→v5 displaced block (6 markers isolated by
~5 cM gaps) was caught by the connected-component quirky finder (`scripts/map_tools.R`)
and removed, collapsing that gap **10.22 → 0.76 cM** and chr7 **154.5 → 144.85 cM**
(diagnosis in the [QTL-recovery notebook](teonam-qtl-recovery.html)).

```{r setup}
library(here)
library(data.table)
library(ggplot2)
```

```{r load}
d <- fread(here("data/teonam/teonam_v5_native.tsv"))
d <- d[!is.na(cm) & !is.na(chr_v5)][order(chr_v5, cm)]

# R/qtl `map` object: one per-chromosome cM vector, origin shifted to 0
chrs <- sort(unique(d$chr_v5))
map <- lapply(chrs, function(ch) {
  z <- d[chr_v5 == ch]
  v <- z$cm - min(z$cm)
  names(v) <- z$marker
  v
})
names(map) <- as.character(chrs)
class(map) <- "map"
```

## Map size

```{r size-table}
tab <- d[, .(markers = .N, length_cM = round(max(cm) - min(cm), 1)),
  by = .(chr = chr_v5)
][order(chr)]
tot <- data.table(chr = "total", markers = sum(tab$markers), length_cM = round(sum(tab$length_cM), 1))
knitr::kable(
  rbind(tab[, .(chr = as.character(chr), markers, length_cM)], tot),
  col.names = c("chr (v5)", "markers", "length (cM)"),
  caption = "Composite bcsft(1,4) map per chromosome (Chen 2019: avg 1461 cM/subpopulation)."
)
```

## Chromosome lengths

```{r cm-bars, fig.width=9, fig.height=6}
mk <- rbindlist(lapply(chrs, function(ch) data.table(chr = ch, cm = map[[as.character(ch)]])))
ggplot(mk, aes(x = factor(chr), y = cm)) +
  geom_segment(aes(xend = factor(chr), y = 0, yend = cm), linewidth = 6, colour = "grey85") +
  geom_segment(aes(xend = factor(chr), y = cm - 0.15, yend = cm + 0.15),
    linewidth = 6, colour = "grey30", alpha = 0.20
  ) + # marker ticks
  scale_y_reverse(expand = expansion(mult = c(0.02, 0.04))) +
  labs(x = "chromosome (v5)", y = "position (cM)") +
  theme_classic(base_size = 14)
```
