---
title: "TeoNAM STAM — GWAS Manhattan with candidate genes (Chen 2019 Fig 4C)"
subtitle: "Empirical-truth validation: reproduce Chen 2019 STAM on real TeoNAM genotypes"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**What this is.** The STAM GWAS Manhattan (Chen 2019 Fig 4C) reproduced on the
TeoNAM genotypes, with the domestication candidate genes placed at their **real
B73 v5 coordinates** (MaizeGDB `genemodel_locus` → `b73_gene_coords`), a red star
at each gene's genomic position, and the LOD = 5 dotted significance line.

This notebook **loads pre-computed CSVs and plots** (compute lives in the `scripts/`
directory): the per-marker GWAS scans (`teonam_stam_gwas_scan.R` interpolated /
`teonam_stam_gwas_family_imputed.R` family-imputed) and the candidate-gene overlap
(`teonam_candidate_overlap.R`). We show both the **family-imputed** scan (natural
scatter) and the **interpolated/densified** scan (block-terraced — the artifact of
step-interpolated genotypes, see below).

```{r setup}
library(here)
library(data.table)
library(ggplot2)
library(fastman)
library(ggrepel)
library(ggtext)
library(scales)
library(cowplot)
```

```{r helpers}
# Coordinate transformer replicating fastman_gg's internal x layout, so we can
# place gene labels / lines at arbitrary (chr, bp). From inv4m make_manhattan_plots.Rmd.
get_transformer <- function(m) {
  cmat <- do.call(rbind, lapply(
    split(m[order(m$CHR, m$BP), ], m$CHR[order(m$CHR, m$BP)]),
    function(d) {
      data.frame(
        CHR = d$CHR[1], min_bp = min(d$BP),
        width = max(d$BP) - min(d$BP),
        medgap = if (nrow(d) > 1) median(diff(sort(d$BP))) else 1
      )
    }
  ))
  cmat <- cmat[order(cmat$CHR), ]
  maxgap <- max(cmat$medgap, na.rm = TRUE)
  numc <- nrow(cmat)
  cmat$base <- 0
  cmat$midp <- 0
  cmat$midp[1] <- cmat$width[1] / 2
  for (i in 2:numc) {
    cmat$base[i] <- cmat$base[i - 1] + cmat$width[i - 1] + maxgap
    cmat$midp[i] <- cmat$base[i] + cmat$width[i] / 2
  }
  fac <- numc / cmat$midp[numc]
  cmat$basef <- fac * cmat$base
  function(chr, bp) {
    i <- match(chr, cmat$CHR)
    fac * (bp - cmat$min_bp[i]) + cmat$basef[i]
  }
}

LOD <- 5 # Chen 2019: P < 1e-5 (LOD = 5)

#' STAM GWAS Manhattan with candidate genes at their true v5 positions.
#' Odd chromosomes black, even gray75; y-axis expanded one major-tick unit above
#' the top tick; gene labels repelled into the headroom above the maximum peak
#' (segments back to the red star at each gene's local peak).
#' @param scan_csv per-marker scan (SNP, CHR, BP[v5], P)
#' @param title    plot title
#' @param out_png  optional path to save a PNG
plot_stam_manhattan <- function(scan_csv, title, out_png = NULL, mark = NULL) {
  scan <- if (is.data.frame(scan_csv)) as.data.frame(scan_csv) else as.data.frame(fread(scan_csv))
  scan <- scan[is.finite(scan$P) & scan$P > 0, ]
  m <- scan[order(scan$CHR, scan$BP), c("SNP", "CHR", "BP", "P")]

  ov <- fread(here("results/sim/teonam/stam_candidate_overlap.csv")) # symbol, chr, start (v5)
  # star height = local max -log10P within +/-500 kb IN THIS scan
  ov[, y := vapply(seq_len(.N), function(i) {
    w <- m$P[m$CHR == chr[i] & abs(m$BP - start[i]) <= 5e5]
    if (length(w)) -log10(min(w)) else NA_real_
  }, numeric(1))]
  tr <- get_transformer(m)
  ov[, BPn := tr(chr, start)]
  ov <- ov[is.finite(y)]

  mp <- -log10(min(m$P)) # maximum peak
  ticks <- pretty(c(0, mp))
  unit <- diff(ticks)[1]
  ytop <- max(ticks) + unit # one major-tick unit of headroom

  p <- fastman_gg(
    m = m, snp = "SNP", col = c("black", "gray75"), maxP = mp,
    genomewideline = NULL, suggestiveline = NULL,
    ylab = expression(-log[10](italic(P))), xlab = ""
  ) +
    geom_hline(yintercept = LOD, linetype = "dotted", linewidth = 0.7, color = "black") +
    geom_point(
      data = ov, aes(x = BPn, y = y), inherit.aes = FALSE,
      shape = 8, size = 3, stroke = 1, color = "red"
    ) +
    ggrepel::geom_text_repel(
      data = ov, aes(x = BPn, y = y, label = symbol),
      inherit.aes = FALSE, fontface = "italic", size = 5,
      segment.color = "grey50", min.segment.length = 0, box.padding = 0.5,
      max.overlaps = Inf, ylim = c(mp + 0.3 * unit, ytop)
    ) + # above the max peak
    ggtitle(title) +
    theme_classic(base_size = 18) +
    scale_y_continuous(
      breaks = ticks, limits = c(0, ytop),
      expand = expansion(mult = c(0.01, 0))
    ) +
    theme(
      plot.title = element_markdown(hjust = 1, size = 16, face = "bold"),
      legend.position = "none",
      panel.grid.major.y = element_line(color = "grey90", linewidth = 0.2)
    )

  if (!is.null(mark)) { # dashed marker at an arbitrary locus (e.g. the chr3 ghost)
    xm <- tr(mark$chr, mark$bp)
    p <- p + geom_vline(xintercept = xm, linetype = "dashed", color = "red", linewidth = 0.5)
  }
  if (!is.null(out_png)) ggsave(out_png, p, width = 9, height = 4.6, dpi = 200, bg = "white")
  p
}

#' Chen 2019 Fig 4A: JLM QTL genomic distribution as a lollipop plot.
#' Full 10-chromosome axis is obtained by passing the dense scan to fastman_gg
#' with fully transparent points (col = transparent); the JLM QTL are overlaid as
#' blue stems + points, candidate genes labeled bold-italic above.
#' @param jlm_txt  TASSEL StepwiseOLSModelFitter output (stam_jlm_native1.txt): Name,
#'   Locus, Position (v5 bp), `pr>F`
#' @param scan_csv dense per-marker scan, only used (transparent) to lay out the axis
plot_stam_lollipop <- function(jlm_txt, scan_csv, title, out_png = NULL) {
  scan <- as.data.frame(fread(scan_csv))
  scan <- scan[is.finite(scan$P) & scan$P > 0, ]
  maxis <- scan[order(scan$CHR, scan$BP), c("SNP", "CHR", "BP", "P")]

  j <- fread(jlm_txt)
  j <- j[!(Name %in% c("mean", "Family", "Error"))]
  q <- data.table(
    SNP = j$Name, CHR = as.integer(j$Locus), BP = as.integer(j$Position),
    logP = -log10(as.numeric(j$`pr>F`))
  )
  ov <- fread(here("results/sim/teonam/stam_candidate_overlap.csv")) # symbol, chr, start (v5)
  ov <- ov[symbol %in% c("ts2", "tb1", "tru1", "zmm16", "tsh4")] # JLM Fig-4A candidates (an1 is GWAS-only)
  q[, gene := ""] # assign each candidate gene to its NEAREST JLM QTL (1:1)
  for (i in seq_len(nrow(ov))) {
    idx <- which(q$CHR == ov$chr[i])
    if (length(idx)) {
      k <- idx[which.min(abs(q$BP[idx] - ov$start[i]))]
      q$gene[k] <- ov$symbol[i]
    }
  }
  tr <- get_transformer(maxis)
  q[, BPn := tr(CHR, BP)]

  maxlogP <- max(q$logP)
  brks <- seq(0, floor(maxlogP / 10) * 10, by = 10)
  ytop <- maxlogP * 1.18 # headroom for the top label

  p <- fastman_gg(
    m = maxis, snp = "SNP", col = c("#00000000", "#00000000"),
    maxP = maxlogP, genomewideline = NULL, suggestiveline = NULL,
    ylab = expression(-log[10](italic(P))), xlab = ""
  ) +
    geom_segment(
      data = q, aes(x = BPn, xend = BPn, y = 0, yend = logP),
      inherit.aes = FALSE, color = "blue", linewidth = 0.5
    ) +
    geom_point(
      data = q, aes(x = BPn, y = logP), inherit.aes = FALSE,
      color = "blue", size = 2.5
    ) +
    ggrepel::geom_text_repel(
      data = q[gene != ""], aes(x = BPn, y = logP, label = gene),
      inherit.aes = FALSE, fontface = "bold.italic", size = 5,
      segment.color = NA, box.padding = 0.4, point.padding = 0.3,
      nudge_y = 0.05 * maxlogP, seed = 1, max.overlaps = Inf
    ) +
    ggtitle(title) + theme_classic(base_size = 18) +
    scale_y_continuous(breaks = brks, limits = c(0, ytop), expand = expansion(mult = c(0.01, 0))) +
    theme(
      plot.title = element_markdown(hjust = 1, size = 16, face = "bold"),
      legend.position = "none",
      panel.grid.major.y = element_line(color = "grey90", linewidth = 0.2)
    )
  if (!is.null(out_png)) ggsave(out_png, p, width = 9, height = 4, dpi = 200, bg = "white")
  p
}

#' Coverage-sweep line-Manhattan (geom_line, one line per coverage).
#' −log10P vs genomic position along the union grid; `group = interaction(CHR,
#' coverage)` so lines never connect across chromosomes. Color = coverage on a
#' viridis ramp, the λ = ∞ truth baseline in black. LOD = 5 dotted. NO gene labels.
#' Reused for the control / nNIL / RTIGER panels of the degradation composite.
#' @param sweep_csv combined sweep CSV (SNP, CHR, BP[v5], P, coverage; coverage=Inf
#'   for the λ=∞ baseline rows)
#' @param title panel title (markdown)
#' @param legend show the coverage legend (FALSE for composite panels sharing one)
#' @param base_size theme base size
plot_sweep_line <- function(sweep_csv, title, legend = TRUE, base_size = 14) {
  sw <- as.data.table(fread(sweep_csv))
  sw <- sw[is.finite(P) & P > 0]
  sw[, logP := -log10(P)]
  # x layout from the union grid (every coverage carries the identical marker set)
  tr <- get_transformer(as.data.frame(sw[coverage == sw$coverage[1], .(CHR, BP)]))
  sw[, BPn := tr(CHR, BP)]

  # Legend order: infinity (truth) at top, then coverage in descending order.
  cov_levels <- c("∞", "20", "10", "5", "1", "0.5", "0.2", "0.1")
  sw[, cov_lab := factor(ifelse(is.finite(coverage), formatC(coverage), "∞"),
    levels = cov_levels
  )]
  # Colour mapping is fixed by coverage value (0.1 = darkest viridis ... 20 =
  # lightest; ∞ = black), independent of the legend order above.
  cov_asc <- c("0.1", "0.2", "0.5", "1", "5", "10", "20")
  pal <- setNames(grDevices::hcl.colors(7, "viridis"), cov_asc)
  pal["∞"] <- "black"
  axis_df <- sw[, .(mid = (min(BPn) + max(BPn)) / 2), by = CHR][order(CHR)]
  ytop <- 20 # shared y-limit across all panels for cross-method comparison

  # Draw order (z-stack): 20x first, then descending to 0.1x, and ∞ LAST so the
  # darker low-coverage / black-truth profiles sit on top. dord = -coverage sorts
  # finite coverage descending (20 -> -20 first); ∞ -> Inf sorts last.
  sw[, dord := ifelse(is.finite(coverage), -coverage, Inf)]

  ggplot(
    sw[order(dord, CHR, BP)],
    aes(x = BPn, y = logP, color = cov_lab, group = interaction(CHR, coverage))
  ) +
    geom_line(linewidth = 0.4) +
    geom_hline(yintercept = LOD, linetype = "dotted", linewidth = 0.7, color = "grey30") +
    scale_color_manual(values = pal, name = "coverage (×)", drop = FALSE) +
    scale_x_continuous(breaks = axis_df$mid, labels = axis_df$CHR, expand = c(0.01, 0)) +
    scale_y_continuous(limits = c(0, ytop), expand = expansion(mult = c(0.01, 0))) +
    labs(x = "chromosome", y = expression(-log[10](italic(P))), title = title) +
    theme_classic(base_size = base_size) +
    theme(
      plot.title = element_markdown(hjust = 0, size = base_size, face = "bold"),
      panel.grid.major.y = element_line(color = "grey90", linewidth = 0.2),
      # legend INSIDE the plotting area, top-right corner (top panel only)
      legend.position = if (legend) c(0.995, 0.97) else "none",
      legend.justification = c(1, 1),
      legend.background = element_rect(fill = scales::alpha("white", 0.7), colour = NA),
      legend.key.size = unit(0.85, "lines"),
      legend.title = element_text(size = base_size - 3),
      legend.text = element_text(size = base_size - 4),
      legend.margin = margin(1, 3, 1, 3)
    ) +
    guides(color = guide_legend(override.aes = list(linewidth = 1.2)))
}
```

## Candidate genes: closest marker vs most-significant marker (v5)

For each STAM candidate gene (B73 v5), two different "nearby marker" choices from
the pre-JLM marker set (the 51,004-marker family-imputed GWAS scan): the
**closest** marker (smallest bp distance to the gene body) and the **window-max**
marker (highest −log₁₀P within ±500 kb — a 1 Mb window centered on the gene). They
are **not the same marker** at any gene: the closest marker sits essentially on
the gene but carries little signal, while the significant local signal is a
nearby but distinct marker. This is the reference for what "near the gene" means
in the Fig 4A lollipop below (which labels each gene onto its nearest JLM QTL).

```{r candidate-markers}
sc <- fread(here("data/teonam/stam_gwas_scan_family_imputed.csv"))
sc <- sc[is.finite(P) & P > 0][, logP := -log10(P)]
cand <- c(
  ts2 = "STAM1.1", tb1 = "STAM1.2", tru1 = "STAM3.1",
  zmm16 = "STAM3.2", tsh4 = "STAM7.1", an1 = "GWAS(chr1)"
)
sym <- fread(here("data/ref/maize_v5_gene_symbols.tsv"))
setnames(sym, c("gene_id", "symbol"))
coord <- fread(here("data/ref/b73_gene_coords.tsv"),
  header = FALSE,
  col.names = c("gene_id", "chr", "start", "end")
)
gg <- merge(sym[symbol %in% names(cand)], coord, by = "gene_id")
gg[, chr := as.integer(sub("^chr", "", chr))][, qtl := cand[symbol]]
WIN <- 5e5 # +/-500 kb around the gene body (1 Mb window)
ctab <- rbindlist(lapply(seq_len(nrow(gg)), function(i) {
  ch <- gg$chr[i]
  s <- gg$start[i]
  e <- gg$end[i]
  m <- sc[CHR == ch]
  d <- pmax(0L, pmax(s - m$BP, m$BP - e))
  kc <- which.min(d) # closest to gene body
  w <- m[BP >= s - WIN & BP <= e + WIN]
  kw <- which.max(w$logP) # max signal in window
  data.table(
    gene = gg$symbol[i], QTL = gg$qtl[i], chr = ch, `gene start (v5)` = s,
    `closest bp (v5)` = m$BP[kc], `closest dist (bp)` = d[kc],
    `closest -log10P` = round(m$logP[kc], 2),
    `window-max bp (v5)` = w$BP[kw],
    `window-max dist (bp)` = pmax(0L, pmax(s - w$BP[kw], w$BP[kw] - e)),
    `window-max -log10P` = round(w$logP[kw], 2)
  )
}))
setorder(ctab, chr, `gene start (v5)`)
knitr::kable(ctab, format.args = list(big.mark = ","), caption = paste(
  "STAM candidate genes (B73 v5): the CLOSEST marker vs the highest -log10P marker",
  "within +/-500 kb (family-imputed GWAS scan, 51,004 markers). Positions/distances",
  "in v5 bp; the two markers are distinct at every gene."
))
```

## JLM QTL genomic distribution (Fig 4A)

The STAM QTL detected by JLM as a lollipop plot across all 10 chromosomes
(height = −log10 P from TASSEL's stepwise model), candidate genes bold-italic
above the corresponding QTL. Reproduces Chen 2019 Fig 4A.

```{r lollipop-jlm, fig.width=9, fig.height=4}
p_lolli <- plot_stam_lollipop(
  here("data/teonam/tassel/stam_jlm_native1.txt"),
  here("data/teonam/stam_gwas_scan_family_imputed.csv"),
  "STAM — JLM QTL genomic distribution (TeoNAM)",
  here("results/sim/teonam/stam_jlm_lollipop.png")
)
p_lolli
```

## Family-imputed genotypes (the faithful Fig 4C)

Each marker tested only on its actually-typed lines — natural GWAS scatter.

```{r manhattan-family-imputed, fig.width=9, fig.height=4.4}
p_unimp <- plot_stam_manhattan(
  here("data/teonam/stam_gwas_scan_family_imputed.csv"),
  "STAM — GWAS (TeoNAM, family-imputed), candidate genes at v5 positions",
  here("results/sim/teonam/stam_gwas_manhattan_annotated.png")
)
p_unimp
```

## Step-interpolated grid — artifact only

The same scan on the step-interpolated full GWAS set (`teonam_map_v5_gwas`, 51,004
markers, cross-family gaps densified). This is shown purely to document the
interpolation artifact. Ancestry is a piecewise-constant block signal, so step
interpolation makes contiguous markers identical → identical p-values → **terraced
plateaus** (peaks still correctly located). The GWAS figure above uses the same
markers *without* densification (family-imputed, per-marker).

```{r manhattan-interpolated, fig.width=9, fig.height=4.4}
p_interp <- plot_stam_manhattan(
  here("data/teonam/stam_gwas_scan_interpolated.csv"),
  "STAM — GWAS (TeoNAM, interpolated), candidate genes at v5 positions",
  here("results/sim/teonam/stam_gwas_manhattan_interpolated_annotated.png")
)
p_interp
```

## Composite (Fig 4A lollipop + Fig 4C family-imputed + interpolated)

```{r composite, fig.width=9, fig.height=13}
composite <- cowplot::plot_grid(
  p_lolli, p_unimp, p_interp,
  ncol = 1, labels = c("A", "B", "C"), label_size = 20, align = "v", axis = "lr"
)
ggsave(here("results/sim/teonam/stam_composite.png"), composite,
  width = 9, height = 13, dpi = 200, bg = "white"
)
composite
```

## RTIGER ancestry vs sequencing coverage (degradation sweep)

The panel-C baseline above is the **λ = ∞, perfect-caller** curve (GWAS on the
interpolated *truth* genotypes). All sweeps below run on the same full
**51,004**-marker GWAS grid as the Fig-4C panels above. Here we ask
how low-coverage sequencing degrades
that signal **through the RTIGER ancestry caller**: for each family × coverage
`λ ∈ {0.1, 0.2, 0.5, 1, 5, 10, 20}×` we simulate reads from the real genotypes
(`.draw_counts`, `pi_floor=0, k_decay=1, error=0.01`), call ancestry with
`call_ancestry(caller = "rtiger", rigidity = 2)` (the calibrated `rigidity_star`),
step-interpolate the recovered block onto the same union grid, and rerun the same
`STAM ~ Family + marker` scan. Compute lives in
`scripts/teonam_rtiger_sweep.R --generate`; this chunk only **loads** the combined
sweep CSV and plots.

The read as a **line-Manhattan** (`geom_line`, one line per coverage; chromosomes
never connect because `group = interaction(CHR, coverage)`): coverage on a viridis
ramp, the λ = ∞ truth baseline in **black** as the reference curve, LOD = 5 dotted.
The striking result is how **robust** RTIGER is — ancestry is a long-block signal,
so even 0.1× (~9.5 % of markers carrying a read) reconstructs the blocks well
enough that the *tb1* peak only slips from ≈17.3 (baseline) to ≈16.1. Degradation
is real but lives mostly in HET recovery (het-recall 0.77 at 0.1× → 1.00 at ≥20×),
not in the additive QTL peaks.

```{r line-manhattan, fig.width=9, fig.height=4.6}
p_sweep <- plot_sweep_line(
  here("results/sim/teonam/stam_gwas_rtiger_sweep.csv"),
  "STAM — RTIGER ancestry, GWAS −log10P vs coverage",
  base_size = 15
)
ggsave(here("results/sim/teonam/stam_gwas_rtiger_sweep_manhattan.png"),
  p_sweep,
  width = 9, height = 4.6, dpi = 200, bg = "white"
)
p_sweep
```

## Coverage-degradation composite: interpolation control / nNIL / RTIGER / LB-Impute

The four-panel contrast (`agent/teonam-control-sweep-plan.md`). All four degrade
the **identical** truth mosaics over the same coverage grid
`λ ∈ {0.1, 0.2, 0.5, 1, 5, 10, 20}×` (+ the λ = ∞ interpolated-truth baseline) and
feed the same `STAM ~ Family + marker` GWAS — only the genotype caller differs:

- **A — interpolation control** (`nilHMM::call_gl(prior = "hwe")` + per-RIL
  step-interpolation; `scripts/teonam_control_sweep.R`): the deliberately het-prone
  no-HMM caller — a single ALT read is called HET via the `2p(1−p)` HWE prior term.
- **B — nNIL** (`call_ancestry(caller = "nnil", rrate = rrate_star, err = 0.01)`,
  `f_1 = 0.08`, `f_2 = 0.15`; `scripts/teonam_nnil_sweep.R`): count emission + geometric
  duration.
- **C — RTIGER** (`caller = "rtiger", rigidity = 2`; `scripts/teonam_rtiger_sweep.R`):
  count emission + rigidity duration.
- **D — LB-Impute** (`call_ancestry(caller = "lbimpute", unit = "cm", recombdist = recombdist_star ≈ 5.5)`;
  `scripts/teonam_lbimpute_sweep.R`): coverage-aware emission + distance-dependent
  (map-aware cM) transition with a double-recombination penalty, full-chromosome
  Viterbi (nilHMM native port of Fragoso et al. 2014). `recombdist` is the
  donor-fragment-Dice-optimal `recombdist_star` from the BC1S4 calibration
  (`scripts/02_calibrate.R`), read from `calib_params.csv` — not hardcoded.

The **contrast is het handling**. The three duration/distance-aware HMMs (B, C, D) tie
neighbouring markers into ancestry blocks, so a lone ALT read cannot open a het call —
het-call fraction stays near the ~8 % truth level and the STAM peaks hold near the
baseline down to 0.1×. LB-Impute (D), at the calibrated `recombdist_star ≈ 5.5 cM`,
degrades gracefully like B and C rather than over-smoothing: its *tb1* peak holds
near the baseline across the grid (≈16.3 at 0.1× vs ≈17.3 baseline), never
over-shooting. (The tighter calibrated distance scale allows more transitions than
the earlier hardcoded `recombdist = 50`, which had produced over-long imputed
homozygous blocks and an over-shot peak.) The control (A) decides each site independently, so at low
coverage the HWE prior inflates het to ~20 % (het **excess**) and the *tb1* peak
degrades the other way (≈13.4 at 0.1× vs ≈17.3 baseline). `geom_line`, colored by
coverage, no gene labels, LOD = 5, chromosomes never connect
(`group = interaction(CHR, coverage)`).

**A pericentromeric block-imputation "ghost" on chromosome 3 at 0.1×.** At the
lowest coverage a spurious peak appears near the **middle of chr3 (≈110–115 Mb, on
the chromosome axis label)** — *proximal to* both chr3 STAM candidates, *tru1*
(STAM3.1, ≈152 Mb) and *zmm16* (STAM3.2, ≈176 Mb), which sit in the right half of
the chromosome. It is a **false positive**: the λ = ∞ baseline has no peak there
(≈8.5), whereas the 0.1× line rises to ≈12.4 (excess ≈+3.9 over baseline). The
ghost sits on the proximal shoulder of the chr3 **pericentromere** — the
restricted-recombination zone reported at ≈80–90 Mb in B73 v5 (pan-genome cenhap
boundaries) and independently visible in this study's own native cM map, where
the recombination rate collapses to ≈0.06 cM/Mb across ≈80–100 Mb (~40× below the
≈2 cM/Mb near the telomeres). Recombination is so suppressed there that cM barely
advances across tens of Mb, so with the map-aware cM transition adjacent markers
are separated by ~zero cM and the state stays highly persistent; at 0.1× (only
~9.5 % of markers carrying a read) LB-Impute lays down a long spurious homozygous
block through this low-recombination valley that tracks the phenotype. It is a **block-caller**
effect: the no-HMM interpolation control (A) shows **no** central ghost (excess
≈0), nNIL (B) and RTIGER (C) show milder versions (excess ≈+2.3–2.6), and
**LB-Impute (D) is the most affected** (≈+3.9).

Note this ghost is a genuine consequence of the **calibrated** `recombdist_star`
(≈5.5 cM): with the earlier hardcoded `recombdist = 50` the chr3 middle stayed near
baseline (excess ≈+1) and the over-imputation instead landed on the *right*,
inflating the real *zmm16* region (≈16.7 vs ≈14.8 baseline). The tighter calibrated
distance scale represents the real peaks more faithfully but, interacting with
pericentromeric cM compression that the BC1S4 calibration sim does not model,
produces this central false block at extreme low coverage. Treat the 0.1× chr3-mid
peak as a coverage-limit artifact, not signal.

```{r het-fraction-table}
het_ctrl <- fread(here("results/sim/teonam/stam_control_het_fraction.csv"))[, .(coverage, control = round(het_frac, 3))]
het_nnil <- fread(here("results/sim/teonam/stam_nnil_het_fraction.csv"))[, .(coverage, nnil = round(het_frac, 3))]
het_tab <- merge(het_ctrl, het_nnil, by = "coverage")
het_tab[] # HET-call fraction vs coverage: control (het excess at low λ) vs nNIL (flat ~truth)
```

```{r sweep-composite, fig.width=9, fig.height=16}
# Legend lives INSIDE the top panel (A), top-right; the rest carry none.
pA <- plot_sweep_line(here("results/sim/teonam/stam_gwas_control_sweep.csv"),
  "STAM — interpolation control (GL + HWE)",
  legend = TRUE
)
pB <- plot_sweep_line(here("results/sim/teonam/stam_gwas_nnil_sweep.csv"),
  "STAM — nNIL ancestry",
  legend = FALSE
)
pC <- plot_sweep_line(here("results/sim/teonam/stam_gwas_rtiger_sweep.csv"),
  "STAM — RTIGER ancestry",
  legend = FALSE
)
pD <- plot_sweep_line(here("results/sim/teonam/stam_gwas_lbimpute_sweep.csv"),
  "STAM — LB-Impute ancestry (map-aware cM)",
  legend = FALSE
)

composite <- cowplot::plot_grid(pA, pB, pC, pD,
  ncol = 1, align = "v", axis = "lr",
  labels = c("A", "B", "C", "D"), label_size = 18
)
ggsave(here("results/sim/teonam/stam_sweep_composite.png"), composite,
  width = 9, height = 16, dpi = 200, bg = "white"
)
composite
```

## Is the chr3-mid ghost a real QTL? Structure/kinship control at 0.1× (MLM Q + K, LOCO)

The composite above is an OLS scan (`STAM ~ Family + marker`), which models neither
kinship nor fine population structure — exactly what inflates broad signals in
low-recombination regions. Chen 2019's GWAS instead used a **mixed model with
Q = 5 PCs + K = kinship**. To test whether the 0.1× chr3-pericentromere ghost is a
real QTL or a structure/LD confound, we re-scan the **same 0.1× LB-Impute
genotypes** three ways (compute in `scripts/teonam_mlm_loco_lb01.R` — an EMMAX port:
VanRaden K + 5 PCs + REML δ + P3D GLS; genotype matrix from
`scripts/teonam_lbimpute_sweep.R --generate --save-matrix`):

- **A — OLS (Family):** the composite's model. Ghost present (chr3-mid ≈12.4).
- **B — MLM (Q + K):** kinship + structure controlled (λ_GC = 1.10). The ghost
  **collapses to ≈4.7**, back below the LOD-5 line — the signature of a structure/LD
  confound, not a QTL. (The real *tru1*/*zmm16* peaks are pulled down here too — the
  known MLM proximal over-correction on their own chromosome.)
- **C — LOCO:** kinship built excluding the tested chromosome. The real QTL return
  (*tru1*/159/*zmm16* ≈ 9.6–12.6) — **but so does the ghost (≈10.7)**, because
  dropping chr3 from the kinship removes the correction for its own pericentromeric
  structure. At 0.1× the ghost becomes indistinguishable from a real QTL under LOCO.

Red dashed line = the chr3 pericentromere ghost (~112 Mb); the *tru1*/*zmm16* stars
mark the real distal QTL. **Takeaway:** the ghost is a structure/LD confound that
full-K MLM controls even at 0.1×, but low-coverage block imputation inflates it
enough to nearly survive Q + K and to fully resurface under LOCO — so LOCO peaks in
low-recombination regions cannot be trusted at low coverage.

```{r ghost-mlm-loco, fig.width=9, fig.height=13}
lb01 <- fread(here("results/sim/teonam/stam_gwas_lbimpute_lambda0.1.csv")) # OLS, Family only
ml <- fread(here("results/sim/teonam/mlm_loco_scan_lb01.csv")) # MLM Q+K (p_full) + LOCO (p_loco)

ols_s <- lb01[is.finite(P) & P > 0, .(SNP, CHR, BP, P)]
full_s <- ml[is.finite(p_full) & p_full > 0, .(SNP = marker, CHR = chr, BP = bp, P = p_full)]
loco_s <- ml[is.finite(p_loco) & p_loco > 0, .(SNP = marker, CHR = chr, BP = bp, P = p_loco)]

ghost <- list(chr = 3, bp = 112e6) # chr3 pericentromere ghost locus
pO <- plot_stam_manhattan(ols_s, "STAM — LB-Impute 0.1×, OLS (Family)", mark = ghost)
pF <- plot_stam_manhattan(full_s, "STAM — LB-Impute 0.1×, MLM (Q + K)", mark = ghost)
pL <- plot_stam_manhattan(loco_s, "STAM — LB-Impute 0.1×, LOCO", mark = ghost)

ghost_grid <- cowplot::plot_grid(pO, pF, pL,
  ncol = 1, align = "v", axis = "lr",
  labels = c("A", "B", "C"), label_size = 18
)
ggsave(here("results/sim/teonam/stam_ghost_mlm_loco.png"), ghost_grid,
  width = 9, height = 13, dpi = 200, bg = "white"
)
ghost_grid
```

```{r ghost-region-table}
regmax <- function(dt, pcol, lo, hi) {
  z <- dt[chr == 3 & bp >= lo * 1e6 & bp <= hi * 1e6]
  pv <- z[[pcol]][is.finite(z[[pcol]]) & z[[pcol]] > 0]
  if (length(pv)) round(max(-log10(pv)), 1) else NA_real_
}
olsdt <- copy(lb01)
setnames(olsdt, c("CHR", "BP", "P"), c("chr", "bp", "p_ols"))
gtab <- data.table(
  region = c("pericentromere 80–120 (ghost)", "tru1 ~152", "159-mid", "zmm16 ~176"),
  lo = c(80, 150, 157, 174), hi = c(120, 154, 162, 178)
)
gtab[, `:=`(
  OLS = mapply(function(a, b) regmax(olsdt, "p_ols", a, b), lo, hi),
  `MLM (Q+K)` = mapply(function(a, b) regmax(ml, "p_full", a, b), lo, hi),
  LOCO = mapply(function(a, b) regmax(ml, "p_loco", a, b), lo, hi)
)]
gtab[, .(region, OLS, `MLM (Q+K)`, LOCO)] # chr3 max -log10P at 0.1x; LOD = 5
```
