---
title: "TeoNAM STAM — OLS GWAS on the family-imputed map panel (Fig 4C)"
subtitle: "Per-marker across-family OLS on the FSFHap family-imputed map-panel genotypes — computation, model & QC"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: false
execute:
  warning: false
  message: false
  echo: true
  freeze: auto
---

**What this is.** The *computation* behind the STAM OLS GWAS scan on the family-imputed
map panel (`data/teonam/stam_gwas_scan_family_imputed.csv`), tracked in its own right. It
ports the throwaway `scripts/teonam_stam_gwas_family_imputed.R` into a rendered notebook so
the scan — not just its Manhattan — is reproducible from primary data. The annotated
Fig-4C Manhattan that *consumes* this CSV lives in
[`teonam-qtl-recovery`](teonam-qtl-recovery.html); this notebook **produces** it.

The genotypes are **FSFHap family-imputed** (`teonam_map_v5_gwas`, 51,004 markers — Chen's
map construction, complete *within family*). Each marker is tested only on the families
that typed it (cross-family gaps left as NA, **not** densified), giving the natural GWAS
scatter with no step-interpolation plateau artifact. (`family_imputed` names exactly this
state: FSFHap within-family imputation only, *without* cross-family step-interpolation —
contrast the `interpolated` matrix used by the MLM notebook.)

```{r setup}
library(here)
library(data.table)
library(parallel)
library(readxl)
library(ggplot2)
theme_set(theme_bw(base_size = 12))
THREADS <- max(1L, detectCores() - 2L)
LOD <- 5 # Chen 2019 significance line: P < 1e-5
```

## Inputs (all primary data — no `.rds` intermediates)

```{r inputs}
# FSFHap-imputed recoded dosages (complete within family): 1237 lines x 51,482 markers; NA = cross-family untyped (0/1/2 = W22 / het / teosinte)
geno <- fread(here("data/teonam/TeoNAM_genotype_clean.csv"))

# robust key = family + line, e.g. "TIL01" + "A001"; family from the population column
key <- paste0(sub("^W22", "", geno[[2]]), sub("^.*Line_", "", geno[[1]]))
fam <- factor(sub("^W22", "", geno[[2]]))

# phenotype: STAM (staminate spikelet), joined by line key
ph <- as.data.frame(read_excel(
  here("data/teonam/9250682/TeoNAM_1257RILs_22traits_phenotype_data.xlsx")
))
names(ph)[1] <- "line"
y <- suppressWarnings(as.numeric(setNames(ph$STAM, ph$line)[key]))

# v5 coordinates + consensus cM; restrict the scan to markers that lifted to v5
mc <- fread(here("data/teonam/map_v5_coe2008.tsv"))
chr_by <- setNames(mc$chr_v5, mc$marker)
pos_by <- setNames(mc$pos_v5, mc$marker)
mk <- intersect(names(geno)[-(1:3)], mc$marker)

data.table(
  quantity = c(
    "lines", "markers (raw)", "markers (lifted, scanned)",
    "STAM phenotyped", "families"
  ),
  value = c(
    nrow(geno), ncol(geno) - 3L, length(mk),
    sum(!is.na(y)), nlevels(fam)
  )
)
```

## The model

For each marker $m$, restrict to the lines $i$ with a non-missing dosage
$g_{im}\in\{0,1,2\}$ **and** a non-missing STAM value $y_i$, then compare two
nested least-squares fits:

$$
\textbf{Null:}\quad y_i = \mu + \sum_{f} \beta_f \,\mathbb{1}[\text{family}_i = f] + \varepsilon_i
$$
$$
\textbf{Full:}\quad y_i = \mu + \sum_{f} \beta_f \,\mathbb{1}[\text{family}_i = f] + \gamma\, g_{im} + \varepsilon_i
$$

The marker enters as a **single additive dosage effect $\gamma$ shared across
families** (1 numerator df). Significance is the marginal $F$-test of $\gamma$:

$$
F = \frac{(\mathrm{RSS}_0 - \mathrm{RSS}_1)/1}{\mathrm{RSS}_1/(n - \mathrm{rank})},
\qquad P = \Pr\big(F_{1,\,n-\mathrm{rank}} > F\big).
$$

Family enters as a fixed categorical covariate (levels dropped to those present for
that marker) — it plays the role of the structure term. **This is ordinary least
squares (OLS): there is no kinship/polygenic random effect $K$.** Chen 2019 instead
used a mixed model `y ~ Q(5 PCs) + SNP + K(IBS)`; the missing random $K$ is what
makes ours "not mixed", so peak heights here are expected to run hotter — quantifying
that inflation ($\lambda_{GC}$, QQ) is a separate analysis.

## The scan

```{r scan}
scan1 <- function(m) {
  g <- geno[[m]]
  ok <- !is.na(g) & !is.na(y)
  if (sum(ok) < 20) {
    return(c(NA_real_, sum(ok)))
  }
  gg <- g[ok]
  if (stats::sd(gg) == 0) {
    return(c(NA_real_, sum(ok)))
  }
  yy <- y[ok]
  ff <- droplevels(fam[ok])
  Xr <- if (nlevels(ff) > 1) model.matrix(~ff) else matrix(1, sum(ok), 1)
  n <- sum(ok)
  RSS0 <- sum(lm.fit(Xr, yy)$residuals^2)
  fit <- lm.fit(cbind(Xr, gg), yy)
  RSS1 <- sum(fit$residuals^2)
  df2 <- n - fit$rank
  if (df2 <= 0 || RSS1 <= 0) {
    return(c(NA_real_, n))
  }
  c(pf(((RSS0 - RSS1) / 1) / (RSS1 / df2), 1, df2, lower.tail = FALSE), n)
}
out <- mclapply(mk, scan1, mc.cores = THREADS)
P <- vapply(out, `[`, numeric(1), 1)
N <- vapply(out, `[`, numeric(1), 2)

scan <- data.table(
  SNP = mk, CHR = as.integer(chr_by[mk]),
  BP = as.integer(pos_by[mk]), P = P,
  n = as.integer(N)
)[order(CHR, BP)]
fwrite(scan, here("data/teonam/stam_gwas_scan_family_imputed.csv")) # canonical output
```

## QC and results

```{r results}
fin <- scan[is.finite(P) & P > 0]
data.table(
  metric = c(
    "markers scanned", "markers tested (finite P)",
    "median lines / marker", "max -log10P",
    sprintf("genome-wide sig. (P < 1e-%d)", LOD)
  ),
  value = c(
    nrow(scan), nrow(fin), median(scan$n),
    round(max(-log10(fin$P)), 2), sum(fin$P < 10^(-LOD))
  )
)
```

```{r top-hits}
top <- fin[order(P)][1:10, .(SNP, CHR, BP, n, logP = round(-log10(P), 2))]
knitr::kable(top, caption = "Top 10 STAM associations (family-imputed scan)")
```

```{r per-marker-n, fig.width=7, fig.height=2.8}
ggplot(scan, aes(n)) +
  geom_histogram(bins = 60, fill = "grey30") +
  labs(
    x = "lines tested per marker (families typing that marker)",
    y = "markers", title = "Per-marker sample size — the source of the natural scatter"
  )
```

```{r manhattan, fig.width=9, fig.height=3.4}
# lightweight Manhattan (the annotated Fig-4C version lives in teonam-qtl-recovery)
m <- fin[order(CHR, BP)]
off <- m[, .(mx = max(BP)), by = CHR][order(CHR)]
off[, base := cumsum(shift(mx, fill = 0))]
m <- merge(m, off[, .(CHR, base)], by = "CHR")
m[, x := BP + base]
axis <- m[, .(mid = (min(x) + max(x)) / 2), by = CHR][order(CHR)]
ggplot(m, aes(x, -log10(P), colour = factor(CHR %% 2))) +
  geom_point(size = 0.5) +
  geom_hline(yintercept = LOD, linetype = "dotted") +
  scale_colour_manual(values = c("grey60", "black"), guide = "none") +
  scale_x_continuous(breaks = axis$mid, labels = axis$CHR, expand = c(0.01, 0)) +
  labs(
    x = "chromosome", y = expression(-log[10](italic(P))),
    title = "STAM family-imputed GWAS (OLS: STAM ~ Family + marker)"
  )
```

## Notes

- This notebook is the **canonical, tracked source** for
  `stam_gwas_scan_family_imputed.csv`; it supersedes the untracked
  `scripts/teonam_stam_gwas_family_imputed.R` (identical model, ported verbatim).
- Model is **OLS (a fixed-effects linear model) with no kinship term** —
  deliberately faithful to the throwaway script, *not* to Chen's MLM. The peak-height
  / $\lambda_{GC}$ comparison against Chen's `Q + K` mixed model is deferred to a
  dedicated inflation notebook.
```
