---
title: "ZEAL/BZea Stem-Pubescence (StPu) GWAS + JLM on the SNP50K panel — FDR ≤ 0.05"
subtitle: "Macrohair/pubescence sibling of the StPi notebook: the same SNP50K ancestry pipeline, with the binary stem-pubescence (macrohair-presence) score modeled on the logit scale (empirical logit of the per-genotype pubescent-plot proportion, no spatial correction), and the maize macrohairless1 (mhl1) locus — an uncloned QTL inside the inv9f mexicana inversion on chromosome 9 — as the lead candidate. Significance is Benjamini–Hochberg FDR ≤ 0.05 (per-scan q-values): the Manhattan line is the −log₁₀P at which BH q = 0.05, and candidate/peak calls report q."
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**What this is.** The stem-pubescence (StPu) sibling of
[`zeal-qtl-recovery-stpi-mlm-snp50k-fdr`](zeal-qtl-recovery-stpi-mlm-snp50k-fdr.html), built from the
same real BZea pipeline: field StPu → **per-genotype empirical logit** → SNP50K teosinte-ancestry
dosage → RTIGER ancestry mosaic → K → JLM + MLM. The population is the same near-isogenic
B73×teosinte BC2S3 lines from **5 teosinte taxa**. StPu is a **binary 0/1** macrohair-presence call
per plot (pubescent stem / not); the per-genotype phenotype is the **proportion of pubescent plots**,
`k/n`, and only **72 of 1,423 genotypes (~5%)** are ever pubescent. Like StPi (and unlike the
continuous DTA/DTS/PH), it is modeled on the logit scale, not with a SpATS spatial BLUE.

::: {.callout-important}
## Phenotype: empirical logit of the pubescent-plot proportion (no spatial correction)
StPu is a binary 0/1 macrohair-presence plot score, not a continuous measurement, so a SpATS
genotype-fixed 2-D P-spline (used for the continuous DTA/DTS/PH traits) is misspecified — the SpATS
pipeline (`zeal_spats_blues.R`) in fact returns **"no fit"** for StPu in both fields, and an earlier
forced fit ran the "BLUEs" **below 0** (min −0.034, impossible for a 0–1 score). The per-genotype
value is a bounded **proportion** `k/n` (k pubescent of n plots, pooled across fields), modeled on
the **logit scale** via the Haldane–Anscombe empirical logit `log((k+0.5)/(n−k+0.5))` — finite at
`k=0` and `k=n`, variance-stabilizing, and respecting the [0,1] bound. No spatial correction. This
is a pragmatic logit for the existing linear (EMMAX/TASSEL) pipeline; a full binomial GLMM would be
the stricter model. With only **72 of 1,423 genotypes ever pubescent (~5%)**, StPu is a
**very low-prevalence** trait — sparser than StPi — so the logit sharpens the ordering among the
pubescent genotypes but cannot add power the sparse phenotype lacks. Built by `zeal_stpu_direct.R`
(the StPi-style `zeal_stpi_direct.R` analog).
:::

::: {.callout-important}
## Genotype at 0.4× skim: why ancestry, not per-SNP
SNP50K is 51,991 teosinte-vs-B73 sites pileup-genotyped at ~0.4× — hard genotypes are
unusable (70% missing, ~0% hom-ALT), so ancestry is inferred from the per-sample allele
**counts**. The **per-SNP genotype (bcftools HWE-posterior) (Panel B)** is λ_GC-deflated (sparse mean-imputation);
the **RTIGER ancestry mosaic (Panel C)** is the usable, block-smoothed genotype (teosinte presence
0.111, matching the established BzeaSeq ~0.101). Only the phenotype changes vs the DTA notebook.
:::

**Trait context.** Stem pubescence is a **macrohair** trait: macrohairs (MHs) are the prominent
epidermal hairs that mark adult vegetative identity and have long been implicated in insect
resistance. The major regulator is **macrohairless1 (`mhl1`)** on the long arm of chromosome 9
(Moose, Lauter & Carlson 2004): recessive *mhl1* alleles eliminate leaf-blade macrohairs, and a
*mexicana*-derived neomorphic allele extends macrohairs from the blade onto the sheath — the "hairy
sheath" pubescence phenotype rare in maize inbreds but common in teosinte. **`mhl1` has never been
cloned** (no causal gene), so it is mapped as a QTL: the three-point map places it between
`glossy15` (`gl15`, proximal) and `umc1120` (distal) on 9L. It falls inside **`inv9f`**, a ~3 Mb
*mexicana*-introgression inversion (v4 chr9:108,615,836–111,785,557 → v5 **chr9:110,501,231–113,605,354**;
Crow *et al.* 2021) that shows a steep *mexicana*-ancestry cline with elevation. In v5 the inversion
sits between the flanking anchors `gl15` (Zm00001eb387280, ~101.0 Mb, proximal — an AP2-like
juvenile→adult regulator upstream of `mhl1`) and `umc1120` (Zm00001eb392660, ~127.2 Mb) / `bk2`
(brittle stalk2, Zm00001eb393070, ~128.9 Mb) distally. The candidate here is therefore the
**`mhl1`/`inv9f` interval itself**, not a single cloned gene. A second, cloned trichome candidate is
**`OCL4`** (Zm00001eb024680, chr1), an HD-ZIP IV transcription factor required for trichome
patterning and anther development in maize (Vernoud *et al.* 2009).

```{r setup}
library(here)
library(data.table)
library(ggplot2)
library(fastman)
library(ggrepel)
library(ggtext)
library(scales)
library(cowplot)
source(here("scripts/teonam_notebook_plots.R")) # shared: plot_manhattan/lollipop, lambda_gc, LOD

# Overlay the published macrohair-density QTL (B73v4 lifted to v5) as translucent red bands.
# Uses the same cumulative-position transformer the Manhattan/lollipop use, built from the SAME
# scan passed to the plot, so bands align exactly with the plotted points.
add_qmh_bands <- function(p, scan) {
  s <- as.data.frame(scan)
  s <- s[is.finite(s$P) & s$P > 0, c("SNP", "CHR", "BP", "P")]
  tr <- get_transformer(s)
  q <- fread(here("data/teonam/stpu_prior_mh_qtl.tsv"))
  q$xmin <- tr(q$chr, q$ci_lo_bp)
  q$xmax <- tr(q$chr, q$ci_hi_bp)
  band <- geom_rect(
    data = q, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
    inherit.aes = FALSE, fill = "#d62728", alpha = 0.10
  )
  p$layers <- c(band, p$layers) # draw beneath the points/line
  p
}
```

```{r load}
mlm <- fread(here("data/zeal/stpu_gwas_mlm_rtiger_mosaic_donor_accession_snp50k.csv")) # Panel C
gt <- fread(here("data/zeal/stpu_gwas_mlm_hwe_post_gt_donor_accession_snp50k.csv")) # Panel B
ols <- fread(here("data/zeal/stpu_gwas_ols_donor_accession_snp50k.csv"))[is.finite(P) & P > 0]
overlap <- here("results/sim/zeal/stpu_candidate_overlap.csv")
lam <- function(p) {
  p <- p[is.finite(p) & p > 0 & p <= 1]
  round(median(qchisq(p, 1, lower.tail = FALSE)) / qchisq(0.5, 1), 2)
}
ALPHA <- 0.05
fdr_lod <- function(p, alpha = ALPHA) {
  p <- p[is.finite(p) & p > 0]
  q <- p.adjust(p, "BH")
  if (!any(q <= alpha)) {
    return(NA_real_)
  }
  -log10(max(p[q <= alpha]))
}
qval <- function(p) p.adjust(p, "BH") # BH q-values aligned to p
```

::: {.callout-note}
## Significance = Benjamini–Hochberg FDR ≤ 0.05
Every Manhattan's dotted line is drawn at the **−log₁₀P where BH q = 0.05** for that scan; a
marker is significant iff its BH q ≤ 0.05. If no marker passes FDR, no line is drawn.
Candidate/peak tables report q.
:::

## Candidate loci: closest marker vs window-max marker

For each StPu trichome/macrohair candidate (B73 v5) — the uncloned macrohair QTL `mhl1` inside the
`inv9f` inversion; the AP2 phase-change factor `gl15` (proximal flank of `mhl1`); the HD-ZIP IV TF
`OCL4`; the WOX3 homeobox genes `wox3a` (ZmWOX3A) / `wox3b`; the SIAMESE-related CDK inhibitor
`smr4` (ZmSMR4); and the SBP-box/SPL family (`tga1`, `tsh4`, `ub2`, `ub3`, and the peak-proximal
`sbp11`/`sbp22`) — the closest marker to the gene body vs the highest
−log₁₀P marker within ±500 kb in the RTIGER-mosaic **MLM (Taxon+K)** scan (Panel C). `pathway`
records the locus role.

```{r candidate-markers}
sc <- copy(mlm)[is.finite(P) & P > 0][, logP := -log10(P)][, q := qval(P)]
gg <- fread(here("data/teonam/stpu_candidate_genes.tsv"))
gg[, chr := as.integer(chr)]
WIN <- 5e5
ctab <- rbindlist(lapply(seq_len(nrow(gg)), function(i) {
  ch <- gg$chr[i]
  s <- gg$start[i]
  e <- gg$end[i]
  m <- sc[CHR == ch]
  if (!nrow(m)) {
    return(NULL)
  }
  d <- pmax(0L, pmax(s - m$BP, m$BP - e))
  kc <- which.min(d)
  w <- m[BP >= s - WIN & BP <= e + WIN]
  kw <- if (nrow(w)) which.max(w$logP) else NA
  data.table(
    gene = gg$symbol[i], pathway = gg$pathway[i], chr = ch, `gene start (v5)` = s,
    `closest Δstart (bp)` = m$BP[kc] - s, `closest -log10P` = round(m$logP[kc], 2),
    `window-max -log10P` = if (nrow(w)) round(w$logP[kw], 2) else NA_real_,
    `window-max q (BH)` = if (nrow(w)) signif(w$q[kw], 2) else NA_real_,
    `FDR<=0.05` = if (nrow(w)) w$q[kw] <= ALPHA else NA
  )
}))
setorder(ctab, chr, `gene start (v5)`)
knitr::kable(ctab,
  format.args = list(big.mark = ","),
  caption = "StPu candidate (B73 v5): closest vs window-max (±500 kb) marker across the mhl1/inv9f interval, RTIGER-mosaic MLM (Taxon+K) scan, with Benjamini–Hochberg q at the window-max marker."
)
```

## OLS → MLM: structure control on a pubescence trait

The fixed-effect OLS (Taxon + marker, no K) vs the MLM (Taxon + K) on the RTIGER mosaic, same
fixed part. OLS is intrinsically inflated (ancestry-block LD); K deflates λ toward 1.

```{r ols-manhattan, fig.width=9, fig.height=4.4}
plot_manhattan(ols,
  sprintf("ZEAL/BZea StPu — OLS (Taxon + marker, no K), RTIGER mosaic (λ_GC = %.2f; dotted = FDR 0.05)", lam(ols$P)),
  overlap_csv = overlap, lod = fdr_lod(ols$P),
  out_png = here("results/sim/zeal/stpu_gwas_ols_manhattan_fdr_snp50k.png")
)
```

```{r ols-vs-mlm}
lam_tab <- data.table(
  model = c("OLS (Taxon + marker, no K)", "MLM (Taxon + K)"),
  markers = c(nrow(ols), nrow(mlm)),
  lambda_GC = c(lam(ols$P), lam(mlm$P)),
  `n FDR<=0.05` = c(sum(qval(ols$P) <= ALPHA), sum(qval(mlm$P) <= ALPHA)),
  max_logP = round(c(max(-log10(ols$P)), max(-log10(mlm$P))), 2)
)
knitr::kable(lam_tab, caption = "OLS vs MLM on the RTIGER mosaic (same Taxon fixed part): K deflates genomic inflation; n markers at BH FDR ≤ 0.05.")
```

```{r ols-vs-mlm-peaks}
ov <- fread(overlap)
peak_at <- function(scan, ch, s) {
  w <- scan[CHR == ch & abs(BP - s) <= 5e5 & is.finite(P) & P > 0]
  if (!nrow(w)) NA_real_ else round(max(-log10(w$P)), 2)
}
cmp <- ov[, .(
  gene = symbol, chr,
  `OLS peak` = mapply(peak_at, list(ols), chr, start),
  `MLM peak` = mapply(peak_at, list(mlm), chr, start)
)]
cmp[, `drop (OLS−MLM)` := round(`OLS peak` - `MLM peak`, 2)]
setorder(cmp, -`OLS peak`)
knitr::kable(cmp, caption = "Local peak -log10P (±500 kb) at the candidate: OLS vs MLM.")
```

## Prior macrohair (MH) QTL overlay (B73 v4 → v5)

Published macrohair-density QTL (Perez-Limón *et al.* 2022, *G3* 12(3):jkab447; a B73×Palomero
Toluqueño highland-maize mapping population) — peak, Bayes support interval, additive
variance explained, and the analyses detecting each — lifted from B73 v4 to v5 and overlaid as red
bands on the JLM and RTIGER-mosaic Manhattans below. None of the StPu FDR peaks fall inside a qMH
interval; the one biologically coherent overlap is **qMH9 ⊃ `mhl1`/`inv9f`** (its Bayes interval
spans most of 9L), and the chr8 StPu peak lies ~2.4 Mb proximal to the tight, high-variance **qMH8**.

```{r prior-mh-qtl}
qmh <- fread(here("data/teonam/stpu_prior_mh_qtl.tsv"))
sc9 <- copy(mlm)[is.finite(P) & P > 0]
mhl_iv <- c(110501231, 113605354) # mhl1/inv9f interval (v5)
best_in <- function(ch, lo, hi) {
  w <- sc9[CHR == ch & BP >= lo & BP <= hi]
  if (nrow(w)) round(max(-log10(w$P)), 2) else NA_real_
}
tabq <- qmh[, .(
  qtl, chr,
  `peak v5 (Mb)` = round(peak_bp / 1e6, 2),
  `Bayes CI v5 (Mb)` = sprintf("%.1f–%.1f", ci_lo_bp / 1e6, ci_hi_bp / 1e6),
  `Var (%)` = var_pct, analysis,
  `max StPu −log10P in CI` = mapply(best_in, chr, ci_lo_bp, ci_hi_bp),
  `contains mhl1/inv9f` = ifelse(chr == 9 & ci_hi_bp >= mhl_iv[1] & ci_lo_bp <= mhl_iv[2], "yes", "no")
)]
knitr::kable(tabq, caption = "Prior macrohair-density QTL from Perez-Limón et al. 2022 (G3 12(3):jkab447), the B73×Palomero Toluqueño highland-maize mapping population (peak + Bayes support interval + additive variance % + detecting analyses; B73v4 lifted to v5). Red bands on the JLM/RTIGER Manhattans below mark these intervals; `max StPu −log10P in CI` is the best RTIGER-mosaic MLM signal inside each.")
```

## JLM QTL genomic distribution (Panel A)

StPu QTL from JLM (TASSEL StepwiseOLS, nested in taxon, enter/exit P=1e-4=LOD4). Model-independent
— the primary analysis, as in Chen 2019.

```{r lollipop-jlm, fig.width=9, fig.height=4}
p_lolli <- add_qmh_bands(plot_lollipop(
  here("results/sim/zeal/tassel/stpu_jlm_native1.txt"), mlm,
  "ZEAL/BZea StPu — JLM QTL genomic distribution (nested in taxon); red bands = prior qMH intervals",
  overlap_csv = overlap, label_genes = c("mhl1", "OCL4", "gl15", "ub2", "sbp22", "tga1", "ub3", "wox3a", "tsh4", "sbp11", "wox3b", "smr4"),
  out_png = here("results/sim/zeal/stpu_jlm_lollipop_snp50k.png")
), mlm)
p_lolli
```

## MLM (Taxon + K), RTIGER ancestry mosaic (Panel C)

```{r manhattan-mlm, fig.width=9, fig.height=4.4}
p_mlm <- add_qmh_bands(plot_manhattan(mlm,
  sprintf("ZEAL/BZea StPu — MLM (Taxon + K), RTIGER mosaic (λ_GC = %.2f; dotted = FDR 0.05; red bands = prior qMH)", lam(mlm$P)),
  overlap_csv = overlap, lod = fdr_lod(mlm$P), out_png = here("results/sim/zeal/stpu_gwas_mlm_rtiger_mosaic_manhattan_fdr_snp50k.png")
), mlm)
p_mlm
cat(sprintf("%d markers pass FDR <= 0.05 (BH); FDR line at -log10P = %s\n", sum(qval(mlm$P) <= ALPHA), ifelse(is.na(fdr_lod(mlm$P)), "NA (none)", sprintf("%.2f", fdr_lod(mlm$P)))))
```

## Composite (A: JLM · B: per-SNP MLM · C: RTIGER-mosaic MLM)

Panel B (per-SNP genotype (bcftools HWE-posterior)) is λ_GC-deflated but catches sharp SNP hits the mosaic
block-terraces away. Panel C (the ancestry mosaic) is the usable, well-calibrated genotype.

```{r composite, fig.width=9, fig.height=13}
p_B <- plot_manhattan(gt,
  sprintf("StPu — MLM (Taxon+K), per-SNP genotype (bcftools HWE-posterior) (λ_GC = %.2f)", lam(gt$P)),
  overlap_csv = overlap, lod = fdr_lod(gt$P)
)
p_C <- add_qmh_bands(plot_manhattan(mlm,
  sprintf("StPu — MLM (Taxon+K), RTIGER ancestry mosaic (λ_GC = %.2f)", lam(mlm$P)),
  overlap_csv = overlap, lod = fdr_lod(mlm$P)
), mlm)
composite <- cowplot::plot_grid(p_lolli, p_B, p_C, ncol = 1, labels = c("A", "B", "C"), label_size = 20, align = "v", axis = "lr")
ggsave(here("results/sim/zeal/stpu_composite_snp50k.png"), composite, width = 9, height = 13, dpi = 200, bg = "white")
composite
```

## Multi-caller StPu scan (control · nNIL · RTIGER · LB-Impute · binhmm)

The same StPu MLM (Taxon+K) scan on each ancestry caller's recovered genotypes, at the real
~0.4× depth. The robust, trait-independent finding is **teosinte-presence recovery**: the
block/bin-aggregated callers (RTIGER 0.111; binhmm 0.101) recover the established BzeaSeq fraction
(~0.10), while per-SNP calling (control) and the per-SNP HMM recalls (nNIL, LB-Impute)
**under-call**. Genomic inflation is phenotype-dependent (reported per caller below). Caller
genotypes are population-level and trait-independent; only the phenotype (StPu) changes.

```{r caller-table}
callers <- c(control = "hwe_post_gt", nNIL = "nnil_mosaic", RTIGER = "rtiger_mosaic", `LB-Impute` = "lbimpute_mosaic", binhmm = "binhmm_mosaic")
gg <- fread(here("data/teonam/stpu_candidate_genes.tsv"))[, chr := as.integer(chr)]
min_cand_q <- function(sc) {
  q <- qval(sc$P)
  vals <- unlist(lapply(seq_len(nrow(gg)), function(i) {
    idx <- which(sc$CHR == gg$chr[i] & abs(sc$BP - gg$start[i]) <= 5e5)
    if (length(idx)) min(q[idx]) else NA_real_
  }))
  if (all(is.na(vals))) NA_real_ else signif(min(vals, na.rm = TRUE), 2)
}
ctab2 <- rbindlist(lapply(names(callers), function(nm) {
  g <- callers[[nm]]
  sc <- fread(here(sprintf("data/zeal/stpu_gwas_mlm_%s_donor_accession_snp50k.csv", g)))
  mf <- here(sprintf("data/zeal/zeal_%s.rds", g))
  pres <- if (file.exists(mf)) {
    st <- readRDS(mf)$state
    round(mean(st %in% c(1L, 2L), na.rm = TRUE), 3)
  } else {
    NA_real_
  }
  sc <- sc[is.finite(P) & P > 0]
  data.table(
    caller = nm, `teosinte presence` = pres, lambda_GC = lam(sc$P),
    `n FDR<=0.05` = sum(qval(sc$P) <= ALPHA),
    `best candidate q (BH)` = min_cand_q(sc)
  )
}))
knitr::kable(ctab2, caption = "StPu MLM (Taxon+K) by ancestry caller at 0.4×, with n markers at BH FDR ≤ 0.05 and the best (min) q across the trichome candidate windows.")
```

```{r caller-composite, fig.width=9, fig.height=20}
panels <- lapply(names(callers), function(nm) {
  sc <- fread(here(sprintf("data/zeal/stpu_gwas_mlm_%s_donor_accession_snp50k.csv", callers[[nm]])))
  plot_manhattan(sc, sprintf("StPu MLM (Taxon+K) — %s (λ_GC = %.2f; FDR 0.05)", nm, lam(sc$P)), overlap_csv = overlap, lod = fdr_lod(sc$P))
})
cc <- cowplot::plot_grid(plotlist = panels, ncol = 1, labels = LETTERS[1:5], label_size = 18, align = "v", axis = "lr")
ggsave(here("results/sim/zeal/stpu_caller_composite_fdr_snp50k.png"), cc, width = 9, height = 20, dpi = 200, bg = "white")
cc
```
