---
title: "ZEAL/BZea DTA — R/qtl (bcsft) on the TeoNAM JLM grid: LOD profile + per-family effects"
subtitle: "Classic R/qtl interval mapping (Haley–Knott) on the BC2S3 RTIGER ancestry mosaic (B73=A / het=H / teosinte=B). Marker set + genetic map = the TeoNAM JLM 0.1-cM set (9,063 markers, TeoNAM cM — NOT computed from ZEAL); the ZEAL mosaic is projected onto those positions. Significance = genome-wide 1000-permutation LOD threshold. Models: no covariate vs + taxon (5-family), and per-family scans with a per-family MAF filter."
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**What this is.** A first R/qtl look at ZEAL/BZea DTA — the classic linkage-mapping route as a
contrast to the EMMAX-MLM / TASSEL-JLM pipeline in the other ZEAL notebooks. A `bcsft` cross
(BC₂S₃), interval mapping by Haley–Knott, and a **1000-permutation** genome-wide threshold. Inputs
produced by `scripts/zeal_rqtl_scan.R` (TRAIT=DTA).

::: {.callout-important}
## Marker set, map, and genotype — read this first
- **No ZEAL genetic map is estimated.** The excess heterozygotes in the mosaic (H ≈ 9% vs
  hom-teosinte ≈ 2%) make a ZEAL `est.map` unreliable, so **every cM here is a TeoNAM distance**,
  not native to ZEAL.
- **Marker set + map = the TeoNAM JLM set** (9,063 markers, 0.1 cM, TeoNAM cM), the *same* grid the
  TeoNAM JLM reproduction uses — so the two analyses are directly comparable.
- ZEAL SNP50K and TeoNAM markers are **disjoint platforms**, so the ZEAL RTIGER-mosaic ancestry is
  **projected** onto the JLM positions (each site takes the nearest ZEAL marker's block state).
- Genotype = ancestry mosaic recoded `0/1/2` → `A/H/B`. This scan has no kinship term; the
  permutation calibrates the threshold to the (structured) null.
:::

```{r setup}
library(here)
library(data.table)
library(qtl)
library(ggplot2)
library(ggrepel)
RQ <- here("results/sim/zeal/rqtl")
stopifnot(dir.exists(RQ)) # run: TRAIT=DTA Rscript scripts/zeal_rqtl_scan.R

# Markers are named S<chr_v5>_<pos_v5> (renamed upstream in zeal_rqtl_scan.R), so a
# peak's v5 position reads straight off its name — no lookup, no v2 confusion.
CCT_BP <- 96175873L # ZmCCT10 (chr10), Chen's DTA10.1
CCT_MB <- CCT_BP / 1e6
bp_of <- function(m) as.integer(sub("^S[0-9]+_", "", m))
ci_str <- function(lbp, hbp) sprintf("%.1f–%.1f", pmin(lbp, hbp) / 1e6, pmax(lbp, hbp) / 1e6)
cct_in <- function(chr, lbp, hbp) {
  fifelse(chr == 10L, fifelse(CCT_BP >= pmin(lbp, hbp) & CCT_BP <= pmax(lbp, hbp), "yes", "no"), "–")
}

load_scan <- function(tag) {
  one <- fread(file.path(RQ, sprintf("zeal_dta_scanone_%s.csv", tag)))
  perms <- readRDS(file.path(RQ, sprintf("zeal_dta_perms_%s.rds", tag)))
  th <- summary(perms, alpha = c(0.05, 0.10))
  list(one = one, thr5 = as.numeric(th[1]), thr10 = as.numeric(th[2]))
}
nc <- load_scan("nocovar")
tx <- load_scan("taxon")
cross <- readRDS(file.path(RQ, "zeal_dta_cross.rds"))
NMARK <- totmar(cross)
```

```{r meta}
g <- pull.geno(cross)
fH <- mean(g == 2, na.rm = TRUE)
fB <- mean(g == 3, na.rm = TRUE)
meta <- data.table(
  metric = c("individuals (DTA)", "markers (TeoNAM JLM grid)", "chromosomes", "H / het", "B / teosinte", "H:B ratio"),
  value = c(
    sum(!is.na(pull.pheno(cross, "DTA"))), NMARK, nchr(cross),
    sprintf("%.3f", fH), sprintf("%.3f", fB), sprintf("%.1f", fH / fB)
  )
)
knitr::kable(meta, caption = "Run composition. Markers = the TeoNAM JLM 0.1-cM grid; genotypes = ZEAL mosaic projected onto it.")
```

## No covariate vs taxon covariate

```{r compare-table}
peakrow <- function(tag) {
  pk <- fread(file.path(RQ, sprintf("zeal_dta_peaks_%s.csv", tag)))
  setnames(pk, old = names(pk), new = sub("^lod\\.", "", names(pk)))
  pk
}
cmp <- data.table(
  model = c("no covariate", "+ taxon (5-family)"),
  `perm 5% LOD` = round(c(nc$thr5, tx$thr5), 2),
  `perm 10% LOD` = round(c(nc$thr10, tx$thr10), 2),
  `max LOD` = round(c(max(nc$one$lod), max(tx$one$lod)), 2),
  `n QTL (FWER 0.05)` = c(nrow(peakrow("nocovar")), nrow(peakrow("taxon")))
)
knitr::kable(cmp, caption = "Adding the 5-family taxon covariate barely shifts the permutation threshold or the QTL count — the pooled chr1 DTA signal is robust to structure correction.")
```

## Genome-wide LOD profile (taxon model, with the no-covariate profile behind it)

Faceted by chromosome. **Blue** = scanone + taxon covariate; **grey** = no covariate. Dashed lines =
the corresponding 1000-permutation 5% thresholds.

```{r lod-profile, fig.width=11, fig.height=4.4}
prof <- rbindlist(list(
  cbind(nc$one[, .(marker, chr, pos, lod)], model = "no covariate"),
  cbind(tx$one[, .(marker, chr, pos, lod)], model = "+ taxon")
))[is.finite(lod)]
prof[, model := factor(model, levels = c("no covariate", "+ taxon"))]
sig <- prof[model == "+ taxon" & lod >= tx$thr5][, .SD[which.max(lod)], by = chr]
pal <- c("no covariate" = "grey65", "+ taxon" = "#2c6fbb")
# ZmCCT10 (chr10 96.18 Mb) mapped onto the TeoNAM cM axis -> mark on the chr10 facet
.n10 <- fread(here("data/teonam/teonam_v5_native.tsv"))[chr_v5 == 10 & is.finite(cm)][order(pos_v5)]
cct10 <- data.table(chr = 10L, cm = approx(.n10$pos_v5, .n10$cm, xout = 96175873)$y)
ytop <- max(prof$lod, na.rm = TRUE)
p <- ggplot(prof, aes(pos, lod, colour = model)) +
  geom_line(linewidth = 0.4) +
  geom_hline(yintercept = nc$thr5, linetype = "dashed", colour = pal[["no covariate"]], linewidth = 0.35) +
  geom_hline(yintercept = tx$thr5, linetype = "dashed", colour = pal[["+ taxon"]], linewidth = 0.35) +
  geom_point(data = sig, colour = "red", size = 1.6) +
  ggrepel::geom_text_repel(
    data = sig, aes(label = sprintf("%s\n%.0f cM · LOD %.1f", marker, pos, lod)),
    size = 2.6, colour = "red", min.segment.length = 0, box.padding = 0.6, show.legend = FALSE
  ) +
  geom_vline(data = cct10, aes(xintercept = cm), inherit.aes = FALSE, linetype = "dotted", colour = "grey30", linewidth = 0.35) +
  geom_point(data = cct10, aes(x = cm, y = ytop * 1.22), inherit.aes = FALSE, shape = 25, fill = "black", colour = "black", size = 2.4) +
  geom_text(data = cct10, aes(x = cm, y = ytop * 1.33, label = "ZmCCT10"), inherit.aes = FALSE, size = 2.7, colour = "grey20") +
  expand_limits(y = ytop * 1.42) +
  facet_grid(~chr, scales = "free_x", space = "free_x", switch = "x") +
  scale_colour_manual(values = pal, name = NULL) +
  labs(
    x = "Chromosome (cM, TeoNAM JLM map)", y = "LOD",
    title = "ZEAL/BZea DTA — R/qtl (bcsft) scanone, Haley–Knott (TeoNAM JLM grid)",
    subtitle = sprintf(
      "%s markers · 1000-perm 5%% threshold: no-cov = %.2f, +taxon = %.2f",
      format(NMARK, big.mark = ","), nc$thr5, tx$thr5
    )
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
    panel.spacing = unit(0.05, "lines"), strip.placement = "outside",
    strip.background = element_blank(), legend.position = "top"
  ) +
  scale_x_continuous(expand = c(0.01, 0))
ggsave(file.path(RQ, "zeal_dta_lod_profile.png"), p, width = 11, height = 4.4, dpi = 200, bg = "white")
p
```

## Significant peaks (taxon model, α = 0.05)

R/qtl's `summary()` reports **at most one peak per chromosome**. That is the table below — position
and the **1.5-LOD confidence interval in v5 Mb**, plus whether **ZmCCT10** (96.2 Mb) falls inside.

```{r peaks}
cipk <- fread(file.path(RQ, "zeal_dta_peaks_ci_taxon.csv"))
knitr::kable(
  cipk[order(chr, -lod), .(
    chr,
    `pos (cM)` = round(pos, 1), `pos (Mb)` = round(bp_of(marker) / 1e6, 1),
    LOD = round(lod, 2), pval, marker,
    `1.5-LOD CI (Mb)` = ci_str(ci_left, ci_right),
    `ZmCCT10 in CI` = cct_in(chr, ci_left, ci_right)
  )],
  caption = "Genome-wide significant QTL, taxon model (1000-perm threshold). One peak per chromosome (R/qtl summary), with the 1.5-LOD interval in v5 Mb and whether ZmCCT10 (chr10, 96.2 Mb) falls inside it."
)
```

### Multi-peak refinement (airmine `refine_peaks`)

A single chromosome can carry **more than one** DTA QTL. To resolve them, `scripts/detect_peaks.R`
(ported from the airmine QTL pipeline) Akima-interpolates each significant chromosome's LOD curve
(10×), runs `pracma::findpeaks` for up to 5 sub-peaks above the permutation threshold, takes each
sub-peak's **1.5-LOD-drop interval**, and resolves overlapping sub-peaks into a maximal set of
**non-overlapping** QTL via an interval graph (igraph independent vertex sets). Positions and the
1.5-LOD-drop confidence intervals are shown in **v5 Mb** (markers are v5-named, so `pos (Mb)` reads
off the name), with the **ZmCCT10-in-CI** flag. Peak splitting from genotyping noise is handled
**before** the scan, not here: the hard RTIGER/projected calls carry het/hom-teosinte miscalls that
punch artificial LOD dives (which would split a QTL and clip its CI). `calc.genoprob` runs the R/qtl
HMM with a realistic `error.prob` so those calls are down-weighted, and the scans use the resulting
genotype **probabilities** (Haley–Knott) / **expected dosage** (per-taxon) rather than raw calls —
so the LOD curve fed to peak detection is already error-corrected.

```{r refined-peaks}
refined <- function(tag) {
  r <- fread(file.path(RQ, sprintf("zeal_dta_peaks_refined_%s.csv", tag)))
  if (nrow(r) == 0) {
    return(r)
  }
  r[order(chr, -lod), .(
    chr,
    `pos (cM)` = round(pos, 1), `pos (Mb)` = round(bp_of(marker) / 1e6, 1),
    LOD = round(lod, 2), `perm 5% thr` = round(thresh, 2), marker,
    `1.5-LOD CI (Mb)` = ci_str(ci_left, ci_right),
    `ZmCCT10 in CI` = cct_in(chr, ci_left, ci_right)
  )]
}
rt <- refined("taxon")
if (nrow(rt) == 0) {
  cat("No refined peaks (no chromosome exceeded the permutation threshold).")
} else {
  knitr::kable(rt, caption = sprintf(
    "Refined multi-peak QTL, taxon model (%d QTL across %d chromosomes) vs %d one-per-chr peaks above. Multiple rows on one chromosome = linked QTL split by the interval-graph deconvolution.",
    nrow(rt), uniqueN(rt$chr), nrow(peakrow("taxon"))
  ))
}
```

## Per-taxon (family-resolved), additive-only with QC

Each taxon's NILs are scanned separately (cf. the Andosol `bcfst_by_donor.R` approach), with QC to
keep low-evidence peaks from surfacing in the small per-family samples (n ≈ 60–380):

- **Phenotype cleaning is upstream:** the DTA BLUEs come from raw plots already cleaned by a
  per-field×taxa **upper Tukey fence** (Q3 + 3·IQR, late tail only) applied before SpATS in
  `zeal_spats_blues.R` — so the very-late plots are removed at source, not re-fenced here.
- **Additive-only model on HMM expected dosage** — LOD from `lm(DTA ~ E[teosinte dosage])`, 1 df,
  where `E[g] = P(het) + 2·P(hom-teo)` comes from the error-aware `calc.genoprob` (so het/hom miscalls
  are down-weighted *before* the scan, not smoothed afterwards); no dominance term, so a near-empty
  homozygous-teosinte class cannot make the fit rank-deficient. QC class-counts below use the hard calls.
- **Marker filters:** (a) **fraction introgressed (HET or ALT) > 0.05** (HET *or* ALT because the
  caller calls het over hom-ALT — BC₂S₃ expectation is ~10.9% hom-teosinte vs ~3.1% het); (b) **≥ 3
  samples in every present genotype class** (empty classes are fine under an additive model — they
  carry no parameter).

```{r per-taxon, fig.width=11, fig.height=4.6}
bt <- fread(file.path(RQ, "zeal_dta_scanone_by_taxon.csv"))[is.finite(lod)]
pal5 <- c(Zd = "#E41A1C", Zh = "#377EB8", Zl = "#4DAF4A", Zv = "#984EA3", Zx = "#FF7F00")
pt <- ggplot(bt, aes(pos, lod, colour = taxon)) +
  geom_line(linewidth = 0.4) +
  geom_hline(
    data = unique(bt[, .(taxon, thr5)]), aes(yintercept = thr5, colour = taxon),
    linetype = "dotted", linewidth = 0.4, show.legend = FALSE
  ) +
  facet_grid(~chr, scales = "free_x", space = "free_x", switch = "x") +
  scale_colour_manual(values = pal5, name = "taxon") +
  labs(
    x = "Chromosome (cM, TeoNAM JLM map)", y = "LOD (additive)",
    title = "ZEAL/BZea DTA — per-taxon additive scan (TeoNAM JLM grid)",
    subtitle = "additive-only; per-family filters: frac introgressed (HET or ALT) > 0.05 & ≥3 per present class"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
    panel.spacing = unit(0.05, "lines"), strip.placement = "outside",
    strip.background = element_blank(), legend.position = "top"
  ) +
  scale_x_continuous(expand = c(0.01, 0))
ggsave(file.path(RQ, "zeal_dta_lod_by_taxon.png"), pt, width = 11, height = 4.6, dpi = 200, bg = "white")
pt
```

```{r per-taxon-peaks}
btpk <- bt[,
  {
    i <- which.max(lod)
    ch <- chr[i]
    L <- lod[i]
    ins <- .SD[chr == ch][lod >= L - 1.5]
    lbp <- min(bp_of(ins$marker))
    hbp <- max(bp_of(ins$marker))
    .(
      chr = ch, `pos (cM)` = round(pos[i], 1), `pos (Mb)` = round(bp_of(marker[i]) / 1e6, 1),
      marker = marker[i], `max LOD` = round(L, 2), `perm 5% thr` = round(thr5[1], 2),
      sig = fifelse(L > thr5[1], "yes", "no"),
      `1.5-LOD CI (Mb)` = sprintf("%.1f–%.1f", lbp / 1e6, hbp / 1e6),
      `ZmCCT10 in CI` = fifelse(ch == 10L, fifelse(CCT_BP >= lbp & CCT_BP <= hbp, "yes", "no"), "–")
    )
  },
  by = taxon
][order(-`max LOD`)]
knitr::kable(btpk, caption = "Genome-wide additive peak per taxon (filtered: frac introgressed > 0.05 & ≥3 per present class), with the per-taxon 1000-permutation 5% threshold. 1.5-LOD interval in v5 Mb and whether ZmCCT10 (chr10) falls inside. 'sig' = the family's peak exceeds its own threshold. One peak per taxon.")
```

### Per-taxon multi-peak refinement

The same `find_subpeaks` deconvolution applied to each family's LOD profile — so a taxon carrying two
QTL on one chromosome (or across the genome) shows all of them, not just its single global maximum.
Thresholded at each family's own per-taxon permutation 5% (`thr5`).

```{r per-taxon-refined}
btr <- fread(file.path(RQ, "zeal_dta_peaks_refined_by_taxon.csv"))
if (nrow(btr) == 0) {
  cat("No per-taxon refined peaks cleared the per-family thresholds.")
} else {
  knitr::kable(
    btr[order(taxon, chr, -lod), .(
      taxon, chr,
      `pos (cM)` = round(pos, 1), `pos (Mb)` = round(bp_of(marker) / 1e6, 1),
      LOD = round(lod, 2), `perm 5% thr` = round(thr5, 2), marker,
      `1.5-LOD CI (Mb)` = ci_str(ci_left, ci_right),
      `ZmCCT10 in CI` = cct_in(chr, ci_left, ci_right)
    )],
    caption = sprintf(
      "Refined per-taxon multi-peak QTL (%d QTL across %d families), 1.5-LOD-drop CIs in v5 Mb.",
      nrow(btr), uniqueN(btr$taxon)
    )
  )
}
```

## Allele effect at the Zd chr10 peak (near ZmCCT10), by taxon

```{r zd-setup}
# derive the effect-plot marker from the current Zd per-family peak (no hardcoding,
# so this cannot drift from the peaks table above)
mk <- bt[taxon == "Zd"][which.max(lod), marker]
zd_lod <- bt[taxon == "Zd"][which.max(lod), round(lod, 2)]
zd_mb <- round(bp_of(mk) / 1e6, 1)
zd_dist <- round(abs(bp_of(mk) - CCT_BP) / 1e6, 1)
# 1.5-LOD interval (v5 Mb) for the Zd chr10 peak + ZmCCT10 containment
.ins <- bt[taxon == "Zd" & chr == 10][lod >= max(lod) - 1.5]
zd_ci_lo <- round(min(bp_of(.ins$marker)) / 1e6, 1)
zd_ci_hi <- round(max(bp_of(.ins$marker)) / 1e6, 1)
zd_cct_in <- CCT_BP >= min(bp_of(.ins$marker)) & CCT_BP <= max(bp_of(.ins$marker))
.g <- pull.geno(cross)[, mk]
.d <- as.numeric(pull.pheno(cross, "DTA"))
.ss <- unique(fread(here("data/zeal/samplesheet_3way.csv"))[, .(pedigree, taxon)], by = "pedigree")
.tx <- .ss[match(as.character(pull.pheno(cross, "id")), pedigree), taxon]
.k <- .tx == "Zd" & is.finite(.d) & !is.na(.g)
zd_ador <- round(coef(lm(.d[.k] ~ c(0, 1, 2)[.g[.k]]))[2], 2)
zd_nbb <- sum(.g[.k] == 3)
zd_bbaa <- round(mean(.d[.k][.g[.k] == 3]) - mean(.d[.k][.g[.k] == 1]), 1)
```

The Zd per-family DTA peak, `` `r mk` `` (chr10 ≈ `r zd_mb` Mb), lies **~`r zd_dist` Mb from ZmCCT10**
(96.2 Mb, Chen's DTA10.1), and its **1.5-LOD interval (`r zd_ci_lo`–`r zd_ci_hi` Mb)
`r if (zd_cct_in) "includes" else "does not include"` ZmCCT10** — so the gene
`r if (zd_cct_in) "sits inside the support interval" else "falls just outside the support interval"`.
In Zd it is a **large-effect, teosinte-late QTL**:
**`r sprintf("%+.2f", zd_ador)` days per teosinte allele, `r sprintf("%+.1f", zd_bbaa)` days B/B−A/A**
(n_BB = `r zd_nbb`), LOD `r zd_lod`. In maize DTA that is a big effect — the scale of the major cloned
flowering genes; the LOD is the *statistical support*, separate from the effect size. Effect by taxon
below: all families shown, but the mean ± SE is drawn **only for genotype classes with ≥ 3 samples** —
undersupported classes (e.g. Zh/Zl A/B, n = 2) show raw points + count, no estimate. Lines beyond the
per-taxa 3·IQR upper fence were dropped upstream. `A/A` = B73, `A/B` = het, `B/B` = homozygous teosinte.

```{r zl-effect, fig.width=10, fig.height=4.2}
gm <- pull.geno(cross)[, mk] # mk from the zd-setup chunk (current Zd peak)
dta <- as.numeric(pull.pheno(cross, "DTA")) # DTA > 100 already dropped upstream
ids <- as.character(pull.pheno(cross, "id"))
ss <- fread(here("data/zeal/samplesheet_3way.csv"))[, .(pedigree, taxon)]
d <- data.table(
  taxon = ss[match(ids, pedigree), taxon],
  geno = factor(c("A/A", "A/B", "B/B")[gm], levels = c("A/A", "A/B", "B/B")), dta
)[!is.na(dta) & !is.na(geno) & taxon %in% c("Zd", "Zh", "Zl", "Zv", "Zx")]
summ <- d[, .(mean = mean(dta), se = sd(dta) / sqrt(.N), n = .N), by = .(taxon, geno)]
shown <- summ[n >= 3] # estimate drawn only for classes with >= 3 samples
ef <- ggplot(d, aes(geno, dta, colour = taxon)) +
  geom_jitter(width = 0.12, alpha = 0.12, size = 0.5) +
  geom_line(data = shown, aes(y = mean, group = taxon), linewidth = 0.7) +
  geom_point(data = shown, aes(y = mean), size = 2.2) +
  geom_errorbar(data = shown, aes(y = mean, ymin = mean - se, ymax = mean + se), width = 0.15) +
  geom_text(data = summ, aes(label = n), y = min(d$dta), vjust = 0, size = 2.5, show.legend = FALSE) +
  facet_wrap(~taxon, nrow = 1) +
  scale_colour_manual(values = pal5, guide = "none") +
  labs(
    x = sprintf("genotype at %s (chr10 ~%s Mb, ~%s Mb from ZmCCT10)", mk, zd_mb, zd_dist), y = "DTA (days)",
    title = sprintf("Allele effect at the Zd chr10 peak — %s, by taxon", mk),
    subtitle = "per-genotype mean DTA ± SE (grey = individual lines; n at base)"
  ) +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())
ggsave(file.path(RQ, sprintf("zeal_dta_%s_effect.png", mk)), ef, width = 10, height = 4.2, dpi = 200, bg = "white")
ef
```

```{r zl-effect-table}
eff <- dcast(summ, taxon ~ geno, value.var = "mean")
ns <- dcast(summ, taxon ~ geno, value.var = "n")
setnames(ns, names(ns)[-1], paste0("n ", names(ns)[-1]), skip_absent = TRUE)
knitr::kable(merge(eff, ns, by = "taxon"),
  digits = 1,
  caption = paste0("Mean DTA by genotype and taxon at ", mk, ", with cell counts.")
)
```

**Read.** *Effect size and statistical support are separate — read effects in days, LOD/n as the
support.* In maize a DTA shift of a few days is a **large** effect, the scale of the major cloned
flowering genes.

Cleaning the DTA BLUEs (per-taxa fence, upstream) drops the earlier chr1 ~254.7 Mb peak — it was over
an uncharacterized universal stress protein and outlier-sensitive — and resolves **two pooled
peaks (+taxon):**

- **chr10 101.4 Mb (`S10_101418767`, LOD 11.28)** — teosinte-**late**, **+3.9 days B/B−A/A
  (+1.08 d/allele), n_BB = 28**. Large effect *and* strongly supported: the point estimate is 5.2 Mb
  distal to **ZmCCT10** (Chen's DTA10.1), but its **1.5-LOD CI (93.2–102.1 Mb) spans ZmCCT10**. This
  is the headline the chr1/USP artifact was masking.
- **chr9 12.4 Mb (`S9_12409400`, LOD 8.09)** — teosinte-**early**, **−1.0 day (−0.95 d/allele),
  n_BB = 31**; no known flowering gene nearby.

Per-family (additive, now with a per-family 1000-perm 5% threshold ~3.5–4.0): four of five families
clear their own threshold. **Zd chr10 96.9 Mb (`S10_96943345`): LOD 8.19 vs thr 3.96 — +6.5 days
B/B−A/A (+2.35 d/allele), n_BB = 8**, a big teosinte-late effect whose point estimate is 0.7 Mb from
**ZmCCT10** and whose **1.5-LOD CI (95.9–100.9 Mb) includes the gene**, echoing the pooled chr10 peak
— the clearest per-family QTL. **Zl chr9 94.3 Mb (`S9_94322165`): LOD 6.83 vs thr 3.52, −6.7 days
(−2.97 d/allele)** — significant but on **only n_BB = 3**, so a large effect on thin support. Zx chr9
4.5 Mb (LOD 4.67 vs 3.73, −2.6 d) and Zv chr3 230.3 Mb (LOD 4.23 vs 3.98, +3.4 d) just clear; Zh does
not (2.48 < 3.92).

So after honest phenotype cleaning the ZEAL DTA scan recovers a **large-effect, well-supported,
teosinte-late QTL in the ZmCCT10 region** — pooled *and* in Zd — consistent with TeoNAM's DTA10.1,
plus a chr9 teosinte-early locus. These are not "modest" effects; they are large, and the chr10 one
is well supported.
