---
title: "ZEAL/BZea Kinki (zigzag culm) GWAS + JLM on the SNP50K panel — 1000-permutation FWER"
subtitle: "Zigzag-culm analog of the StPi notebook: the same SNP50K ancestry pipeline, with the binary kinked-stem score modeled on the logit scale (empirical logit of the per-genotype kinked-plot proportion, no spatial correction; CLY23 only) and a curated maize culm/meristem-architecture candidate set anchored on gn1/knox4 (Gnarley1) and the classical zg1 interval. Significance is a genome-wide FWER from a 1000-permutation null (per scan): the Manhattan line is the -log10P exceeded by only 5% of permutations, and candidate/peak calls report whether they clear it."
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**What this is.** The Kinki (zigzag-culm) sibling of
[`zeal-qtl-recovery-stpi-mlm-snp50k`](zeal-qtl-recovery-stpi-mlm-snp50k.html), built from the
same real BZea pipeline: field Kinki → **per-genotype empirical logit** → SNP50K
teosinte-ancestry dosage → RTIGER ancestry mosaic → K → JLM + MLM. The population is the same
near-isogenic B73×teosinte BC2S3 lines from **5 teosinte taxa**. Kinki is the classic maize
**"zigzag culm"** phenotype (Eyster 1920, *J Hered* 11:349): dwarfed plants with shortened,
thickened, arching ear-region internodes that give the stalk a kinked, zigzag appearance. It was
scored **CLY23 only** as an ordinal 0/1/2 severity, binarized here to "kinked" (score ≥ 1); the
per-genotype phenotype is the proportion of kinked plots, `k/n`.

::: {.callout-important}
## Phenotype: empirical logit of the kinked-plot proportion (no spatial correction)
Kinki is an ordinal 0/1/2 severity score, not a continuous measurement, so a SpATS
genotype-fixed 2-D P-spline (used for the continuous DTA/DTS/PH/EH traits) is misspecified. The
score is binarized to kinked presence (≥ 1) and the per-genotype value is a bounded **proportion**
`k/n` (k kinked of n plots), modeled on the **logit scale** via the Haldane–Anscombe empirical
logit `log((k+0.5)/(n−k+0.5))` — finite at `k=0` and `k=n`, variance-stabilizing, and respecting
the [0,1] bound. Kinki was scored in **CLY23 only** (a single field, absent in CLY25), so there is
no cross-field union and no spatial surface to correct — consistent with the binary StPi/StPu
treatment. This is a pragmatic logit for the existing linear (EMMAX/TASSEL) pipeline; a full
binomial GLMM would be the stricter model.
:::

::: {.callout-important}
## Genotype at 0.4× skim: why ancestry, not per-SNP
SNP50K is 51,991 teosinte-vs-B73 sites pileup-genotyped at ~0.4× — hard genotypes are
unusable (70% missing, ~0% hom-ALT), so ancestry is inferred from the per-sample allele
**counts**. The **per-SNP genotype (bcftools HWE-posterior) (Panel B)** is λ_GC-deflated (sparse mean-imputation);
the **RTIGER ancestry mosaic (Panel C)** is the usable, block-smoothed genotype (teosinte presence
0.111, matching the established BzeaSeq ~0.101). Only the phenotype changes vs the StPi notebook.
:::

**Trait context.** The zigzag culm is a **culm / shoot-meristem-architecture** defect: shortened,
mis-oriented internodes and kinked stalks. The most direct a-priori candidate is **`gn1`/`knox4`
(Gnarley1)** — a dominant KNOX-homeobox mutant that produces curving/twisting culms with short
internodes, and the intended comparison gene for this phenotype. The candidate set assembles the
loci most likely to underlie a kinked-culm phenotype: the **KNOX / KNOX-regulatory** axis (`gn1`,
`lg3`, and the KNOX-repressor `rs2`, whose recessive loss gives compressed/twisted culms via ectopic
KNOX), **GA / dwarf internode** genes (`an1`, `d9`, `ct2`), **CLV–WUS meristem-size** receptors
(`fea2`, `td1`), **auxin / inflorescence-meristem** genes (`bif2`, `ba1`, `spi1`, `ra1`),
cell-division-plane control (`tan1`), the cytokinin receptor `hsf1`, and pigment/linkage **anchors**
for the classical **`zg1`** interval on chr 6L (`pl1`, `sm1`) plus the chr3 anchor `a1`. Because
teosinte ancestry is **partly collinear with K**, the mixed model absorbs ancestry-correlated
variance, and the model-independent **JLM** (nested-in-family stepwise) is the primary QTL tool.

```{r setup}
library(here)
library(data.table)
library(ggplot2)
library(fastman)
library(ggrepel)
library(ggtext)
library(scales)
library(cowplot)
source(here("scripts/teonam_notebook_plots.R")) # shared: plot_manhattan/lollipop, lambda_gc, LOD
```

```{r load}
mlm <- fread(here("data/zeal/kinki_gwas_mlm_rtiger_mosaic_donor_accession_snp50k.csv")) # Panel C
gt <- fread(here("data/zeal/kinki_gwas_mlm_hwe_post_gt_donor_accession_snp50k.csv")) # Panel B
ols <- fread(here("data/zeal/kinki_gwas_ols_donor_accession_snp50k.csv"))[is.finite(P) & P > 0]
overlap <- here("results/sim/zeal/kinki_candidate_overlap.csv")
lam <- function(p) {
  p <- p[is.finite(p) & p > 0 & p <= 1]
  round(median(qchisq(p, 1, lower.tail = FALSE)) / qchisq(0.5, 1), 2)
}
# Significance = genome-wide FWER from a 1000-permutation null (scripts/zeal_gwas_perm.R),
# stored per (trait, model, geno) in data/zeal/gwas_perm_thresholds.csv.
TR <- "KINKI"
thr <- function(model, geno, a = 0.05) gwas_thr(TR, model, geno, a)
thr_ols <- thr("ols", "rtiger_mosaic")
thr_mlm <- thr("mlm", "rtiger_mosaic")
thr_gt <- thr("mlm", "hwe_post_gt")
nsig <- function(p, t) sum(-log10(p) > t, na.rm = TRUE)
```

::: {.callout-note}
## Significance = 1000-permutation genome-wide FWER (5%)
Every Manhattan's dotted line is the **−log₁₀P exceeded by only 5% of 1000 permutations** of
that scan (a genome-wide max-−log₁₀P null; `scripts/zeal_gwas_perm.R`), stored per
(trait, model, geno) in `data/zeal/gwas_perm_thresholds.csv`. A marker is significant iff its
−log₁₀P clears that line. The line's height differs per panel with the null. Candidate/peak
tables report whether the marker clears the 5% FWER line.
:::

## Candidate gene registry

Curated evidence for each **candidate** (tier, causal direction, evidence type, phenotype, and a DOI link) from the tracked registry (`reference/candidate_genes.csv` × `candidate_evidence.csv`). These are the genes starred on the Manhattans and labelled on the lollipop; loci without B73 v5 coordinates still appear here but are not plotted.

```{r candidate-registry}
candidate_table(TR)
```

## Candidate genes: closest marker vs window-max marker

For each maize culm/meristem-architecture candidate (B73 v5), the closest marker to the gene body
vs the highest −log₁₀P marker within ±500 kb in the RTIGER-mosaic **MLM (Family+K)** scan (Panel C).
`pathway` records the biological role each locus acts through (KNOX / GA-dwarf / meristem / anchor).

```{r candidate-markers}
sc <- copy(mlm)[is.finite(P) & P > 0][, logP := -log10(P)]
gg <- fread(here("data/teonam/kinki_candidate_genes.tsv"))
gg[, chr := as.integer(chr)]
WIN <- 5e5
ctab <- rbindlist(lapply(seq_len(nrow(gg)), function(i) {
  ch <- gg$chr[i]
  s <- gg$start[i]
  e <- gg$end[i]
  m <- sc[CHR == ch]
  if (!nrow(m)) {
    return(NULL)
  }
  d <- pmax(0L, pmax(s - m$BP, m$BP - e))
  kc <- which.min(d)
  w <- m[BP >= s - WIN & BP <= e + WIN]
  kw <- if (nrow(w)) which.max(w$logP) else NA
  data.table(
    gene = gg$symbol[i], pathway = gg$pathway[i], chr = ch, `gene start (v5)` = s,
    `closest Δstart (bp)` = m$BP[kc] - s, `closest -log10P` = round(m$logP[kc], 2),
    `window-max -log10P` = if (nrow(w)) round(w$logP[kw], 2) else NA_real_,
    `clears perm 5%` = if (nrow(w)) (w$logP[kw] >= thr_mlm) else NA
  )
}))
setorder(ctab, chr, `gene start (v5)`)
knitr::kable(ctab,
  format.args = list(big.mark = ","),
  caption = sprintf("Kinki candidates (B73 v5; culm/meristem-architecture loci): closest vs window-max (±500 kb) marker, RTIGER-mosaic MLM (Family+K) scan; 'clears perm 5%%' = window-max exceeds the 1000-permutation FWER line (-log10P = %.2f).", thr_mlm)
)
```

## OLS → MLM: structure control

Here **Family** is the **82 donor-accession** fixed factor (parent-plant-level structure correction) that the MLM and per-marker OLS scans were actually computed with — a stronger correction than the 5 donor **taxa** the JLM nests on. It likely over-corrects population structure and costs some power; a taxon-level (5) MLM/OLS rerun is a pending comparison (see `TERMINOLOGY.md`).

The fixed-effect OLS (Family + marker, no K) vs the MLM (Family + K) on the RTIGER mosaic, same
fixed part. OLS is intrinsically inflated (ancestry-block LD); K deflates λ toward 1.

```{r ols-manhattan, fig.width=9, fig.height=4.4}
plot_manhattan(ols,
  sprintf("ZEAL/BZea Kinki — OLS (Family + marker, no K), RTIGER mosaic (λ_GC = %.2f; dotted = perm FWER 5%%)", lam(ols$P)),
  overlap_csv = overlap, lod = thr_ols,
  out_png = here("results/sim/zeal/kinki_gwas_ols_manhattan_fdr_snp50k.png")
)
```

```{r ols-vs-mlm}
lam_tab <- data.table(
  model = c("OLS (Family + marker, no K)", "MLM (Family + K)"),
  markers = c(nrow(ols), nrow(mlm)),
  lambda_GC = c(lam(ols$P), lam(mlm$P)),
  `n > perm 5%` = c(nsig(ols$P, thr_ols), nsig(mlm$P, thr_mlm)),
  max_logP = round(c(max(-log10(ols$P)), max(-log10(mlm$P))), 2)
)
knitr::kable(lam_tab, caption = "OLS vs MLM on the RTIGER mosaic (same Family fixed part): K deflates genomic inflation; n markers above each scan's 1000-permutation 5% FWER line.")
```

```{r ols-vs-mlm-peaks}
ov <- fread(overlap)
peak_at <- function(scan, ch, s) {
  w <- scan[CHR == ch & abs(BP - s) <= 5e5 & is.finite(P) & P > 0]
  if (!nrow(w)) NA_real_ else round(max(-log10(w$P)), 2)
}
cmp <- ov[, .(
  gene = symbol, chr,
  `OLS peak` = mapply(peak_at, list(ols), chr, start),
  `MLM peak` = mapply(peak_at, list(mlm), chr, start)
)]
cmp[, `drop (OLS−MLM)` := round(`OLS peak` - `MLM peak`, 2)]
setorder(cmp, -`OLS peak`)
knitr::kable(cmp, caption = "Local peak -log10P (±500 kb) per candidate: OLS vs MLM.")
```

## JLM QTL genomic distribution (Panel A)

Kinki QTL from JLM (TASSEL StepwiseOLS, nested in taxon, entry by a 1000-permutation empirical α = 0.05 (Chen-style genome-wide FWER; per-trait)). Model-independent
— the primary analysis, as in Chen 2019.

```{r lollipop-jlm, fig.width=9, fig.height=4}
p_lolli <- plot_lollipop(
  here("results/sim/zeal/tassel/kinki_jlm_native1.txt"), mlm,
  "ZEAL/BZea Kinki — JLM QTL genomic distribution (nested in taxon)",
  overlap_csv = overlap,
  out_png = here("results/sim/zeal/kinki_jlm_lollipop_snp50k.png")
)
p_lolli
```

## MLM (Family + K), RTIGER ancestry mosaic (Panel C)

```{r manhattan-mlm, fig.width=9, fig.height=4.4}
p_mlm <- plot_manhattan(mlm,
  sprintf("ZEAL/BZea Kinki — MLM (Family + K), RTIGER ancestry mosaic (λ_GC = %.2f; dotted = perm FWER 5%%)", lam(mlm$P)),
  overlap_csv = overlap, lod = thr_mlm, out_png = here("results/sim/zeal/kinki_gwas_mlm_rtiger_mosaic_manhattan_fdr_snp50k.png")
)
p_mlm
cat(sprintf("%d markers clear the 5%% FWER line (-log10P = %.2f)\n", nsig(mlm$P, thr_mlm), thr_mlm))
```

## Composite (A: JLM · B: per-SNP MLM · C: RTIGER-mosaic MLM)

Panel B (per-SNP genotype (bcftools HWE-posterior)) is λ_GC-deflated but catches sharp SNP hits the mosaic
block-terraces away. Panel C (the ancestry mosaic) is the usable, well-calibrated genotype.

```{r composite, fig.width=9, fig.height=13}
p_B <- plot_manhattan(gt,
  sprintf("Kinki — MLM (Family+K), per-SNP genotype (bcftools HWE-posterior) (λ_GC = %.2f)", lam(gt$P)),
  overlap_csv = overlap, lod = thr_gt
)
p_C <- plot_manhattan(mlm,
  sprintf("Kinki — MLM (Family+K), RTIGER ancestry mosaic (λ_GC = %.2f)", lam(mlm$P)),
  overlap_csv = overlap, lod = thr_mlm
)
composite <- cowplot::plot_grid(p_lolli, p_B, p_C, ncol = 1, labels = c("A", "B", "C"), label_size = 20, align = "v", axis = "lr")
ggsave(here("results/sim/zeal/kinki_composite_snp50k.png"), composite, width = 9, height = 13, dpi = 200, bg = "white")
composite
```

## Multi-caller Kinki scan (control · nNIL · RTIGER · LB-Impute · binhmm)

The same Kinki MLM (Family+K) scan on each ancestry caller's recovered genotypes, at the real
~0.4× depth. The robust, trait-independent finding is **teosinte-presence recovery**: the
block/bin-aggregated callers (RTIGER 0.111; binhmm 0.101) recover the established BzeaSeq fraction
(~0.10), while per-SNP calling (control) and the per-SNP HMM recalls (nNIL, LB-Impute)
**under-call**. Genomic inflation is phenotype-dependent (reported per caller below). Caller
genotypes are population-level and trait-independent; only the phenotype (Kinki) changes.

```{r caller-table}
callers <- c(control = "hwe_post_gt", nNIL = "nnil_mosaic", RTIGER = "rtiger_mosaic", `LB-Impute` = "lbimpute_mosaic", binhmm = "binhmm_mosaic")
gg <- fread(here("data/teonam/kinki_candidate_genes.tsv"))[, chr := as.integer(chr)]
max_cand_logp <- function(sc) {
  lp <- -log10(sc$P)
  vals <- unlist(lapply(seq_len(nrow(gg)), function(i) {
    idx <- which(sc$CHR == gg$chr[i] & abs(sc$BP - gg$start[i]) <= 5e5)
    if (length(idx)) max(lp[idx], na.rm = TRUE) else NA_real_
  }))
  if (all(is.na(vals))) NA_real_ else round(max(vals, na.rm = TRUE), 2)
}
ctab2 <- rbindlist(lapply(names(callers), function(nm) {
  g <- callers[[nm]]
  sc <- fread(here(sprintf("data/zeal/kinki_gwas_mlm_%s_donor_accession_snp50k.csv", g)))
  mf <- here(sprintf("data/zeal/zeal_%s.rds", g))
  pres <- if (file.exists(mf)) {
    st <- readRDS(mf)$state
    round(mean(st %in% c(1L, 2L), na.rm = TRUE), 3)
  } else {
    NA_real_
  }
  sc <- sc[is.finite(P) & P > 0]
  data.table(
    caller = nm, `teosinte presence` = pres, lambda_GC = lam(sc$P),
    `n > perm 5%` = nsig(sc$P, gwas_thr(TR, "mlm", g)),
    `best candidate -log10P` = max_cand_logp(sc)
  )
}))
knitr::kable(ctab2, caption = "Kinki MLM (Family+K) by ancestry caller at 0.4×, with n markers above the caller's own 1000-permutation 5% FWER line and the best (max) -log10P across the culm/meristem candidate windows.")
```

```{r caller-composite, fig.width=9, fig.height=20}
panels <- lapply(names(callers), function(nm) {
  sc <- fread(here(sprintf("data/zeal/kinki_gwas_mlm_%s_donor_accession_snp50k.csv", callers[[nm]])))
  plot_manhattan(sc, sprintf("Kinki MLM (Family+K) — %s (λ_GC = %.2f; perm FWER 5%%)", nm, lam(sc$P)), overlap_csv = overlap, lod = gwas_thr(TR, "mlm", callers[[nm]]))
})
cc <- cowplot::plot_grid(plotlist = panels, ncol = 1, labels = LETTERS[1:5], label_size = 18, align = "v", axis = "lr")
ggsave(here("results/sim/zeal/kinki_caller_composite_fdr_snp50k.png"), cc, width = 9, height = 20, dpi = 200, bg = "white")
cc
```
