---
title: "ZEAL/BZea Leaf-Greenness (SPAD) GWAS + JLM on the SNP50K panel — 1000-permutation FWER"
subtitle: "The leaf-greenness analog of the PH/EH notebooks: same SNP50K ancestry pipeline, phenotype = per-field SpATS SPAD BLUE (leaf greenness index; scored in CLY25 only), candidate set = maize chlorophyll biosynthesis / catabolism-stay-green / light-harvesting / GLK–NAC regulator genes. Significance is a genome-wide FWER from a 1000-permutation null (per scan): the Manhattan line is the −log₁₀P exceeded by only 5% of permutations, and candidate/peak calls report whether they clear it."
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**What this is.** The leaf-greenness (SPAD) sibling of
[`zeal-qtl-recovery-ph-mlm-snp50k`](zeal-qtl-recovery-ph-mlm-snp50k.html), built from the
same real BZea pipeline end-to-end: field SPAD → SpATS spatial BLUE → SNP50K teosinte-ancestry
dosage → RTIGER ancestry mosaic → K → JLM + MLM. **SPAD** is the leaf greenness index read by a
SPAD meter on one leaf above the primary ear node — a proxy for leaf chlorophyll content. It was
scored in **CLY25 only**, so the BLUE is a **single-field** SpATS mean (no cross-field union). Only
the phenotype changes vs the PH/EH notebooks.

::: {.callout-note}
## Phenotype provenance (single-SPAD, spatially corrected) — exactly what was used
- **Column:** `SPAD` in the CLY25 `B5_BZea_eval` fieldbook sheet (`CLY25-Fieldbook.xlsx`) — a
  SPAD-meter reading on one leaf above the primary ear node, dated **Jul 2025** (`SPAD_date`).
- **BLUE:** per contiguous map grid, a **SpATS** 2-D P-spline `PSANOVA(Row, Range)` with genotype
  **fixed** + row/range random effects and |t|>4 outlier removal (`scripts/zeal_spats_blues.R`,
  `TRAIT=SPAD PHENO=blue`); the genotype value is the mean of its grid BLUEs. GWAS on the
  RTIGER-mosaic MLM with `FAMILY_COL=donor_accession`.
- **CLY25 only:** CLY23's `SPAD1` is a placeholder (all `1.0`, no dates) and its real reading
  `SPAD2` (dated **Aug 2023**) is a different measurement, **not** merged into this single-SPAD
  BLUE. The two measurement dates are analysed separately in the sibling notebooks
  [`spad20das`](zeal-qtl-recovery-spad20das-mlm-snp50k.html) and
  [`spad36das`](zeal-qtl-recovery-spad36das-mlm-snp50k.html), which use the zealbrowser
  phenotype values **as-is** (no additional spatial correction).
:::

::: {.callout-important}
## Genotype at 0.4× skim: why ancestry, not per-SNP
SNP50K is 51,991 teosinte-vs-B73 sites pileup-genotyped at ~0.4× — hard genotypes are
unusable (70% missing, ~0% hom-ALT), so ancestry is inferred from the per-sample allele
**counts**. The **per-SNP genotype (bcftools HWE-posterior) (Panel B)** is therefore λ_GC-deflated (sparse
mean-imputation); the **RTIGER ancestry mosaic (Panel C)** is the usable, block-smoothed
genotype (teosinte presence 0.111, matching the established BzeaSeq ~0.101). Identical in mechanism
to the DTA/PH notebooks — only the phenotype changes.
:::

**Trait context.** SPAD is a **physiological / leaf-chlorophyll** trait, not an architecture or
flowering trait. The candidate set spans the maize chlorophyll pathway and its regulators: **chlorophyll
biosynthesis** (`chlh1`, `oy1` = *oil yellow1*/Mg-chelatase I — the classic maize SPAD gene, `gtr3`,
`urod`, `por1`, `por2`), **chlorophyll catabolism / stay-green** (`pph`, `nye1`/`nye2`/`sgrl1` =
SGR/STAY-GREEN, `lls1` = *lethal leaf spot1*/PaO, `nyc1` = chlorophyll b reductase), the
**light-harvesting** antenna (`lhcb10`, `cab1`, `cab2`), and greenness **regulators**
(`nactf132`/ZmNAC132, and the GOLDEN2-LIKE TFs `g2`, `glk1`). As in the other traits, teosinte
ancestry is structured and partly collinear with K, so the model-independent **JLM**
(nested-in-family stepwise) is the primary QTL tool and the MLM absorbs ancestry-correlated variance.

```{r setup}
library(here)
library(data.table)
library(ggplot2)
library(fastman)
library(ggrepel)
library(ggtext)
library(scales)
library(cowplot)
source(here("scripts/teonam_notebook_plots.R")) # shared: plot_manhattan/lollipop, lambda_gc, LOD
```

```{r load}
mlm <- fread(here("data/zeal/spad_gwas_mlm_rtiger_mosaic_donor_accession_snp50k.csv")) # Panel C
gt <- fread(here("data/zeal/spad_gwas_mlm_hwe_post_gt_donor_accession_snp50k.csv")) # Panel B
ols <- fread(here("data/zeal/spad_gwas_ols_donor_accession_snp50k.csv"))[is.finite(P) & P > 0]
overlap <- here("results/sim/zeal/spad_candidate_overlap.csv")
lam <- function(p) {
  p <- p[is.finite(p) & p > 0 & p <= 1]
  round(median(qchisq(p, 1, lower.tail = FALSE)) / qchisq(0.5, 1), 2)
}
# Significance = genome-wide FWER from a 1000-permutation null (scripts/zeal_gwas_perm.R),
# stored per (trait, model, geno) in data/zeal/gwas_perm_thresholds.csv.
TR <- "SPAD"
thr <- function(model, geno, a = 0.05) gwas_thr(TR, model, geno, a)
thr_ols <- thr("ols", "rtiger_mosaic")
thr_mlm <- thr("mlm", "rtiger_mosaic")
thr_gt <- thr("mlm", "hwe_post_gt")
nsig <- function(p, t) sum(-log10(p) > t, na.rm = TRUE)
```

::: {.callout-note}
## Significance = 1000-permutation genome-wide FWER (5%)
Every Manhattan's dotted line is the **−log₁₀P exceeded by only 5% of 1000 permutations** of
that scan (a genome-wide max-−log₁₀P null; `scripts/zeal_gwas_perm.R`), stored per
(trait, model, geno) in `data/zeal/gwas_perm_thresholds.csv`. A marker is significant iff its
−log₁₀P clears that line. The line's height differs per panel with the null. Candidate/peak
tables report whether the marker clears the 5% FWER line.
:::

## Candidate gene registry

Curated evidence for each **candidate** (tier, causal direction, evidence type, phenotype, and a DOI link) from the tracked registry (`reference/candidate_genes.csv` × `candidate_evidence.csv`). These are the genes starred on the Manhattans and labelled on the lollipop; loci without B73 v5 coordinates still appear here but are not plotted.

```{r candidate-registry}
candidate_table(TR)
```

## Candidate genes: closest marker vs window-max marker

For each maize chlorophyll/greenness candidate (B73 v5), the closest marker to the gene body vs the
highest −log₁₀P marker within ±500 kb in the RTIGER-mosaic **MLM (Family+K)** scan (Panel C).
`pathway` records the chlorophyll-pathway role (biosynthesis / catabolism-stay-green /
light-harvesting / regulator).

```{r candidate-markers}
sc <- copy(mlm)[is.finite(P) & P > 0][, logP := -log10(P)]
gg <- fread(here("data/teonam/spad_candidate_genes.tsv"))
gg[, chr := as.integer(chr)]
WIN <- 5e5
ctab <- rbindlist(lapply(seq_len(nrow(gg)), function(i) {
  ch <- gg$chr[i]
  s <- gg$start[i]
  e <- gg$end[i]
  m <- sc[CHR == ch]
  if (!nrow(m)) {
    return(NULL)
  }
  d <- pmax(0L, pmax(s - m$BP, m$BP - e))
  kc <- which.min(d)
  w <- m[BP >= s - WIN & BP <= e + WIN]
  kw <- if (nrow(w)) which.max(w$logP) else NA
  data.table(
    gene = gg$symbol[i], pathway = gg$pathway[i], chr = ch, `gene start (v5)` = s,
    `closest Δstart (bp)` = m$BP[kc] - s, `closest -log10P` = round(m$logP[kc], 2),
    `window-max -log10P` = if (nrow(w)) round(w$logP[kw], 2) else NA_real_,
    `clears perm 5%` = if (nrow(w)) (w$logP[kw] >= thr_mlm) else NA
  )
}))
setorder(ctab, chr, `gene start (v5)`)
knitr::kable(ctab,
  format.args = list(big.mark = ","),
  caption = sprintf("SPAD candidates (B73 v5; maize chlorophyll biosynthesis / catabolism / light-harvesting / regulators): closest vs window-max (±500 kb) marker, RTIGER-mosaic MLM (Family+K) scan; 'clears perm 5%%' = window-max exceeds the 1000-permutation FWER line (-log10P = %.2f).", thr_mlm)
)
```

## OLS → MLM: structure control

Here **Family** is the **82 donor-accession** fixed factor (parent-plant-level structure correction) that the MLM and per-marker OLS scans were actually computed with — a stronger correction than the 5 donor **taxa** the JLM nests on. It likely over-corrects population structure and costs some power; a taxon-level (5) MLM/OLS rerun is a pending comparison (see `TERMINOLOGY.md`).

The fixed-effect OLS (Family + marker, no K) vs the MLM (Family + K) on the RTIGER mosaic, same
fixed part. OLS is intrinsically inflated (ancestry-block LD); K deflates λ toward 1.

```{r ols-manhattan, fig.width=9, fig.height=4.4}
plot_manhattan(ols,
  sprintf("ZEAL/BZea SPAD — OLS (Family + marker, no K), RTIGER mosaic (λ_GC = %.2f; dotted = perm FWER 5%%)", lam(ols$P)),
  overlap_csv = overlap, lod = thr_ols,
  out_png = here("results/sim/zeal/spad_gwas_ols_manhattan_fdr_snp50k.png")
)
```

```{r ols-vs-mlm}
lam_tab <- data.table(
  model = c("OLS (Family + marker, no K)", "MLM (Family + K)"),
  markers = c(nrow(ols), nrow(mlm)),
  lambda_GC = c(lam(ols$P), lam(mlm$P)),
  `n > perm 5%` = c(nsig(ols$P, thr_ols), nsig(mlm$P, thr_mlm)),
  max_logP = round(c(max(-log10(ols$P)), max(-log10(mlm$P))), 2)
)
knitr::kable(lam_tab, caption = "OLS vs MLM on the RTIGER mosaic (same Family fixed part): K deflates genomic inflation; n markers above each scan's 1000-permutation 5% FWER line.")
```

```{r ols-vs-mlm-peaks}
ov <- fread(overlap)
peak_at <- function(scan, ch, s) {
  w <- scan[CHR == ch & abs(BP - s) <= 5e5 & is.finite(P) & P > 0]
  if (!nrow(w)) NA_real_ else round(max(-log10(w$P)), 2)
}
cmp <- ov[, .(
  gene = symbol, chr,
  `OLS peak` = mapply(peak_at, list(ols), chr, start),
  `MLM peak` = mapply(peak_at, list(mlm), chr, start)
)]
cmp[, `drop (OLS−MLM)` := round(`OLS peak` - `MLM peak`, 2)]
setorder(cmp, -`OLS peak`)
knitr::kable(cmp, caption = "Local peak -log10P (±500 kb) per candidate: OLS vs MLM.")
```

## JLM QTL genomic distribution (Panel A)

SPAD QTL from JLM (TASSEL StepwiseOLS, nested in taxon, entry by a 1000-permutation empirical α = 0.05 (Chen-style genome-wide FWER; per-trait)). Model-independent
— the primary analysis, as in Chen 2019.

```{r lollipop-jlm, fig.width=9, fig.height=4}
p_lolli <- plot_lollipop(
  here("results/sim/zeal/tassel/spad_jlm_native1.txt"), mlm,
  "ZEAL/BZea SPAD — JLM QTL genomic distribution (nested in taxon)",
  overlap_csv = overlap,
  out_png = here("results/sim/zeal/spad_jlm_lollipop_snp50k.png")
)
p_lolli
```

## MLM (Family + K), RTIGER ancestry mosaic (Panel C)

```{r manhattan-mlm, fig.width=9, fig.height=4.4}
p_mlm <- plot_manhattan(mlm,
  sprintf("ZEAL/BZea SPAD — MLM (Family + K), RTIGER ancestry mosaic (λ_GC = %.2f; dotted = perm FWER 5%%)", lam(mlm$P)),
  overlap_csv = overlap, lod = thr_mlm, out_png = here("results/sim/zeal/spad_gwas_mlm_rtiger_mosaic_manhattan_fdr_snp50k.png")
)
p_mlm
cat(sprintf("%d markers clear the 5%% FWER line (-log10P = %.2f)\n", nsig(mlm$P, thr_mlm), thr_mlm))
```

## Composite (A: JLM · B: per-SNP MLM · C: RTIGER-mosaic MLM)

Panel B (per-SNP genotype (bcftools HWE-posterior)) is λ_GC-deflated but catches sharp SNP hits the mosaic
block-terraces away. Panel C (the ancestry mosaic) is the usable, well-calibrated genotype.

```{r composite, fig.width=9, fig.height=13}
p_B <- plot_manhattan(gt,
  sprintf("SPAD — MLM (Family+K), per-SNP genotype (bcftools HWE-posterior) (λ_GC = %.2f)", lam(gt$P)),
  overlap_csv = overlap, lod = thr_gt
)
p_C <- plot_manhattan(mlm,
  sprintf("SPAD — MLM (Family+K), RTIGER ancestry mosaic (λ_GC = %.2f)", lam(mlm$P)),
  overlap_csv = overlap, lod = thr_mlm
)
composite <- cowplot::plot_grid(p_lolli, p_B, p_C, ncol = 1, labels = c("A", "B", "C"), label_size = 20, align = "v", axis = "lr")
ggsave(here("results/sim/zeal/spad_composite_snp50k.png"), composite, width = 9, height = 13, dpi = 200, bg = "white")
composite
```

## Multi-caller SPAD scan (control · nNIL · RTIGER · LB-Impute · binhmm)

The same SPAD MLM (Family+K) scan on each ancestry caller's recovered genotypes, at the real
~0.4× depth. The robust, trait-independent finding is **teosinte-presence recovery**: the
block/bin-aggregated callers (RTIGER 0.111; binhmm 0.101) recover the established BzeaSeq fraction
(~0.10), while per-SNP calling (control) and the per-SNP HMM recalls (nNIL, LB-Impute)
**under-call**. Genomic inflation is phenotype-dependent (reported per caller below). Caller
genotypes are population-level and trait-independent; only the phenotype (SPAD) changes.

```{r caller-table}
callers <- c(control = "hwe_post_gt", nNIL = "nnil_mosaic", RTIGER = "rtiger_mosaic", `LB-Impute` = "lbimpute_mosaic", binhmm = "binhmm_mosaic")
gg <- fread(here("data/teonam/spad_candidate_genes.tsv"))[, chr := as.integer(chr)]
max_cand_logp <- function(sc) {
  lp <- -log10(sc$P)
  vals <- unlist(lapply(seq_len(nrow(gg)), function(i) {
    idx <- which(sc$CHR == gg$chr[i] & abs(sc$BP - gg$start[i]) <= 5e5)
    if (length(idx)) max(lp[idx], na.rm = TRUE) else NA_real_
  }))
  if (all(is.na(vals))) NA_real_ else round(max(vals, na.rm = TRUE), 2)
}
ctab2 <- rbindlist(lapply(names(callers), function(nm) {
  g <- callers[[nm]]
  sc <- fread(here(sprintf("data/zeal/spad_gwas_mlm_%s_donor_accession_snp50k.csv", g)))
  mf <- here(sprintf("data/zeal/zeal_%s.rds", g))
  pres <- if (file.exists(mf)) {
    st <- readRDS(mf)$state
    round(mean(st %in% c(1L, 2L), na.rm = TRUE), 3)
  } else {
    NA_real_
  }
  sc <- sc[is.finite(P) & P > 0]
  data.table(
    caller = nm, `teosinte presence` = pres, lambda_GC = lam(sc$P),
    `n > perm 5%` = nsig(sc$P, gwas_thr(TR, "mlm", g)),
    `best candidate -log10P` = max_cand_logp(sc)
  )
}))
knitr::kable(ctab2, caption = "SPAD MLM (Family+K) by ancestry caller at 0.4×, with n markers above the caller's own 1000-permutation 5% FWER line and the best (max) -log10P across the chlorophyll candidate windows.")
```

```{r caller-composite, fig.width=9, fig.height=20}
panels <- lapply(names(callers), function(nm) {
  sc <- fread(here(sprintf("data/zeal/spad_gwas_mlm_%s_donor_accession_snp50k.csv", callers[[nm]])))
  plot_manhattan(sc, sprintf("SPAD MLM (Family+K) — %s (λ_GC = %.2f; perm FWER 5%%)", nm, lam(sc$P)), overlap_csv = overlap, lod = gwas_thr(TR, "mlm", callers[[nm]]))
})
cc <- cowplot::plot_grid(plotlist = panels, ncol = 1, labels = LETTERS[1:5], label_size = 18, align = "v", axis = "lr")
ggsave(here("results/sim/zeal/spad_caller_composite_fdr_snp50k.png"), cc, width = 9, height = 20, dpi = 200, bg = "white")
cc
```
