---
title: "ZEAL Ear-Number (EN) GWAS + JLM on the SNP50K panel — 1000-permutation FWER"
subtitle: "The ear-number analog of the PH/EH notebooks: same SNP50K ancestry pipeline, phenotype = per-field SpATS ear-number BLUEs, candidate set = the curated maize ear-number / prolificacy cloned-gene panel (core gt1/tb1/tru1 + inflorescence-architecture branching/meristem/KRN genes). Significance is a genome-wide FWER from a 1000-permutation null (per scan): the Manhattan line is the −log₁₀P exceeded by only 5% of permutations, and candidate/peak calls report whether they clear it."
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**What this is.** The ear-number (EN) sibling of
[`zeal-qtl-recovery-eh-mlm-snp50k`](zeal-qtl-recovery-eh-mlm-snp50k.html), built from the
same real ZEAL pipeline end-to-end: field EN → SpATS spatial BLUEs → SNP50K teosinte-ancestry
dosage → rtiger ancestry mosaic → K → JLM + MLM. The population is the same near-isogenic
B73×teosinte BC2S3 lines from **5 teosinte taxa**. EN is **ear number** — the count of ears borne
per plant — and the per-genotype value is the per-field SpATS mean across the two fields (CLY23-D4,
CLY25-B5). Only the phenotype and the candidate panel change vs the EH notebook. A closely related
[`prolificacy (Prolif)`](zeal-qtl-recovery-prolif-mlm-snp50k.html) notebook shares this candidate
panel.

::: {.callout-important}
## Genotype at 0.4× skim: why ancestry, not per-SNP
SNP50K is 51,991 teosinte-vs-B73 sites pileup-genotyped at ~0.4× — hard genotypes are
unusable (70% missing, ~0% hom-ALT), so ancestry is inferred from the per-sample allele
**counts**. The **per-SNP genotype (bcftools HWE-posterior) (Panel B)** is therefore λ_GC-deflated (sparse
mean-imputation); the **rtiger ancestry mosaic (Panel C)** is the usable, block-smoothed
genotype (teosinte presence 0.111, matching the established BzeaSeq ~0.101). This is the
SNP50K-specific reason ancestry inference is mandatory, and is identical to the DTA/PH/EH notebooks —
only the phenotype changes.
:::

**Trait context.** Ear number / prolificacy is a **domestication-and-inflorescence-architecture**
trait: teosinte is highly branched and prolific (many small ears on long lateral branches), while
B73 has strong apical dominance and typically one or two ears. The trait is therefore expected to
respond strongly to teosinte introgression. The candidate panel is the curated maize ear-number /
prolificacy cloned-gene set, split into two tiers: the **core ear-number regulators** — `gt1`
(*grassy tillers1*; HD-ZIP I TF underlying the **prol1.1** QTL, represses axillary ear-bud growth,
teosinte allele → prolific ears), `tb1` (*teosinte branched1*; the classic domestication TCP TF whose
*Hopscotch* insertion drives apical dominance and reduced ear number), and `tru1` (*tassels replace
upper ears1*; a direct tb1 target promoting ear-vs-tassel identity) — and a second tier of
**inflorescence-architecture / meristem / kernel-row-number** genes (`ub2`/`ub3` SBP-box branching,
`ra1`/`ra2`/`ra3` ramosa branch determinacy, `fea2`/`fea3`/`td1` CLV–WUS meristem-size receptors).
Because teosinte ancestry is broadly ear-number-increasing and **partly collinear with K**, the
mixed model absorbs ancestry-correlated variance, and the model-independent **JLM** (nested-in-family
stepwise) is the primary QTL tool — with `gt1`/`tb1` the two loci most likely to surface as
headline, effect-bearing QTL rather than polygenic background.

```{r setup}
library(here)
library(data.table)
library(ggplot2)
library(fastman)
library(ggrepel)
library(ggtext)
library(scales)
library(cowplot)
source(here("scripts/teonam_notebook_plots.R")) # shared: plot_manhattan/lollipop, lambda_gc, LOD
```

```{r load}
mlm <- fread(here("data/zeal/en_gwas_mlm_rtiger_mosaic_donor_accession_snp50k.csv")) # Panel C
gt <- fread(here("data/zeal/en_gwas_mlm_hwe_post_gt_donor_accession_snp50k.csv")) # Panel B
ols <- fread(here("data/zeal/en_gwas_ols_donor_accession_snp50k.csv"))[is.finite(P) & P > 0]
overlap <- here("results/sim/zeal/en_candidate_overlap.csv")
lam <- function(p) {
  p <- p[is.finite(p) & p > 0 & p <= 1]
  round(median(qchisq(p, 1, lower.tail = FALSE)) / qchisq(0.5, 1), 2)
}
# Significance = genome-wide FWER from a 1000-permutation null (scripts/zeal_gwas_perm.R),
# stored per (trait, model, geno) in data/zeal/gwas_perm_thresholds.csv.
TR <- "EN"
thr <- function(model, geno, a = 0.05) gwas_thr(TR, model, geno, a)
thr_ols <- thr("ols", "rtiger_mosaic")
thr_mlm <- thr("mlm", "rtiger_mosaic")
thr_gt <- thr("mlm", "hwe_post_gt")
nsig <- function(p, t) sum(-log10(p) > t, na.rm = TRUE)
```

::: {.callout-note}
## Significance = 1000-permutation genome-wide FWER (5%)
Every Manhattan's dotted line is the **−log₁₀P exceeded by only 5% of 1000 permutations** of
that scan (a genome-wide max-−log₁₀P null; `scripts/zeal_gwas_perm.R`), stored per
(trait, model, geno) in `data/zeal/gwas_perm_thresholds.csv`. A marker is significant iff its
−log₁₀P clears that line. The line's height differs per panel with the null. Candidate/peak
tables report whether the marker clears the 5% FWER line.
:::

## Candidate gene registry

Curated evidence for each **candidate** (tier, causal direction, evidence type, phenotype, and a DOI link) from the tracked registry (`reference/candidate_genes.csv` × `candidate_evidence.csv`). These are the genes starred on the Manhattans and labelled on the lollipop; loci without B73 v5 coordinates still appear here but are not plotted.

```{r candidate-registry}
candidate_table(TR)
```

## Candidate genes: closest marker vs window-max marker

For each maize ear-number / prolificacy candidate (B73 v5), the closest marker to the gene body vs
the highest −log₁₀P marker within ±500 kb in the rtiger-mosaic **MLM (Family+K)** scan (Panel C).
Candidates are the curated cloned ear-number / prolificacy panel (core `gt1`/`tb1`/`tru1` +
inflorescence-architecture branching/meristem/KRN genes; v5 coordinates); `pathway` records the
tier and protein/role each acts in.

```{r candidate-markers}
sc <- copy(mlm)[is.finite(P) & P > 0][, logP := -log10(P)]
gg <- fread(here("data/teonam/en_candidate_genes.tsv"))
gg[, chr := as.integer(chr)][, qtl := ifelse(qtl_chen2019 == "" | is.na(qtl_chen2019), paste0("EN(chr", chr, ")"), qtl_chen2019)]
WIN <- 5e5
ctab <- rbindlist(lapply(seq_len(nrow(gg)), function(i) {
  ch <- gg$chr[i]
  s <- gg$start[i]
  e <- gg$end[i]
  m <- sc[CHR == ch]
  if (!nrow(m)) {
    return(NULL)
  }
  d <- pmax(0L, pmax(s - m$BP, m$BP - e))
  kc <- which.min(d)
  w <- m[BP >= s - WIN & BP <= e + WIN]
  kw <- if (nrow(w)) which.max(w$logP) else NA
  data.table(
    gene = gg$symbol[i], pathway = gg$pathway[i], chr = ch, `gene start (v5)` = s,
    `closest Δstart (bp)` = m$BP[kc] - s, `closest -log10P` = round(m$logP[kc], 2),
    `window-max -log10P` = if (nrow(w)) round(w$logP[kw], 2) else NA_real_,
    `clears perm 5%` = if (nrow(w)) (w$logP[kw] >= thr_mlm) else NA
  )
}))
setorder(ctab, chr, `gene start (v5)`)
knitr::kable(ctab,
  format.args = list(big.mark = ","),
  caption = sprintf("EN candidates (B73 v5; curated maize ear-number / prolificacy panel): closest vs window-max (±500 kb) marker, rtiger-mosaic MLM (Family+K) scan; 'clears perm 5%%' = window-max exceeds the 1000-permutation FWER line (-log10P = %.2f).", thr_mlm)
)
```

## OLS → MLM: structure control

Here **Family** is the **82 donor-accession** fixed factor (parent-plant-level structure correction) that the MLM and per-marker OLS scans were actually computed with — a stronger correction than the 5 donor **taxa** the JLM nests on. It likely over-corrects population structure and costs some power; a taxon-level (5) MLM/OLS rerun is a pending comparison (see `TERMINOLOGY.md`).

The fixed-effect OLS (Family + marker, no K) vs the MLM (Family + K) on the rtiger mosaic, same
fixed part. OLS is intrinsically inflated (ancestry-block LD, unfixable by any fixed factor);
K deflates λ toward 1 and removes the ancestry-correlated bulk that a near-isogenic,
teosinte-ancestry-structured ear-number panel carries.

```{r ols-manhattan, fig.width=9, fig.height=4.4}
plot_manhattan(ols,
  sprintf("ZEAL EN — OLS (Family + marker, no K), rtiger mosaic (λ_GC = %.2f; dotted = perm FWER 5%%)", lam(ols$P)),
  overlap_csv = overlap, lod = thr_ols,
  out_png = here("results/sim/zeal/en_gwas_ols_manhattan_fdr_snp50k.png")
)
```

The OLS Manhattan (no K) is genome-wide inflated (ancestry-block LD) — the same structure
artifact the MLM K term removes below.

```{r ols-vs-mlm}
lam_tab <- data.table(
  model = c("OLS (Family + marker, no K)", "MLM (Family + K)"),
  markers = c(nrow(ols), nrow(mlm)),
  lambda_GC = c(lam(ols$P), lam(mlm$P)),
  `n > perm 5%` = c(nsig(ols$P, thr_ols), nsig(mlm$P, thr_mlm)),
  max_logP = round(c(max(-log10(ols$P)), max(-log10(mlm$P))), 2)
)
knitr::kable(lam_tab, caption = "OLS vs MLM on the rtiger mosaic (same Family fixed part): K deflates genomic inflation; n markers above each scan's 1000-permutation 5% FWER line.")
```

```{r ols-vs-mlm-peaks}
ov <- fread(overlap)
peak_at <- function(scan, ch, s) {
  w <- scan[CHR == ch & abs(BP - s) <= 5e5 & is.finite(P) & P > 0]
  if (!nrow(w)) NA_real_ else round(max(-log10(w$P)), 2)
}
cmp <- ov[, .(
  gene = symbol, chr,
  `OLS peak` = mapply(peak_at, list(ols), chr, start),
  `MLM peak` = mapply(peak_at, list(mlm), chr, start)
)]
cmp[, `drop (OLS−MLM)` := round(`OLS peak` - `MLM peak`, 2)]
setorder(cmp, -`OLS peak`)
knitr::kable(cmp, caption = "Local peak -log10P (±500 kb) per candidate: OLS vs MLM. K suppresses the ancestry-collinear ear-number signal.")
```

## JLM QTL genomic distribution (Panel A)

EN QTL from JLM (TASSEL StepwiseOLS, nested in taxon, entry by a 1000-permutation empirical α = 0.05 (Chen-style genome-wide FWER; per-trait)). JLM QTL are
selected by the stepwise enter/exit P (a model-selection gate), independent of the per-scan
permutation FWER line used for the GWAS Manhattans below. Model-independent — the primary EN analysis, as
in Chen 2019.

```{r lollipop-jlm, fig.width=9, fig.height=4}
p_lolli <- plot_lollipop(
  here("results/sim/zeal/tassel/en_jlm_native1.txt"), mlm,
  "ZEAL EN — JLM QTL genomic distribution (nested in taxon)",
  overlap_csv = overlap,
  out_png = here("results/sim/zeal/en_jlm_lollipop_snp50k.png")
)
p_lolli
```

## MLM (Family + K), rtiger ancestry mosaic (Panel C)

```{r manhattan-mlm, fig.width=9, fig.height=4.4}
p_mlm <- plot_manhattan(mlm,
  sprintf("ZEAL EN — MLM (Family + K), rtiger ancestry mosaic (λ_GC = %.2f; dotted = perm FWER 5%%)", lam(mlm$P)),
  overlap_csv = overlap, lod = thr_mlm, out_png = here("results/sim/zeal/en_gwas_mlm_rtiger_mosaic_manhattan_fdr_snp50k.png")
)
p_mlm
cat(sprintf("%d markers clear the 5%% FWER line (-log10P = %.2f)\n", nsig(mlm$P, thr_mlm), thr_mlm))
```

## Composite (A: JLM · B: per-SNP MLM · C: rtiger-mosaic MLM)

Panel B (per-SNP genotype (bcftools HWE-posterior)) is λ_GC-deflated — the 70%-missing per-SNP panel at 0.4×
is sparse — but catches sharp SNP hits the mosaic block-terraces away. Panel C (the ancestry
mosaic) is the usable, well-calibrated genotype. The B↔C contrast is the notebook's
per-SNP-scatter vs ancestry-block-smoothing, and the demonstration that ancestry inference is
mandatory at low coverage — identical in mechanism to the DTA/PH/EH notebooks.

```{r composite, fig.width=9, fig.height=13}
p_B <- plot_manhattan(gt,
  sprintf("EN — MLM (Family+K), per-SNP genotype (bcftools HWE-posterior) (λ_GC = %.2f)", lam(gt$P)),
  overlap_csv = overlap, lod = thr_gt
)
p_C <- plot_manhattan(mlm,
  sprintf("EN — MLM (Family+K), rtiger ancestry mosaic (λ_GC = %.2f)", lam(mlm$P)),
  overlap_csv = overlap, lod = thr_mlm
)
composite <- cowplot::plot_grid(p_lolli, p_B, p_C, ncol = 1, labels = c("A", "B", "C"), label_size = 20, align = "v", axis = "lr")
ggsave(here("results/sim/zeal/en_composite_snp50k.png"), composite, width = 9, height = 13, dpi = 200, bg = "white")
composite
```

## Multi-caller EN scan (control · nnil · rtiger · lbimpute · binhmm)

The same EN MLM (Family+K) scan on each ancestry caller's recovered genotypes, at the real
~0.4× depth (not a coverage sweep). The robust, trait-independent finding is **teosinte-presence
recovery**: the block/bin-aggregated callers (rtiger 0.111; binhmm 0.101) recover the established
BzeaSeq fraction (~0.10), while per-SNP calling (control) and the per-SNP HMM recalls (nnil,
lbimpute) **under-call**. Genomic inflation is phenotype-dependent (reported per caller below).
Caller genotypes are population-level and trait-independent; only the phenotype (EN) changes.

```{r caller-table}
callers <- c(control = "hwe_post_gt", nnil = "nnil_mosaic", rtiger = "rtiger_mosaic", lbimpute = "lbimpute_mosaic", binhmm = "binhmm_mosaic")
gg <- fread(here("data/teonam/en_candidate_genes.tsv"))[, chr := as.integer(chr)]
# best (max -log10P) EN candidate window per caller
max_cand_logp <- function(sc) {
  lp <- -log10(sc$P)
  vals <- unlist(lapply(seq_len(nrow(gg)), function(i) {
    idx <- which(sc$CHR == gg$chr[i] & abs(sc$BP - gg$start[i]) <= 5e5)
    if (length(idx)) max(lp[idx], na.rm = TRUE) else NA_real_
  }))
  if (all(is.na(vals))) NA_real_ else round(max(vals, na.rm = TRUE), 2)
}
ctab2 <- rbindlist(lapply(names(callers), function(nm) {
  g <- callers[[nm]]
  sc <- fread(here(sprintf("data/zeal/en_gwas_mlm_%s_donor_accession_snp50k.csv", g)))
  mf <- here(sprintf("data/zeal/zeal_%s.rds", g))
  pres <- if (file.exists(mf)) {
    st <- readRDS(mf)$state
    round(mean(st %in% c(1L, 2L), na.rm = TRUE), 3)
  } else {
    NA_real_
  }
  sc <- sc[is.finite(P) & P > 0]
  data.table(
    caller = nm, `teosinte presence` = pres, lambda_GC = lam(sc$P),
    `n > perm 5%` = nsig(sc$P, gwas_thr(TR, "mlm", g)),
    `best candidate -log10P` = max_cand_logp(sc)
  )
}))
knitr::kable(ctab2, caption = "EN MLM (Family+K) by ancestry caller at 0.4×, with n markers above the caller's own 1000-permutation 5% FWER line and the best (max) -log10P across the EN ear-number / prolificacy candidate windows.")
```

```{r caller-composite, fig.width=9, fig.height=20}
panels <- lapply(names(callers), function(nm) {
  sc <- fread(here(sprintf("data/zeal/en_gwas_mlm_%s_donor_accession_snp50k.csv", callers[[nm]])))
  plot_manhattan(sc, sprintf("EN MLM (Family+K) — %s (λ_GC = %.2f; perm FWER 5%%)", nm, lam(sc$P)), overlap_csv = overlap, lod = gwas_thr(TR, "mlm", callers[[nm]]))
})
cc <- cowplot::plot_grid(plotlist = panels, ncol = 1, labels = LETTERS[1:5], label_size = 18, align = "v", axis = "lr")
ggsave(here("results/sim/zeal/en_caller_composite_fdr_snp50k.png"), cc, width = 9, height = 20, dpi = 200, bg = "white")
cc
```
