---
title: "ZEAL/BZea Leaf-Greenness SPAD (20 DAS) GWAS + JLM on the SNP50K panel ÔÇö FDR Ôëñ 0.05"
subtitle: "SPAD leaf greenness at 20 days after sowing (20DAS). Unlike the spatially-corrected single-SPAD sibling, the phenotype here is the zealbrowser value used AS-IS (per-line mean of the relabeled Bzea_merged.csv column, no additional spatial correction). Same SNP50K ancestry pipeline and maize chlorophyll candidate panel. Significance is BenjaminiÔÇôHochberg FDR Ôëñ 0.05 (per-scan q-values)."
author: "Fausto Rodr├¡guez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**What this is.** One of three SPAD notebooks. SPAD leaf greenness was measured at **two dates**
(20 and 36 days after sowing). This notebook maps the **20DAS** reading; its siblings map the
**36DAS** reading ([`spad36das`](zeal-qtl-recovery-spad36das-mlm-snp50k-fdr.html)) and a single
spatially-corrected SPAD BLUE ([`spad`](zeal-qtl-recovery-spad-mlm-snp50k-fdr.html)). Same SNP50K
ancestry pipeline ÔåÆ RTIGER mosaic ÔåÆ K ÔåÆ JLM + MLM.

::: {.callout-note}
## Phenotype provenance (20DAS) ÔÇö used AS-IS
- **Source:** column `SPAD_leaf_greenness_20DAS` in `agent/introfinder/Bzea_merged.csv` (the
  relabeled phenotype file the zealbrowser reads).
- **Per genotype:** the **mean over that line's rows** (`scripts/zeal_spad_dates.R`,
  `TRAIT=SPAD20DAS PHENO=direct`). This is the only aggregation applied.
- **No additional correction:** the merged file has no plot/row/range, so **no SpATS / spatial
  correction is done here** ÔÇö the values are taken as provided (they may already have been adjusted
  upstream). This is the deliberate contrast with the spatially-corrected single-SPAD notebook.
- GWAS on the RTIGER-mosaic MLM with `FAMILY_COL=donor_accession`.
:::

::: {.callout-important}
## Genotype at 0.4├ù skim: why ancestry, not per-SNP
SNP50K is 51,991 teosinte-vs-B73 sites pileup-genotyped at ~0.4├ù ÔÇö hard genotypes are
unusable (70% missing, ~0% hom-ALT), so ancestry is inferred from the per-sample allele
**counts**. The **per-SNP genotype (bcftools HWE-posterior) (Panel B)** is therefore ╬╗_GC-deflated (sparse
mean-imputation); the **RTIGER ancestry mosaic (Panel C)** is the usable, block-smoothed
genotype (teosinte presence 0.111, matching the established BzeaSeq ~0.101).
:::

**Trait context.** SPAD is a leaf-chlorophyll proxy. The candidate set spans the maize chlorophyll
pathway and its regulators: biosynthesis (`chlh1`, `oy1`, `gtr3`, `urod`, `por1/2`), catabolism /
stay-green (`pph`, `nye1/2`, `sgrl1`, `lls1`, `nyc1`), light-harvesting (`lhcb10`, `cab1/2`), and
regulators (`nactf132`, GOLDEN2-LIKE `g2`/`glk1`). The model-independent **JLM** is the primary QTL
tool.

```{r setup}
library(here)
library(data.table)
library(ggplot2)
library(fastman)
library(ggrepel)
library(ggtext)
library(scales)
library(cowplot)
source(here("scripts/teonam_notebook_plots.R"))
```

```{r load}
mlm <- fread(here("data/zeal/spad20das_gwas_mlm_rtiger_mosaic_donor_accession_snp50k.csv"))
gt <- fread(here("data/zeal/spad20das_gwas_mlm_hwe_post_gt_donor_accession_snp50k.csv"))
ols <- fread(here("data/zeal/spad20das_gwas_ols_donor_accession_snp50k.csv"))[is.finite(P) & P > 0]
overlap <- here("results/sim/zeal/spad20das_candidate_overlap.csv")
lam <- function(p) {
  p <- p[is.finite(p) & p > 0 & p <= 1]
  round(median(qchisq(p, 1, lower.tail = FALSE)) / qchisq(0.5, 1), 2)
}
ALPHA <- 0.05
fdr_lod <- function(p, alpha = ALPHA) {
  p <- p[is.finite(p) & p > 0]
  q <- p.adjust(p, "BH")
  if (!any(q <= alpha)) {
    return(NA_real_)
  }
  -log10(max(p[q <= alpha]))
}
qval <- function(p) p.adjust(p, "BH")
```

::: {.callout-note}
## Significance = BenjaminiÔÇôHochberg FDR Ôëñ 0.05
Every Manhattan's dotted line is the **ÔêÆlogÔéüÔéÇP where BH q = 0.05** for that scan; a marker is
significant iff BH q Ôëñ 0.05. If none pass, no line is drawn. Candidate/peak tables report q.
:::

## Candidate genes: closest marker vs window-max marker

```{r candidate-markers}
sc <- copy(mlm)[is.finite(P) & P > 0][, logP := -log10(P)][, q := qval(P)]
gg <- fread(here("data/teonam/spad20das_candidate_genes.tsv"))
gg[, chr := as.integer(chr)]
WIN <- 5e5
ctab <- rbindlist(lapply(seq_len(nrow(gg)), function(i) {
  ch <- gg$chr[i]
  s <- gg$start[i]
  e <- gg$end[i]
  m <- sc[CHR == ch]
  if (!nrow(m)) {
    return(NULL)
  }
  d <- pmax(0L, pmax(s - m$BP, m$BP - e))
  kc <- which.min(d)
  w <- m[BP >= s - WIN & BP <= e + WIN]
  kw <- if (nrow(w)) which.max(w$logP) else NA
  data.table(
    gene = gg$symbol[i], pathway = gg$pathway[i], chr = ch, `gene start (v5)` = s,
    `closest ╬östart (bp)` = m$BP[kc] - s, `closest -log10P` = round(m$logP[kc], 2),
    `window-max -log10P` = if (nrow(w)) round(w$logP[kw], 2) else NA_real_,
    `window-max q (BH)` = if (nrow(w)) signif(w$q[kw], 2) else NA_real_,
    `FDR<=0.05` = if (nrow(w)) w$q[kw] <= ALPHA else NA
  )
}))
setorder(ctab, chr, `gene start (v5)`)
knitr::kable(ctab,
  format.args = list(big.mark = ","),
  caption = "SPAD (20DAS) candidates (B73 v5 maize chlorophyll panel): closest vs window-max (┬▒500 kb) marker, RTIGER-mosaic MLM (Taxon+K) scan, with BenjaminiÔÇôHochberg q at the window-max marker."
)
```

## OLS ÔåÆ MLM: structure control

```{r ols-manhattan, fig.width=9, fig.height=4.4}
plot_manhattan(ols,
  sprintf("ZEAL/BZea SPAD 20DAS ÔÇö OLS (Taxon + marker, no K), RTIGER mosaic (╬╗_GC = %.2f; dotted = FDR 0.05)", lam(ols$P)),
  overlap_csv = overlap, lod = fdr_lod(ols$P),
  out_png = here("results/sim/zeal/spad20das_gwas_ols_manhattan_fdr_snp50k.png")
)
```

```{r ols-vs-mlm}
lam_tab <- data.table(
  model = c("OLS (Taxon + marker, no K)", "MLM (Taxon + K)"),
  markers = c(nrow(ols), nrow(mlm)),
  lambda_GC = c(lam(ols$P), lam(mlm$P)),
  `n FDR<=0.05` = c(sum(qval(ols$P) <= ALPHA), sum(qval(mlm$P) <= ALPHA)),
  max_logP = round(c(max(-log10(ols$P)), max(-log10(mlm$P))), 2)
)
knitr::kable(lam_tab, caption = "OLS vs MLM on the RTIGER mosaic (same Taxon fixed part): K deflates genomic inflation; n markers at BH FDR Ôëñ 0.05.")
```

```{r ols-vs-mlm-peaks}
ov <- fread(overlap)
peak_at <- function(scan, ch, s) {
  w <- scan[CHR == ch & abs(BP - s) <= 5e5 & is.finite(P) & P > 0]
  if (!nrow(w)) NA_real_ else round(max(-log10(w$P)), 2)
}
cmp <- ov[, .(
  gene = symbol, chr,
  `OLS peak` = mapply(peak_at, list(ols), chr, start),
  `MLM peak` = mapply(peak_at, list(mlm), chr, start)
)]
cmp[, `drop (OLSÔêÆMLM)` := round(`OLS peak` - `MLM peak`, 2)]
setorder(cmp, -`OLS peak`)
knitr::kable(cmp, caption = "Local peak -log10P (┬▒500 kb) per candidate: OLS vs MLM.")
```

## JLM QTL genomic distribution (Panel A)

```{r lollipop-jlm, fig.width=9, fig.height=4}
p_lolli <- plot_lollipop(
  here("results/sim/zeal/tassel/spad20das_jlm_native1.txt"), mlm,
  "ZEAL/BZea SPAD 20DAS ÔÇö JLM QTL genomic distribution (nested in taxon)",
  overlap_csv = overlap, label_genes = c("chlh1", "gtr3", "urod", "oy1", "por1", "por2", "pph", "nye1", "nye2", "sgrl1", "lls1", "nyc1", "lhcb10", "cab1", "cab2", "nactf132", "g2", "glk1"),
  out_png = here("results/sim/zeal/spad20das_jlm_lollipop_snp50k.png")
)
p_lolli
```

## MLM (Taxon + K), RTIGER ancestry mosaic (Panel C)

```{r manhattan-mlm, fig.width=9, fig.height=4.4}
p_mlm <- plot_manhattan(mlm,
  sprintf("ZEAL/BZea SPAD 20DAS ÔÇö MLM (Taxon + K), RTIGER ancestry mosaic (╬╗_GC = %.2f; dotted = FDR 0.05)", lam(mlm$P)),
  overlap_csv = overlap, lod = fdr_lod(mlm$P), out_png = here("results/sim/zeal/spad20das_gwas_mlm_rtiger_mosaic_manhattan_fdr_snp50k.png")
)
p_mlm
cat(sprintf("%d markers pass FDR <= 0.05 (BH); FDR line at -log10P = %s\n", sum(qval(mlm$P) <= ALPHA), ifelse(is.na(fdr_lod(mlm$P)), "NA (none)", sprintf("%.2f", fdr_lod(mlm$P)))))
```

## Composite (A: JLM ┬À B: per-SNP MLM ┬À C: RTIGER-mosaic MLM)

```{r composite, fig.width=9, fig.height=13}
p_B <- plot_manhattan(gt,
  sprintf("SPAD 20DAS ÔÇö MLM (Taxon+K), per-SNP genotype (bcftools HWE-posterior) (╬╗_GC = %.2f)", lam(gt$P)),
  overlap_csv = overlap, lod = fdr_lod(gt$P)
)
p_C <- plot_manhattan(mlm,
  sprintf("SPAD 20DAS ÔÇö MLM (Taxon+K), RTIGER ancestry mosaic (╬╗_GC = %.2f)", lam(mlm$P)),
  overlap_csv = overlap, lod = fdr_lod(mlm$P)
)
composite <- cowplot::plot_grid(p_lolli, p_B, p_C, ncol = 1, labels = c("A", "B", "C"), label_size = 20, align = "v", axis = "lr")
ggsave(here("results/sim/zeal/spad20das_composite_snp50k.png"), composite, width = 9, height = 13, dpi = 200, bg = "white")
composite
```

## Multi-caller SPAD 20DAS scan (control ┬À nNIL ┬À RTIGER ┬À LB-Impute ┬À binhmm)

```{r caller-table}
callers <- c(control = "hwe_post_gt", nNIL = "nnil_mosaic", RTIGER = "rtiger_mosaic", `LB-Impute` = "lbimpute_mosaic", binhmm = "binhmm_mosaic")
gg <- fread(here("data/teonam/spad20das_candidate_genes.tsv"))[, chr := as.integer(chr)]
min_cand_q <- function(sc) {
  q <- qval(sc$P)
  vals <- unlist(lapply(seq_len(nrow(gg)), function(i) {
    idx <- which(sc$CHR == gg$chr[i] & abs(sc$BP - gg$start[i]) <= 5e5)
    if (length(idx)) min(q[idx]) else NA_real_
  }))
  if (all(is.na(vals))) NA_real_ else signif(min(vals, na.rm = TRUE), 2)
}
ctab2 <- rbindlist(lapply(names(callers), function(nm) {
  g <- callers[[nm]]
  sc <- fread(here(sprintf("data/zeal/spad20das_gwas_mlm_%s_donor_accession_snp50k.csv", g)))
  mf <- here(sprintf("data/zeal/zeal_%s.rds", g))
  pres <- if (file.exists(mf)) {
    st <- readRDS(mf)$state
    round(mean(st %in% c(1L, 2L), na.rm = TRUE), 3)
  } else {
    NA_real_
  }
  sc <- sc[is.finite(P) & P > 0]
  data.table(
    caller = nm, `teosinte presence` = pres, lambda_GC = lam(sc$P),
    `n FDR<=0.05` = sum(qval(sc$P) <= ALPHA),
    `best candidate q (BH)` = min_cand_q(sc)
  )
}))
knitr::kable(ctab2, caption = "SPAD 20DAS MLM (Taxon+K) by ancestry caller at 0.4├ù, with n markers at BH FDR Ôëñ 0.05 and the best (min) q across the chlorophyll candidate windows.")
```

```{r caller-composite, fig.width=9, fig.height=20}
panels <- lapply(names(callers), function(nm) {
  sc <- fread(here(sprintf("data/zeal/spad20das_gwas_mlm_%s_donor_accession_snp50k.csv", callers[[nm]])))
  plot_manhattan(sc, sprintf("SPAD 20DAS MLM (Taxon+K) ÔÇö %s (╬╗_GC = %.2f; FDR 0.05)", nm, lam(sc$P)), overlap_csv = overlap, lod = fdr_lod(sc$P))
})
cc <- cowplot::plot_grid(plotlist = panels, ncol = 1, labels = LETTERS[1:5], label_size = 18, align = "v", axis = "lr")
ggsave(here("results/sim/zeal/spad20das_caller_composite_fdr_snp50k.png"), cc, width = 9, height = 20, dpi = 200, bg = "white")
cc
```
