---
title: "ZEAL/BZea DTA GWAS + JLM on the SNP50K panel ÔÇö 1000-permutation FWER"
subtitle: "GWAS + JLM on the SNP50K panel. Significance is a genome-wide FWER from a 1000-permutation null (per scan): the Manhattan line is the -log10P exceeded by only 5% of permutations, and candidate/peak calls report whether they clear it."
author: "Fausto Rodr├şguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**What this is.** The ZEAL/BZea analog of
[`teonam-qtl-recovery-dta-mlm-118k`](teonam-qtl-recovery-dta-mlm-118k.html), built from real
BZea data end-to-end: field DTA Ôćĺ SpATS spatial BLUEs Ôćĺ SNP50K teosinte-ancestry dosage Ôćĺ
RTIGER ancestry mosaic Ôćĺ K Ôćĺ JLM + MLM. The population is 1,403 near-isogenic B73├Śteosinte
BC2S3 lines from **5 teosinte taxa** (Zx *mexicana* 573, Zv *parviglumis* 383, Zd
*diploperennis* 254, Zl *luxurians* 118, Zh *huehuetenangensis* 61) ÔÇö the analog of TeoNAM's
5 TIL families. The DTA BLUE is the per-field SpATS mean across two fields (CLY23-D4, CLY25-B5;
cross-year genetic corr r = 0.675).

::: {.callout-important}
## Genotype at 0.4├Ś skim: why ancestry, not per-SNP
SNP50K is 51,991 teosinte-vs-B73 sites pileup-genotyped at ~0.4├Ś ÔÇö hard genotypes are
unusable (70% missing, ~0% hom-ALT), so ancestry is inferred from the per-sample allele
**counts**. The **per-SNP genotype (bcftools HWE-posterior) (Panel B)** is therefore ╬╗_GC-deflated (sparse
mean-imputation); the **RTIGER ancestry mosaic (Panel C)** is the usable, block-smoothed
genotype (teosinte presence 0.111, matching the established BzeaSeq ~0.101). This is the
ZEAL-specific reason ancestry inference is mandatory ÔÇö and the direct analog of the
per-SNP-vs-block contrast in the TeoNAM notebook.
:::

The central result mirrors TeoNAM: **K = teosinte-ancestry relatedness is collinear with the
flowering QTL**, so the mixed model absorbs the very signal it protects ÔÇö the flowering
candidates (`ZmCCT10`, `ZCN8`) are suppressed under MLM. As in Chen 2019, the DTA QTL are
therefore the province of **JLM** (nested-in-family stepwise), not the MLM Manhattan.

```{r setup}
library(here)
library(data.table)
library(ggplot2)
library(fastman)
library(ggrepel)
library(ggtext)
library(scales)
library(cowplot)
source(here("scripts/teonam_notebook_plots.R")) # shared: plot_manhattan/lollipop, lambda_gc, LOD
```

```{r load}
mlm <- fread(here("data/zeal/dta_gwas_mlm_rtiger_mosaic_donor_accession_snp50k.csv")) # Panel C
gt <- fread(here("data/zeal/dta_gwas_mlm_hwe_post_gt_donor_accession_snp50k.csv")) # Panel B
ols <- fread(here("data/zeal/dta_gwas_ols_donor_accession_snp50k.csv"))[is.finite(P) & P > 0]
overlap <- here("results/sim/zeal/dta_candidate_overlap.csv")
lam <- function(p) {
  p <- p[is.finite(p) & p > 0 & p <= 1]
  round(median(qchisq(p, 1, lower.tail = FALSE)) / qchisq(0.5, 1), 2)
}
# Significance = genome-wide FWER from a 1000-permutation null (scripts/zeal_gwas_perm.R),
# stored per (trait, model, geno) in data/zeal/gwas_perm_thresholds.csv. thr() reads the
# -log10P line for a panel; a marker is significant iff its -log10P exceeds that line.
TR <- "DTA"
thr <- function(model, geno, a = 0.05) gwas_thr(TR, model, geno, a)
thr_ols <- thr("ols", "rtiger_mosaic") # OLS runs on the RTIGER mosaic
thr_mlm <- thr("mlm", "rtiger_mosaic") # Panel C
thr_gt <- thr("mlm", "hwe_post_gt") # Panel B (per-SNP)
nsig <- function(p, t) sum(-log10(p) > t, na.rm = TRUE) # markers above the FWER line
```

::: {.callout-note}
## Significance = 1000-permutation genome-wide FWER (5%)
Every Manhattan's dotted line is the **ÔłĺlogÔéüÔéÇP exceeded by only 5% of 1000 permutations** of
that scan (a genome-wide max-ÔłĺlogÔéüÔéÇP null; `scripts/zeal_gwas_perm.R`), stored per
(trait, model, geno) in `data/zeal/gwas_perm_thresholds.csv`. A marker is significant iff its
ÔłĺlogÔéüÔéÇP clears that line. The line's height differs per panel with the null. Candidate/peak
tables report whether the marker clears the 5% FWER line.
:::

## Candidate gene registry

Curated evidence for each **candidate** (tier, causal direction, evidence type, phenotype, and a DOI link) from the tracked registry (`reference/candidate_genes.csv` ├Ś `candidate_evidence.csv`). These are the genes starred on the Manhattans and labelled on the lollipop; loci without B73 v5 coordinates still appear here but are not plotted.

```{r candidate-registry}
candidate_table(TR)
```

## Candidate genes: closest marker vs window-max marker

For each maize flowering candidate (B73 v5), the closest marker to the gene body vs the
highest ÔłĺlogÔéüÔéÇP marker within ┬▒500 kb in the RTIGER-mosaic **MLM (Taxon+K)** scan (Panel C).
The same v5 flowering candidates as TeoNAM (these are maize genes; teosinte introgression at
them drives DTA in both populations).

```{r candidate-markers}
sc <- copy(mlm)[is.finite(P) & P > 0][, logP := -log10(P)]
gg <- fread(here("data/teonam/dta_candidate_genes.tsv"))
gg[, chr := as.integer(chr)][, qtl := ifelse(qtl_chen2019 == "" | is.na(qtl_chen2019), paste0("DTA(chr", chr, ")"), qtl_chen2019)]
WIN <- 5e5
ctab <- rbindlist(lapply(seq_len(nrow(gg)), function(i) {
  ch <- gg$chr[i]
  s <- gg$start[i]
  e <- gg$end[i]
  m <- sc[CHR == ch]
  if (!nrow(m)) {
    return(NULL)
  }
  d <- pmax(0L, pmax(s - m$BP, m$BP - e))
  kc <- which.min(d)
  w <- m[BP >= s - WIN & BP <= e + WIN]
  kw <- if (nrow(w)) which.max(w$logP) else NA
  data.table(
    gene = gg$symbol[i], QTL = gg$qtl[i], chr = ch, `gene start (v5)` = s,
    `closest ╬östart (bp)` = m$BP[kc] - s, `closest -log10P` = round(m$logP[kc], 2),
    `window-max -log10P` = if (nrow(w)) round(w$logP[kw], 2) else NA_real_,
    `clears perm 5%` = if (nrow(w)) (w$logP[kw] >= thr_mlm) else NA
  )
}))
setorder(ctab, chr, `gene start (v5)`)
knitr::kable(ctab,
  format.args = list(big.mark = ","),
  caption = sprintf("DTA candidates (B73 v5): closest vs window-max (┬▒500 kb) marker, RTIGER-mosaic MLM (Taxon+K) scan; 'clears perm 5%%' = window-max exceeds the 1000-permutation FWER line (-log10P = %.2f).", thr_mlm)
)
```

## OLS Ôćĺ MLM: the flowering confound and structure control

The fixed-effect OLS (Taxon + marker, no K) vs the MLM (Taxon + K) on the RTIGER mosaic, same
fixed part. OLS is intrinsically inflated (ancestry-block LD, unfixable by any fixed factor);
K deflates ╬╗ toward 1 **and** collapses the flowering peaks (`ZmCCT10`), the ancestry-QTL
confound.

```{r ols-manhattan, fig.width=9, fig.height=4.4}
plot_manhattan(ols,
  sprintf("ZEAL/BZea DTA ÔÇö OLS (Taxon + marker, no K), RTIGER mosaic (╬╗_GC = %.2f; dotted = perm FWER 5%%)", lam(ols$P)),
  overlap_csv = overlap, lod = thr_ols,
  out_png = here("results/sim/zeal/dta_gwas_ols_manhattan_fdr_snp50k.png")
)
```

The OLS Manhattan (no K) is genome-wide inflated (╬╗_GC Ôëł 2.4, ancestry-block LD) ÔÇö many
markers cross even the permutation line, and *ZmCCT10* is nominally strong ÔÇö precisely the structure
artifact the MLM K term removes below.

```{r ols-vs-mlm}
lam_tab <- data.table(
  model = c("OLS (Taxon + marker, no K)", "MLM (Taxon + K)"),
  markers = c(nrow(ols), nrow(mlm)),
  lambda_GC = c(lam(ols$P), lam(mlm$P)),
  `n > perm 5%` = c(nsig(ols$P, thr_ols), nsig(mlm$P, thr_mlm)),
  max_logP = round(c(max(-log10(ols$P)), max(-log10(mlm$P))), 2)
)
knitr::kable(lam_tab, caption = "OLS vs MLM on the RTIGER mosaic (same Taxon fixed part): K deflates genomic inflation; n markers above each scan's 1000-permutation 5% FWER line.")
```

```{r ols-vs-mlm-peaks}
ov <- fread(overlap)
peak_at <- function(scan, ch, s) {
  w <- scan[CHR == ch & abs(BP - s) <= 5e5 & is.finite(P) & P > 0]
  if (!nrow(w)) NA_real_ else round(max(-log10(w$P)), 2)
}
cmp <- ov[, .(
  gene = symbol, chr,
  `OLS peak` = mapply(peak_at, list(ols), chr, start),
  `MLM peak` = mapply(peak_at, list(mlm), chr, start)
)]
cmp[, `drop (OLSÔłĺMLM)` := round(`OLS peak` - `MLM peak`, 2)]
setorder(cmp, -`OLS peak`)
knitr::kable(cmp, caption = "Local peak -log10P (┬▒500 kb) per candidate: OLS vs MLM. K suppresses the ancestry-collinear flowering peaks.")
```

## JLM QTL genomic distribution (Panel A)

DTA QTL from JLM (TASSEL StepwiseOLS, nested in taxon, entry by a 1000-permutation empirical ╬▒ = 0.05 (Chen-style genome-wide FWER; per-trait)). JLM QTL are
selected by the stepwise enter/exit P (a model-selection gate), independent of the per-scan
permutation FWER line used for the GWAS Manhattans below. Model-independent ÔÇö Chen's primary DTA analysis.

```{r lollipop-jlm, fig.width=9, fig.height=4}
p_lolli <- plot_lollipop(
  here("results/sim/zeal/tassel/dta_jlm_native1.txt"), mlm,
  "ZEAL/BZea DTA ÔÇö JLM QTL genomic distribution (nested in taxon)",
  overlap_csv = overlap,
  out_png = here("results/sim/zeal/dta_jlm_lollipop_snp50k.png")
)
p_lolli
```

## MLM (Taxon + K), RTIGER ancestry mosaic (Panel C)

```{r manhattan-mlm, fig.width=9, fig.height=4.4}
p_mlm <- plot_manhattan(mlm,
  sprintf("ZEAL/BZea DTA ÔÇö MLM (Taxon + K), RTIGER ancestry mosaic (╬╗_GC = %.2f; dotted = perm FWER 5%%)", lam(mlm$P)),
  overlap_csv = overlap, lod = thr_mlm, out_png = here("results/sim/zeal/dta_gwas_mlm_rtiger_mosaic_manhattan_fdr_snp50k.png")
)
p_mlm
cat(sprintf("%d markers clear the 5%% FWER line (-log10P = %.2f)\n", nsig(mlm$P, thr_mlm), thr_mlm))
```

## Composite (A: JLM ┬Ě B: per-SNP MLM ┬Ě C: RTIGER-mosaic MLM)

Panel B (per-SNP genotype (bcftools HWE-posterior)) is ╬╗_GC-deflated ÔÇö the 70%-missing per-SNP panel at 0.4├Ś
is sparse ÔÇö but catches sharp SNP hits (`ZCN8`) the mosaic block-terraces away. Panel C (the
ancestry mosaic) is the usable, well-calibrated genotype. The BÔćöC contrast is the notebook's
per-SNP-scatter vs ancestry-block-smoothing, and the ZEAL demonstration that ancestry
inference is mandatory at low coverage.

```{r composite, fig.width=9, fig.height=13}
p_B <- plot_manhattan(gt,
  sprintf("DTA ÔÇö MLM (Taxon+K), per-SNP genotype (bcftools HWE-posterior) (╬╗_GC = %.2f; dotted = perm FWER 5%%)", lam(gt$P)),
  overlap_csv = overlap, lod = thr_gt
)
p_C <- plot_manhattan(mlm,
  sprintf("DTA ÔÇö MLM (Taxon+K), RTIGER ancestry mosaic (╬╗_GC = %.2f; dotted = perm FWER 5%%)", lam(mlm$P)),
  overlap_csv = overlap, lod = thr_mlm
)
composite <- cowplot::plot_grid(p_lolli, p_B, p_C, ncol = 1, labels = c("A", "B", "C"), label_size = 20, align = "v", axis = "lr")
ggsave(here("results/sim/zeal/dta_composite_snp50k.png"), composite, width = 9, height = 13, dpi = 200, bg = "white")
composite
```

## Multi-caller DTA scan (control ┬Ě nNIL ┬Ě RTIGER ┬Ě LB-Impute ┬Ě binhmm)

The ZEAL analog of the TeoNAM DTA five-caller composite: the same DTA MLM (Taxon+K) scan on
each ancestry caller's recovered genotypes, at the real ~0.4├Ś depth (not a coverage sweep).
The scientific content here is **caller reliability at low coverage**: the ancestry mosaic
must be **block/bin-aggregated** ÔÇö the two aggregated callers (RTIGER per-donor segments;
binhmm 1 Mb bins) recover the correct teosinte fraction (~0.10, matching the established
BzeaSeq), while per-SNP calling (control) and the fresh per-SNP HMM recalls (nNIL, LB-Impute)
**under-call** and give ╬╗-deflated scans.

```{r caller-table}
callers <- c(control = "hwe_post_gt", nNIL = "nnil_mosaic", RTIGER = "rtiger_mosaic", `LB-Impute` = "lbimpute_mosaic", binhmm = "binhmm_mosaic")
tb1 <- fread(here("data/teonam/dta_candidate_genes.tsv"))[symbol == "zmcct10"]
ctab2 <- rbindlist(lapply(names(callers), function(nm) {
  g <- callers[[nm]]
  sc <- fread(here(sprintf("data/zeal/dta_gwas_mlm_%s_donor_accession_snp50k.csv", g)))
  mf <- here(sprintf("data/zeal/zeal_%s.rds", g))
  pres <- if (file.exists(mf)) {
    st <- readRDS(mf)$state
    round(mean(st %in% c(1L, 2L), na.rm = TRUE), 3)
  } else {
    NA_real_
  }
  sc <- sc[is.finite(P) & P > 0]
  z <- sc[CHR == tb1$chr & abs(BP - tb1$start) <= 5e5]
  data.table(
    caller = nm, `teosinte presence` = pres, lambda_GC = lam(sc$P),
    `n > perm 5%` = nsig(sc$P, gwas_thr(TR, "mlm", g)),
    `zmcct10 -log10P` = if (nrow(z)) round(max(z[is.finite(P) & P > 0, -log10(P)]), 2) else NA_real_
  )
}))
knitr::kable(ctab2, caption = "DTA MLM (Taxon+K) by ancestry caller at 0.4├Ś, with n markers above the caller's own 1000-permutation 5% FWER line and the zmcct10 window-max -log10P. Aggregated callers (RTIGER, binhmm) recover teosinte presence ~0.10; per-SNP (control) and per-SNP recalls (nNIL, LB-Impute) under-call and deflate ╬╗_GC.")
```

```{r caller-composite, fig.width=9, fig.height=20}
panels <- lapply(names(callers), function(nm) {
  sc <- fread(here(sprintf("data/zeal/dta_gwas_mlm_%s_donor_accession_snp50k.csv", callers[[nm]])))
  plot_manhattan(sc, sprintf("DTA MLM (Taxon+K) ÔÇö %s (╬╗_GC = %.2f; perm FWER 5%%)", nm, lam(sc$P)), overlap_csv = overlap, lod = gwas_thr(TR, "mlm", callers[[nm]]))
})
cc <- cowplot::plot_grid(plotlist = panels, ncol = 1, labels = LETTERS[1:5], label_size = 18, align = "v", axis = "lr")
ggsave(here("results/sim/zeal/dta_caller_composite_fdr_snp50k.png"), cc, width = 9, height = 20, dpi = 200, bg = "white")
cc
```
