---
title: "ZEAL flowering — photoperiod QTL (marker × macroenvironment) R/qtl scan"
subtitle: "The day-length contrast: the same BC2S3 NILs phenotyped in a long-day temperate macroenvironment (Clayton NC, ~35.7°N) and a short-day subtropical one (CUCBA, Jalisco, ~20.7°N). QTL × macroenvironment (G×ME) interaction by R/qtl Approach B: a doubled cross with an environment covariate, interaction LOD = scanone(full: taxon+env, marker×env) − scanone(additive: taxon+env). Genome-wide 1000-permutation, line-level interaction threshold. Loci with a significant G×ME are photoperiod-sensitivity candidates."
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

**What this is.** The photoperiod scan for ZEAL flowering time. Macroenvironment (**ME**) = day-length
class: **long-day temperate** (Clayton CLY23 + CLY25 BLUEs) vs **short-day subtropical** (CUCBA,
Jalisco BLUE). A locus whose teosinte-allele effect on flowering **changes between day lengths**
shows up as a **marker × ME interaction** — that is photoperiod sensitivity. Same `bcsft` cross,
TeoNAM JLM map, and Haley–Knott machinery as the additive DTA/DTS notebooks. Inputs produced by
`scripts/zeal_mexg_scan.R` (`TRAIT=DTA` / `TRAIT=DTS`; writes `mexg_<trait>_scan.csv` /
`mexg_<trait>_result.rds`), whose CUCBA BLUEs come from `scripts/zeal_cucba_pheno.R`.

::: {.callout-important}
## Method (R/qtl Approach B) and what is / isn't solid
- **Doubled cross:** each NIL contributes two records — env 0 = Clayton (long day), env 1 = CUCBA
  (short day) — with identical genotype. `addcovar = [taxon, env]`; the **interaction LOD** is
  `scanone(intcovar = env)` − `scanone(additive)`, i.e. the marker × ME term.
- **Threshold** = genome-wide **1000-permutation**, permuted at the **line level** (breaking
  genotype↔phenotype while keeping the within-line env pairing), so it accounts for the two
  records-per-line non-independence that a naïve permutation would ignore.
- **Provisional:** the short-day side is **CUCBA only** (La Soledad flowering equals CUCBA minus a
  constant 3 days — no independent signal — so it was excluded; the cause is not yet established,
  most likely a planting-date offset or a recording artifact, to be confirmed against the raw dates
  with the Jalisco team), one **unreplicated** environment; and the CUCBA↔pedigree join is the
  Genealogía crosswalk (the mapped subset), pending the definitive three-way sample sheet. Treat as a
  strong first pass, not a final result.
- These interaction loci are **distinct from the additive short-day QTL** (see the CUCBA notebooks)
  and from the main-effect Clayton QTL — read them as their own signal.
:::

```{r setup}
library(here)
library(data.table)
library(qtl)
library(ggplot2)
library(ggrepel)
RQ <- here("results/sim/zeal/rqtl")
bp_of <- function(m) as.integer(sub("^S[0-9]+_", "", m))

# flowering candidates (v5) on the TeoNAM cM axis, per trait
.nat <- fread(here("data/teonam/teonam_v5_native.tsv"))
load_cand <- function(trait_lc) {
  cand <- fread(here(sprintf("data/teonam/%s_candidate_genes.tsv", trait_lc)))
  cand[, mid := (start + end) / 2]
  cand[, cm := mapply(function(c, b) {
    n <- .nat[chr_v5 == c & is.finite(cm)][order(pos_v5)]
    if (nrow(n) < 2) NA_real_ else approx(n$pos_v5, n$cm, xout = b, rule = 2)$y
  }, chr, mid)]
  cand[]
}

# per-trait GxE peaks: per-chr max real marker, 1.5-LOD CI, candidate-in-CI, sig flag
gxe_peaks <- function(res, cand) {
  g <- as.data.table(res$gxe, keep.rownames = "marker")
  g <- g[grepl("^S[0-9]+_", marker)][is.finite(lod)]
  thr5 <- as.numeric(res$thr[2])
  g[,
    {
      i <- which.max(lod)
      L <- lod[i]
      ins <- .SD[lod >= L - 1.5]
      lbp <- min(bp_of(ins$marker))
      hbp <- max(bp_of(ins$marker))
      cin <- cand[chr == .BY$chr & mid >= lbp & mid <= hbp, symbol]
      .(
        `pos (cM)` = round(pos[i], 1), `pos (Mb)` = round(bp_of(marker[i]) / 1e6, 1),
        marker = marker[i], `interaction LOD` = round(L, 2), `perm 5% thr` = round(thr5, 2),
        sig = fifelse(L >= thr5, "yes", "no"),
        `1.5-LOD CI (Mb)` = sprintf("%.1f–%.1f", lbp / 1e6, hbp / 1e6),
        `candidate in CI` = if (length(cin)) paste(cin, collapse = ", ") else "–"
      )
    },
    by = chr
  ][order(-`interaction LOD`)]
}

# per-trait GxE genome-wide profile (interaction LOD, cM axis, candidates + threshold)
gxe_profile <- function(res, cand, trait) {
  g <- as.data.table(res$gxe, keep.rownames = "marker")[is.finite(lod)]
  thr5 <- as.numeric(res$thr[2])
  nperm <- length(res$permmax)
  sig <- g[grepl("^S[0-9]+_", marker) & lod >= thr5][, .SD[which.max(lod)], by = chr]
  cl <- cand[chr %in% unique(g$chr) & is.finite(cm)]
  ytop <- max(g$lod, na.rm = TRUE)
  ggplot(g, aes(pos, lod)) +
    geom_vline(data = cl, aes(xintercept = cm), inherit.aes = FALSE, linetype = "dotted", colour = "grey55", linewidth = 0.3) +
    geom_line(colour = "#2c6fbb", linewidth = 0.4) +
    geom_hline(yintercept = thr5, linetype = "dashed", colour = "red", linewidth = 0.35) +
    geom_point(data = sig, colour = "red", size = 1.6) +
    ggrepel::geom_text_repel(
      data = cl, aes(x = cm, y = ytop * 1.18, label = symbol), inherit.aes = FALSE,
      size = 2.2, colour = "grey25", angle = 90, hjust = 0, direction = "x", segment.size = 0.2, min.segment.length = 0
    ) +
    expand_limits(y = ytop * 1.45) +
    facet_grid(~chr, scales = "free_x", space = "free_x", switch = "x") +
    labs(
      x = "Chromosome (cM, TeoNAM JLM map)", y = "marker × ME interaction LOD",
      title = sprintf("ZEAL %s GxE — photoperiod (marker × macroenvironment) interaction", trait),
      subtitle = sprintf("short-day CUCBA vs long-day Clayton · 1000-perm line-level 5%% threshold = %.2f · dotted = flowering candidates", thr5)
    ) +
    theme_classic(base_size = 11) +
    theme(
      axis.text.x = element_blank(), axis.ticks.x = element_blank(),
      panel.spacing = unit(0.05, "lines"), strip.placement = "outside",
      strip.background = element_blank()
    ) +
    scale_x_continuous(expand = c(0.01, 0))
}

has_trait <- function(trait_lc) file.exists(file.path(RQ, sprintf("mexg_%s_result.rds", trait_lc)))
```

## DTA GxE

```{r dta-gxe, fig.width=11, fig.height=4.4}
if (!has_trait("dta")) {
  cat("mexg_dta_result.rds not found — run: TRAIT=DTA Rscript scripts/zeal_mexg_scan.R")
} else {
  res <- readRDS(file.path(RQ, "mexg_dta_result.rds"))
  cand <- load_cand("dta")
  print(gxe_profile(res, cand, "DTA"))
}
```

```{r dta-gxe-peaks}
if (has_trait("dta")) {
  res <- readRDS(file.path(RQ, "mexg_dta_result.rds"))
  pk <- gxe_peaks(res, load_cand("dta"))
  knitr::kable(pk, caption = sprintf(
    "DTA photoperiod (marker × ME) interaction: per-chromosome peak, %d lines, 1000-perm line-level 5%% threshold = %.2f. 'sig' = clears the interaction threshold. 1.5-LOD interval and any flowering candidate inside.",
    res$n, as.numeric(res$thr[2])
  ))
}
```

## DTS GxE

```{r dts-gxe, fig.width=11, fig.height=4.4}
if (!has_trait("dts")) {
  cat("mexg_dts_result.rds not found yet — run: TRAIT=DTS Rscript scripts/zeal_mexg_scan.R (this notebook picks it up on re-render).")
} else {
  res <- readRDS(file.path(RQ, "mexg_dts_result.rds"))
  print(gxe_profile(res, load_cand("dts"), "DTS"))
}
```

```{r dts-gxe-peaks}
if (has_trait("dts")) {
  res <- readRDS(file.path(RQ, "mexg_dts_result.rds"))
  pk <- gxe_peaks(res, load_cand("dts"))
  knitr::kable(pk, caption = sprintf(
    "DTS photoperiod (marker × ME) interaction: per-chromosome peak, %d lines, 1000-perm line-level 5%% threshold = %.2f.",
    res$n, as.numeric(res$thr[2])
  ))
}
```

## Reading the scan

A significant **marker × ME** peak means the teosinte allele's effect on flowering **differs between
short-day (CUCBA) and long-day (Clayton)** — i.e. the locus is **photoperiod-sensitive**. These are
not the same as the additive flowering QTL: a locus can have a large, consistent effect in both day
lengths (strong additive, no interaction) or a modest average effect that flips sign with day length
(weak additive, strong interaction). Candidate genes are marked for orientation only — a peak's
identity needs a coordinate check against the v5 annotation, not assignment by chromosome. All results
are **provisional** pending real La Soledad flowering (or additional replication) and the definitive
three-way sample sheet; the short-day side is currently one unreplicated environment (CUCBA).
